### 2016-06-06 paper/paper.tex
mostly finished an outline of the paper

%-----------------------------------------------------------------------------
%
%               Template for sigplanconf LaTeX Class
%
% Name:         sigplanconf-template.tex
%
% Purpose:      A template for sigplanconf.cls, which is a LaTeX 2e class
%               file for SIGPLAN conference proceedings.
%
% Guide:        Refer to "Author's Guide to the ACM SIGPLAN Class,"
%               sigplanconf-guide.pdf
%
% Author:       Paul C. Anagnostopoulos
%               Windfall Software
%               978 371-2316
%               paul@windfall.com
%
% Created:      15 February 2005
%
%-----------------------------------------------------------------------------

\documentclass[preprint]{sigplanconf}

% The following \documentclass options may be useful:

% preprint      Remove this option only once the paper is in final form.
% 10pt          To set in 10-point type instead of 9-point.
% 11pt          To set in 11-point type instead of 9-point.
% numbers       To obtain numeric citation style instead of author/year.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amsmath}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{amsthm}
\makeatletter
\def\th@definition{%
  \thm@notefont{}% same as heading font
  \normalfont % body font
}
\makeatother
\theoremstyle{definition}
\newtheorem{problem}{Problem}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\usepackage{courier}
\usepackage{listings}
\lstset{
    language=Haskell,
    basicstyle=\ttfamily,
    keywordstyle=\ttfamily,
    literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
             {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
    %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
    %commentstyle=\color{gray}
}
\lstMakeShortInline|

\newcommand{\cL}{{\cal L}}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{filecontents}{paper.bib}
@article{goldberg1991,
  title={What every computer scientist should know about floating-point arithmetic},
  author={Goldberg, David},
  journal={ACM Computing Surveys (CSUR)},
  volume={23},
  number={1},
  pages={5--48},
  year={1991},
  publisher={ACM}
}

@book{higham2002,
  title={Accuracy and stability of numerical algorithms},
  author={Higham, Nicholas J},
  year={2002},
  publisher={Siam}
}

@article{swierstra2008,
  title={Data types {\`a} la carte},
  author={Swierstra, Wouter},
  journal={Journal of functional programming},
  volume={18},
  number={04},
  pages={423--436},
  year={2008},
  publisher={Cambridge Univ Press}
}

@book{hamming2012,
  title={Numerical methods for scientists and engineers},
  author={Hamming, Richard},
  year={2012},
  publisher={Courier Corporation}
}

@inproceedings{gupta2015,
  title={Deep Learning with Limited Numerical Precision},
  author={Gupta, Suyog and Agrawal, Ankur and Gopalakrishnan, Kailash and Narayanan, Pritish},
  booktitle={Proceedings of the 32nd International Conference on Machine Learning (ICML-15)},
  pages={1737--1746},
  year={2015}
}
\end{filecontents}
\immediate\write18{bibtex paper}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\begin{document}

\special{papersize=8.5in,11in}
\setlength{\pdfpageheight}{\paperheight}
\setlength{\pdfpagewidth}{\paperwidth}

\conferenceinfo{CONF 'yy}{Month d--d, 20yy, City, ST, Country}
\copyrightyear{20yy}
\copyrightdata{978-1-nnnn-nnnn-n/yy/mm}
\copyrightdoi{nnnnnnn.nnnnnnn}

% Uncomment the publication rights you want to use.
%\publicationrights{transferred}
%\publicationrights{licensed}     % this is the default
%\publicationrights{author-pays}

\titlebanner{Preprint}        % These are ignored unless
\preprintfooter{Preprint}   % 'preprint' option specified.

%\title{What's the point of homoiconicity?}
\title{The (floating) point of homoiconicity}
\subtitle{Functional Pearl}

\authorinfo{Mike Izbicki}
           {UC Riverside}
           {mike@izbicki.me}
%\authorinfo{Name2\and Name3}
           %{Affiliation2/3}
           %{Email2/3}

\maketitle

\begin{abstract}
%This pearl uses Haskell's type classes to enable a limited form of \emph{homoiconicity}.
%%Specifically, we show how to convert any Haskell98 function that is parametric in its return type into an isomorphic abstract syntax tree.
%Specifically, we show how to convert a certain set of polymorphic functions into isomorphic abstract syntax trees.
%But this is really a pearl about the story of numerical programming in Haskell.

%Numeric programming in Haskell sucks.
%Actually, it sucks in every language because it requires floating point numbers,
%and floating point arithmetic is so difficult to work with.
%The goal of this pearl is to make numeric programming suck just a little bit less.
%This statement is tautologically true---numeric programming requires floating point arithmetic, and working with floats sucks in every language.
%I think this makes numeric programming in Haskell suck just a little bit less.

Floating point numbers are dangerous.
Due to their limited precision,
seemingly correct mathematical formulae can give wildly inaccurate results.
This pearl uses Haskell's type system to mitigate this danger.

The main observation is that many numerical Haskell functions are \emph{homoiconic}.
That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
The construction does not use template Haskell, and so works at both compile and run time.
We use these ASTs to solve two problems with floating point numbers:
\begin{enumerate}
\item
We show how to write functions that automatically stabilize (and optimize!) mathematical formulae.
This lets the programmer write naive floating point code;
the compiler will worry about improving the numeric stability and speed later.
\item
We show how to specify type class laws for floating point numbers that can be easily and automatically tested.
This is a difficult unsolved problem because floating point numbers do not obey many of the traditional algebraic laws like associativity.
\end{enumerate}
We demonstrate these techniques using GHC's builtin numeric hierarchy,
but these techniques generalize to user defined numeric hierarchies as well.

%First, we show how to convert a certain type of commonly used mathematical function into isomorphic abstract syntax trees.
%Specifically, we show that a subset of Haskell functions (which we call \emph{homogeneous}) can easily be converted into an isomorphic abstract syntax tree.

%I hate numerical programming in Haskell,
%but I hate numerical programming in every language.
%It requires floating point numbers,
%which
%
%I hate floating point numbers.

%This pearl provides a technique that makes working with approximate algebraic structures more convenient.
%The prototypical example of an approximate structure is floating point arithmetic, which is famously not a semigroup because it is not associative.
%The technique has two parts.
%First, it lets us construct an FAlgebra (and hence a syntax tree) for a numerical class hierarchy.
%Second, we show how to treat topology as an algebraic structure, and that floating point types have a useful topology.
%This topology can be used to automatically construct properties for approximate algebraic structures that can be easily checked using quickcheck or smallcheck.
\end{abstract}

\category{CR-number}{subcategory}{third-level}

% general terms are not compulsory anymore,
% you may leave them out
%\terms
%term1, term2

\keywords
floating point, homoiconic, type classes

\section{Introduction}

Imagine you're writing a machine learning library in Haskell.
At some point, you'll need to implement the log-logistic function
\begin{equation}
f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
\label{eq:log-logistic}
\end{equation}
This function is used in logistic regression and many families of neural networks.
As a first attempt, you might write:
\begin{lstlisting}
logLogistic1 :: Floating x => x -> x
logLogistic1 x = log(1/(1+exp(-x)))
\end{lstlisting}
Unfortunately, this straightforward implementation is numerically unstable.
When specialized to the |Double| type,
the |exp| function overflows to infinity for inputs greater than (approximately) $709$.
%This overflow causes |logLogistic| to take the |log| of 0, which is undefined.
Therefore, on an input of $-710$,
|logLogisitic1| will return negative infinity when the true answer is actually very close to $-710$.
Such an inaccurate answer is unacceptable,
so our function must be rewritten.

Writing numerically stable functions is a notoriously difficult black art.
(See \cite{goldberg1991,higham2002,hamming2012} for some classic introductions.)
Fortunately, a simple technique can be used to improve the stability of the log-logistic function.
First, check our input to see if it would overflow the |exp| function.
If no overflow would occur, then calculate our function as normal;
but if overflow would occur, then use an approximation that does not rely on a call to |exp|.
Performing this transformation gives us:
\begin{lstlisting}
logLogistic2 :: (Floating x, Ord x) => x->x
logLogistic2 x = if x < (-709)
    then x
    else log(1/1+exp(-x))
\end{lstlisting}
%This new function is numerically stable, but we've introduced two new problems:
We've made our function numerically stable, but we've introduced two new problems:

\begin{problem}[obfuscation]
The purpose of the if statement in |logLogistic2| is non-obvious.
To understand its purpose,
the reader must carefully consider the limits of floating point arithmetic.
Anyone who doesn't regularly work with floating point numbers is unlikely to discover this purpose on their own.
By transforming our function to have the correct floating point semantics,
we've obfuscated the function's purpose and made it harder for others to understand and contribute to our library.
\end{problem}

\begin{problem}[specialization]
The \lstinline{exp} function when specialized to \lstinline{Float}s overflows on values greater than (approximately) 88.
This means that |logLogistic2| will give incorrect answers for inputs between 88 and 709.
To properly handle both |Float|s and |Double|s,
we must write two separate functions.
In practice, we may want to support many other numeric types as well.
Half precision (16 bit) floats and fixed point arithmetic are becoming increasingly popular in machine learning applications due to their increased speed, smaller memory footprint, and lower energy consumption \cite{gupta2015}.
We need to manually reimplement the log-logistic function specialized to every numeric type we want our library to support.
\end{problem}

These problems are not unique to the log-logistic function.
All functions over the real or complex numbers need to be transformed for floating point stability,
and all such manual transformations suffer from these problems.
%(In fact, the log-logistic transformation described above might be considered an exceptionally simple case.)
%There is a large body of literature on the theory of automating these transformations,
%but relatively little on improving the programmer's experience of dealing with floating point numbers.
%This pearl focuses on the latter.
This pearl presents a method for alleviating these problems.
Specifically, we will see how to write a function
\begin{lstlisting}
stabilize :: (Floating x, Ord x)
    => (x -> x) -> (x -> x)
\end{lstlisting}
that automatically takes an input function like |logLogistic1| and outputs a numerically stable version of that function.
Unfortunately, this introduces a third problem:

\begin{problem}[unlawfulness]
%How do we know what transformations are valid?
The standard way to encode valid program transformations in Haskell is through type class laws.
A classic example is the |Semigroup| type class
\begin{lstlisting}
class Semigroup a where
    (+) :: a -> a -> a
\end{lstlisting}
which has the law that |(+)| should be associative.
%That is, for every type |a| that is a valid instance of |Semigroup|,
%|(a1+a2)+a3::a| should be equal to |a1+(a2+a3)::a|.
That is, if the type |a| is an instance of |Semigroup|,
and the variables |a1|, |a2|, and |a3| all have type |a|,
then |(a1+a2)+a3| should be equal to |a1+(a2+a3)|.
This is not true of floating point numbers, however.
The expression |(0.1+0.2)+0.3| is famously not equal to |0.1+(0.2+0.3)| for |Double|s.
But there is clearly a sense in which |Double|s are approximately associative,
and performing an associativity transformation cannot change our answer ``too much.''
%How can we formalize this notion?
In order for our |stabilize| method to be generic and useful,
we need a way to formalize this notion.
\end{problem}

This pearl attempts to solve all three problems.
First, we provide a mechanical procedure for constructing FAlgebras from what we call \emph{homogeneous} type classes.
(Both terms will be defined formally in the next section.)
These FAlgebras make our numeric functions \emph{homoiconic};
that is, we can easily get a syntax tree representing the function.
Second, we show how these syntax trees let us easily manipulate our numeric functions to implement the |stabilize| function described above.
Third, we introduce a notion of \emph{topological class laws}.
These class laws let us formalize the intuition that floating point numbers approximately obey class laws.
Finally, we extend our construction of FAlgebras to the \emph{heterogeneous} case.
Heterogeneous type classes let us represent the complicated numeric type class hierarchies that arise in practice when working with numeric types.

\section{Homogeneous Functions are Homoiconic}
Consider the simple numeric class hierarchy shown in Figure \ref{code:group}.
With this class hierarchy, we can write a function that measures the degree to which the |(+)| operation obeys the associative rule.
\begin{lstlisting}
associator :: Group a => a -> a -> a -> a
associator a1 a2 a3
    = ((a1+a2)+a3) - (a1+(a2+a3))
\end{lstlisting}
We call the |associator| function \emph{homogeneous} because:
\begin{enumerate}
\item
The function's return type and all of its parameters are the same type variable.
\item
The constraints contain exactly one type class, and
that type class has exactly one parameter.
\end{enumerate}
Notice that each of the class methods for |Semigroup|, |Monoid|, and |Group| are also homogeneous.
Therefore, we will also say these classes are homogeneous.

In this section, we will see how to convert any homogeneous function into an isomorphic abstract syntax tree (AST).
This ability is called \emph{homoiconicity},
and is usually associated with Lisps.
We will then see several methods for evaluating these ASTs.
In subsequent sections, we will manipulate these ASTs to make floating point operations more convenient.

\begin{figure}
\begin{lstlisting}
class Semigroup a where
    (+) :: a -> a -> a

class Semigroup a => Monoid a where
    zero :: a

class Monoid a => Group a where
    (-) :: a -> a -> a
    negate :: a -> a
\end{lstlisting}
\caption{A simple numeric hierarchy that obeys the restrictions for homogeneous FAlgebras.}
\label{code:group}
\end{figure}

%\begin{figure}
%\begin{lstlisting}
%class Num a where
    %(+), (-), (*)       :: a -> a -> a
    %negate              :: a -> a
    %abs, signum         :: a -> a
    %fromInteger         :: Integer -> a
%
%class Num a => Fractional a where
    %(/)                 :: a -> a -> a
    %recip               :: a -> a
    %fromRational        :: Rational -> a
%
%class Fractional a => Floating a where
    %pi                  :: a
    %exp, log, sqrt      :: a -> a
    %(**), logBase       :: a -> a -> a
    %sin, cos, tan       :: a -> a
    %asin, acos, atan    :: a -> a
    %sinh, cosh, tanh    :: a -> a
    %asinh, acosh, atanh :: a -> a
%
%class Eq a where
    %(==), (/=)          :: a -> a -> Bool
%
%class Eq a => Ord a where
    %compare         :: a -> a -> Ordering
    %(<),(<=),(>),(>=)   :: a -> a -> Bool
    %max, min            :: a -> a -> a
%\end{lstlisting}
%\caption{
    %A portion of the numeric and comparison hierarchies defined in GHC's Prelude.
    %Many class methods satisfy the requirements for homogeneous FAlgebras, except:
    %%|fromInteger| and |fromRational| have non-variable parameters;
    %%and |(==)|, |(/=)|, |(<)|, |(<=)|, |(>)|, and |(>=)| have non-variable return types.
%}
%\label{code:ghc}
%\end{figure}

\subsection{A type class for FAlgebras}
Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
See for example \cite{swierstra2008}.
The heart of our construction is the |HomFAlgebra| type class, defined as
\begin{lstlisting}
class
    ( Functor (HomSig alg)
    , Foldable (HomSig alg)
    ) => HomFAlgebra (alg::Type->Constraint)
        where
    data HomSig alg a
    runHomSig :: alg a => HomSig alg a -> a
\end{lstlisting}
The |HomFAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
This means that instances of |HomFAlgebra| will be other type classes (instead of types or type constructors).
The |Hom| in |HomFAlgebra| stands for homogeneous.
Most of the names used in this section will be prefixed with |Hom| to differentiate them from related names used in the later heterogeneous constructions.

The easiest way to explain the |HomFAlgebra| class is by example.
We'll start with the |Semigroup| type class defined above.
\begin{lstlisting}
instance HomFAlgebra Semigroup where
    data HomSig Semigroup a = HomSig_plus a a
    runHomSig (HomSig_plus a1 a2) = a1+a2
\end{lstlisting}
The |HomSig| associated data family is meant to encode what mathematicians call the \emph{signature} of the FAlgebra.
That is, for every class method, |HomSig| should have a constructor with the same number of parameters.
Since |Semigroup| has a single function that takes two parameters,
the |HomSig Semigroup| data instance has a single constructor that takes two parameters.
The |runHomSig| function evaluates the signature.
For each constructor in |HomSig|, |runHomSig| should apply the corresponding function to the constructor's parameters.

To complete the |HomFAlgebra Semigroup| instance, we must also define the following superclass instances.
\begin{lstlisting}
instance Functor (HomSig Semigroup) where
    fmap f (HomSig_plus a1 a2)
        = HomSig_plus (f a1) (f a2)

instance Foldable (HomSig Semigroup) where
    foldr f b (HomSig_plus a1 a2)
        = f a2 (f a1 b)
\end{lstlisting}
As usual, there is only a single valid |Functor| instance.
There are multiple valid |Foldable| instances depending on the order in which |foldr| considers the constructors' parameters,
but for our purposes the order is not important.

Instances get slightly more complicated when the parameter class has superclasses.
In this case, the |HomSig| data instance must also have a constructor for each superclass that stores the superclass's signature.
This can be seen in the |Monoid| and |Group| instances below.
%The |HomFAlgebra| instances of |Monoid| and |Group| demonstrate this.
\begin{lstlisting}
instance HomFAlgebra Monoid where
    data HomSig Monoid a
        = HomSig_zero
        | HomSig_Semigroup_Monoid
            (HomSig Semigroup a)
    runSig HomSig_zero = zero
    runSig (HomSig_Semigroup_Monoid s)
        = runHomSig s

instance HomFAlgebra Group where
    data HomSig Group a
        = HomSig_negate a
        | HomSig_minus a a
        | HomSig_Monoid
            (HomSig_Monoid_Group a)
    runSig (HomSig_negate a) = a
    runSig (HomSig_minus a1 a2) = a1-a2
    runSig (HomSig_Monoid_Group s)
        = runHomSig s
\end{lstlisting}
Notice that the data instance |HomSig Group| does not need a constructor for the |Semigroup| signature because this is included in the |Monoid| signature.
The corresponding |Functor| and |Foldable| instances have been omitted for brevity,
but can be found in the paper's git repo.

%The procedure for generating |HomFAlgebra| instances is purely mechanical.
%The ?? library automates the process with the |mkHomFAlgebra| template Haskell function.

\subsection{Constructing the AST}

We can convert the signature of an FAlgebra into an AST as follows:
\begin{lstlisting}
type HomAST alg a = HomFree (HomSig alg) a

data HomFree f a
    = HomPure a
    | HomFree (f (HomFree f a))
\end{lstlisting}
Intuitively,
the |HomPure| constructor represents a leaf in the AST,
and the |HomFree| constructor represents a branch.
The parameter to |HomFree| is filled by the appropriate |HomSig| data instance.
The constructor used for the |HomSig| instance corresponds to an operation,
and the parameters to the constructor will contain ASTs nested recursively.

For example, consider the expression |(1+2)+3|.
We can create a corresponding AST using |Semigroup|'s |HomFAlgebra| instance.
\begin{lstlisting}
HomFree
  (HomSig_plus
    (HomFree
      (HomSig_plus
        (HomPure 1)
        (HomPure 2)
      )
    )
    (HomPure 3)
  )
\end{lstlisting}
We can also create a corresponding AST for any class that has |Semigroup| as a superclass.
Below is the |Group| AST.
\begin{lstlisting}
HomFree
  (HomSig_Monoid_Group
    (HomSig_Semigroup_Monoid
      (HomSig_plus
        (HomFree
          (HomSig_Monoid_Group
            (HomSig_Semigroup_Monoid
              (HomSig_plus
                (HomPure 1)
                (HomPure 2)
              )
            )
          )
        )
        (HomPure 3)
      )
    )
  )
\end{lstlisting}
Notice that the |HomSig_plus| constructor must be embedded into the |Group|'s |HomSig| with calls to |HomSig_Monoid_Group| and |HomSig_Semigroup_Monoid|.

Constructing these ASTs by hand is a tedious process.
There is a considerable amount of boilerplate in the |Group| AST above as a result of the need to embed the |HomSig_plus| constructor.
We can avoid this boilerplate with the following type class.
\begin{lstlisting}
class
    ( HomFAlgebra alg1
    , HomFAlgebra alg2
    ) => HomView alg1 alg2
        where
    embedHomSig
        :: HomSig alg1 a -> HomSig alg2 a
    unsafeExtractHomSig
        :: HomSig alg2 a -> HomSig alg1 a
\end{lstlisting}
Instances of the |HomView| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
all such relationships should have a corresponding instance.
For example, |Semigroup| is a superclass of |Group|,
so we should have a |HomView Semigroup Group| instance.
\begin{lstlisting}
instance HomView Semigroup Group
    embedHomSig s
        = HomSig_Monoid_Group (embedHomSig s)
    unsafeExtractHomSig (HomSig_Monoid_Group s)
        = unsafeExtractHomSig s
\end{lstlisting}
Because of this property, we know that the |HomSig alg1 a| data instance can always be embedded into a |HomSig alg2 a| data instance.
But it is only sometimes true that we can extract a |HomSig alg1 a|,
which is why the corresponding function is labeled |unsafe|.

Armed with the |HomView| class, we can create a uniform representation for all class's ASTs.
The expression |(1+2)+3| can now be represented as:
\begin{lstlisting}
HomFree
  (embedHomSig
    (HomSig_plus
      (HomFree
        (embedHomSig
          (HomSig_plus
            (HomPure 1)
            (HomPure 2)
          )
        )
      )
      (HomPure 3)
    )
  )
\end{lstlisting}

%Every FAlgebra can be embedded into itself.
%\begin{lstlisting}
%instance HomView alg alg where
    %embedSig = id
    %unsafeExtractSig = id
%\end{lstlisting}

%\begin{lstlisting}
%instance View Semigroup Monoid where
    %embedSig = Sig_Monoid_Semigroup
    %unsafeExtractSig (Sig_Monoid_Semigroup s) = s

%instance View Monoid Group where
    %embedSig s = Sig_Group_Monoid (embedSig s)
    %unsafeExtractSig (Sig_Group_Monoid s) = unsafeExtractSig s
%\end{lstlisting}

%\begin{lstlisting}
%instance View Semigroup Group where
    %embedSig s = Sig_Group_Monoid (embedSig s)
    %unsafeExtractSig (Sig_Group_Monoid s) = unsafeExtractSig s
%\end{lstlisting}

Using our uniform representation of an AST,
we can make the AST itself an instance of any homogeneous type class.
Again, we'll approach the procedure by example.
We can write a |Semigroup| instance for |HomAST Semigroup a| as:
\begin{lstlisting}
instance
    ( HomView Semigroup alg
    ) => Semigroup (HomFree (HomSig alg) a)
        where
    (+) e1 e2 = HomFree $ embedHomSig $
            HomSig_plus e1 e2
\end{lstlisting}
For each function in the type class,
we simply embed the function's constructor into the AST by prepending the expression |HomFree $ embedHomSig $|.
The |Monoid| and |Group| instances follow a similar pattern.
\begin{lstlisting}
instance
    ( HomView Semigroup alg
    , HomView Monoid alg
    ) => Monoid (HomFree (HomSig alg) a)
        where
    zero = HomFree $ embedHomSig $
        HomSig_zero

instance
    ( HomView Semigroup alg
    , HomView Monoid alg
    , HomView Group alg
    ) => Group (HomFree (HomSig alg) a)
        where
    negate e  = HomFree $ embedHomSig $
        HomSig_negate e
    (-) e1 e2 = HomFree $ embedHomSig $
        HomSig_minus e1 e2
\end{lstlisting}
We can easily create ASTs using these instances.
The |Semigroup| AST can be created with the expression
\begin{lstlisting}
(HomPure 1+HomPure 2)+HomPure 3
    :: HomAST Semigroup a
\end{lstlisting}
And the |Group| AST can be created by just changing the type signature.
\begin{lstlisting}
(HomPure 1+HomPure 2)+HomPure 3
    :: HomAST Group a
\end{lstlisting}

\subsection{Showing the AST}
Now that we can create our ASTs, we want to actually do something with them!
The simplest task is to convert them into a |String| for display.
This is an easy procedure of writing |Show| instances for the |HomSig| and |HomFree| data types.

The are no tricks in the |Show| instance for |Semigroup|'s |HomSig|.
\begin{lstlisting}
instance Show a =>
    Show (HomSig Semigroup a)
        where
    show (HomSig_plus a1 a2)
        = show a1++"+"++show a2
\end{lstlisting}
For each operation constructor, we simply show the parameters and the operation connecting them.
For type classes with superclasses, we also need to recursively show the superclasses as can be seen in the |Monoid| and |Group| instances below.
\begin{lstlisting}
instance Show a =>
    Show (HomSig Monoid a)
        where
    show (HomSig_Semigroup_Monoid s)
        = show s
    show (HomSig_zero) = "zero"

instance Show a =>
    Show (HomSig Group a)
        where
    show (HomSig_Monoid_Group s)
        = show s
    show (HomSig_negate a)
        = "negate "++show a
    show (HomSig_minus a1 a2)
        = show a1++"-"++show a2
\end{lstlisting}
The |HomFree| type's |Show| instance is similarly straightforward..
\begin{lstlisting}
instance (Show a, Show (f (HomFree f a)))
    => Show (HomFree f a)
        where
    show (HomPure a) = show a
    show (HomFree f) = "("++show f++")"
\end{lstlisting}
If the syntax tree contains only a single leaf, we just show that leaf.
If the syntax tree contains a branch,
then put parenthesis around the branch so that we don't lose information about the order of operations, and show the functor in the middle.

\subsection{Evaluating the syntax tree}
Another useful task for ASTs is to evaluate them,
generating the value they represent.
This data type is called the free monad because as long as the parameter |f| is a |Functor|,
then |HomFree| has a valid |Monad| instance.
Since all |Monad|s are by necessity also |Functor|s,
|HomFree| has a valid |Functor| instance as well.
\begin{lstlisting}
instance Functor f => Functor (HomFree f) where
    fmap g (HomPure a) = HomPure (g a)
    fmap g (HomFree f)
        = HomFree (fmap (fmap g) f)
\end{lstlisting}
For our purposes, that's all the structure we'll need.

First, we construct a function that ``runs'' the syntax tree, converting it into a single value.
\begin{lstlisting}
runHomAST :: (HomFAlgebra alg, alg a)
    => HomAST alg a -> a
runHomAST (HomPure a) = a
runHomAST (HomFree f) = runHomSig (fmap evalHom f)
\end{lstlisting}
If our syntax tree already consists of just a single value (i.e. it is the |HomePure| constructor),
then we just return that value.
Otherwise, we use the |Functor| instance to recursively convert the |HomSig|s into types of |a|, then a call to |runHomSig| combines the results.

\subsection{Homoiconicity at last!}
We are finally ready to see that our construction makes homogeneous functions homoiconic.
If we open GHCi and enter the following expression
\begin{lstlisting}
ghci> let ast=associator (HomPure a1) (HomPure a2) (HomPure a3)
\end{lstlisting}
the display shows our AST
\begin{lstlisting}
\end{lstlisting}
We can also evaluate our AST and see that doing so gives the same results as running the original expression.
\begin{lstlisting}
ghci> runHomAST ast

ghci> associator a1 a2 a3

\end{lstlisting}
We are now ready to use our ASTs to make working with numeric functions easier.
%\subsection{The Prelude's Numeric Hierarchy}

\section{Implementing the stabilize function}

We've seen how to easily construct the ASTs,
now it's time to perform computations on them.
The standard way to deconstruct an AST is via pattern matching.
We'll use the |ViewPatterns| and |PatternSynonyms| language extensions to simplify this task.
Then we'll see three examples of improving floating point accuracy.

\subsection{Pattern matching the AST}

We'll begin with a very simple example.
Consider the following simple, but numerically unstable function.
\begin{lstlisting}
testFunc1 :: Floating a => a -> a
testFunc1 = log(exp a)
\end{lstlisting}
As in our |logLogistic1| function, |testFunc| is unstable when specialized to |Double|s
because input values greater than (approximately) 710 cause the |exp| function to overflow to infinity;
the log of infinity is still infinity;
so the returned answer is infinity.
But clearly this is not the correct answer.
Mathematically, the $\log$ is the inverse of $\exp$, so the final result should be 711.

First, we'll see how to properly manipulate the AST to fix this problem.
%We need a function of type |AST Floating a -> AST Floating a|.
We're trying to convert the expression |log (exp x)| into simply |x|.
To do this, we can just pattern match on the AST for the left hand side of the expression,
and replace it with the right hand side.
\begin{lstlisting}
logexpAST1 ::
    AST Floating a -> AST Floating a
logexpAST1
    (Free (Sig_log (Free (Sig_exp a)))) = a
logexpAST1 a = a
\end{lstlisting}
This function has two problems.
First, our pattern matching expression is large and obnoxious.
Second, it's not generic.

We can fix the second problem using the |ViewPatterns| language extension.
\begin{lstlisting}
logexpAST2 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST2
    (Free (unsafeExtractSig -> Sig_log
    (Free (unsafeExtractSig -> Sig_exp a))))
    = a
logexpAST2 a = a
\end{lstlisting}
But this made our pattern matches even uglier!
Fortunately, we can use the |PatternSynonyms| extension to greatly simplify these patterns.
\footnote{
    %These patterns can be made bidirectional using the |embedSig| function.
    For ease of presentation, only the unidirectional case is shown.
}
\begin{lstlisting}
pattern AST_log ::
    ( View Floating alg
    ) => AST alg a -> AST alg a
pattern AST_log e
    <- Free (unsafeExtractSig -> Sig_log e)

pattern AST_exp ::
    ( View Floating alg
    ) => AST alg a -> AST alg a
pattern AST_exp e
    <- Free (unsafeExtractSig -> Sig_exp e)
\end{lstlisting}
Armed with these pattern synonyms, we can rewrite our function in the much simpler form below.
\begin{lstlisting}
logexpAST3 :: View Floating alg
    => AST alg a -> AST alg a
logexpAST3 (AST_log (AST_exp a)) = a
logexpAST3 a = a
\end{lstlisting}
Now that we can manipulate the AST, let's apply it to our function.
\begin{lstlisting}
transform1 :: forall alg x.
    ( FAlgebra alg
    , alg x
    ) => (AST alg x -> AST alg x)
      -> (AST alg x -> AST alg x)
      -> (x -> x)
transform1 go f x = runHomAST $ go $ f (Pure x)
\end{lstlisting}

But is our function really correct?
\begin{lstlisting}
testFunc2 :: Floating a => a -> a
testFunc2 = 1+log(exp a)
\end{lstlisting}
What if we call

\begin{lstlisting}
logexpAST4 :: View Floating alg => AST alg a -> AST alg a
logexpAST4 (AST_log (AST_exp a)) = a
logexpAST4 (Free f) = Free (fmap logexpAST4 f)
logexpAST4 (Pure a) = Pure a
\end{lstlisting}

\begin{lstlisting}
testFunc :: Floating a => a -> a
testFunc a = log(exp a)

ghci> logexpAST4 (testFunc 2) :: AST Floating Double
((fromInteger 1)+(fromInteger 2))

testFunc2 :: Floating a => a -> a
testFunc2 a = 1+exp(exp(exp(log(log(log a)))))

ghci> fixAST stabilizeAST (testFunc2 var1 :: AST Floating Var)
((fromInteger 1)+var1)
\end{lstlisting}

\begin{lstlisting}
fixAST :: (Eq (Sig alg (Free (Sig alg) a)), Eq a) => (AST alg a -> AST alg a) -> AST alg a -> AST alg a
fixAST f ast = if ast==ast'
    then ast
    else fixAST f ast'
    where
        ast' = f ast
\end{lstlisting}

\subsection{More complex rules}

The rules in in the previous section are relatively simple.
They could just as easily have been written using GHC's rewrite rules syntax.
Rewrite rules are the primary mechanism for optimising source code in GHC.
One optimization that rewrite rules cannot support is constant folding,
and there is code in GHC specially to perform this constant folding.

\begin{lstlisting}
foldConstants :: View Num alg => AST alg a -> AST alg a
foldConstants (AST_plus   (AST_fromInteger a1) (AST_fromInteger a2)) = AST_fromInteger (a1+a2)
foldConstants (AST_minus  (AST_fromInteger a1) (AST_fromInteger a2)) = AST_fromInteger (a1-a2)
foldConstants (AST_mul    (AST_fromInteger a1) (AST_fromInteger a2)) = AST_fromInteger (a1*a2)
foldConstants (AST_negate (AST_fromInteger a1)) = AST_fromInteger (negate a1)
foldConstants (Free sig) = Free (fmap foldConstants sig)
foldConstants (Pure a  ) = Pure a
\end{lstlisting}

\begin{lstlisting}
constExpr :: Num a => a
constExpr = 4+2*(8-2)-1

-- ghci> fixAST foldConstants constExpr
-- (fromInteger 18)

constFunc :: Num a => a -> a -> a
constFunc x1 x2 = x1*2-(7-2)*x2

-- ghci> fixAST foldConstants (constFunc 2 3)
-- (fromInteger -11)
-- ghci> fixAST foldConstants (constFunc var1 var2)
-- ((var1*(fromInteger 2))-((fromInteger 5)*var2))
\end{lstlisting}

\subsection{The log-logistic function}

\begin{lstlisting}
stabAST :: Eq a => AST FloatingOrd a -> AST FloatingOrd a
stabAST
    (AST_log
        (AST_div
            (AST_fromInteger 1)
            (AST_plus
                (AST_fromInteger 1)
                (AST_exp
                    (AST_negate x)
                )
            )
        )
    )
    = logLogistic2 x
\end{lstlisting}

\section{Making floats lawful}

\subsection{Varieties and exact laws}

A class law can be defined by a pair of ASTs that substituted for each other without changing the semantics of a program.
For convenience, we assign each law a name and put it in the following data type.
\begin{lstlisting}
data Law alg = Law
    { lawName :: String
    , lhs     :: HomAST alg Var
    , rhs     :: HomAST alg Var
    }
\end{lstlisting}
A \emph{variety} is an algebraic structure that has a set of associated laws.
We can encode a variety in Haskell as:
\begin{lstlisting}
class HomFAlgebra alg => HomVariety alg where
    laws :: [Law alg]
\end{lstlisting}
Again, notice that instances of this class will be other type classes.

To encode the laws of a type class, we create an appropriate |HomVariety| instance.
The |Semigroup| class has the following instance defining what it means to be associative.
\begin{lstlisting}
instance HomVariety Semigroup where
    laws = [ Law
        { lawName = "associative"
        , lhs = (var1+var2)+var3
        , rhs = var1+(var2+var3)
        } ]
\end{lstlisting}

\subsection{Listing the laws}

When working with a large numerical hierarchy,
it is easy to forget the exact laws that a class must obey.
Wouldn't it be nice if there were an easy way to ask the repl what the class laws are?

With this encoding, there is.
First, we define a function that pretty prints a |Law|.
\begin{lstlisting}
printLaw :: Show (HomAST alg Var)
    => Law alg -> IO ()
printLaw law = do
    putStrLn ( "  "++lawName law++":"      )
    putStrLn ( "    lhs: "++show (lhs law) )
    putStrLn ( "    rhs: "++show (rhs law) )
\end{lstlisting}
Then we define a function that calls |printLaw| on each class law.
\begin{lstlisting}
printLaws :: forall alg.
    ( HomVariety alg
    , Show (HomAST alg Var)
    ) => Proxy alg -> IO ()
printLaws palg = do
    forM_ (laws::[Law alg]) printLaw
\end{lstlisting}
Now, if we enter the command |printLaws @Semigroup| into the REPL,
we get a nice reminder that every instance of |Semigroup| should be associative.

\subsection{The floating point topology}

Checking if two floating point numbers are equal is almost always a bad idea.
Better practice is to check if the two numbers are ``close enough'' for your application.
We can formalize this notion via topology.

In mathematics, a \emph{topology} is a set $X$ equipped with a family of subsets of $X$ called $\tau$ such that:
%In mathematics, a \emph{topology} is a set $X$ and a set of neighborhoods $\tau \subseteq 2^X$ such that:
\begin{enumerate}
\item
$\tau$ contains the empty set and the set $X$.
\item
$\tau$ is closed under an arbitrary number of unions.
\item
$\tau$ is closed under finitely many intersections.
\end{enumerate}
We call an element of $\tau$ a neighborhood.

To encode the notion of a topology in Haskell,
we first must encode the idea of a set.

\begin{lstlisting}
class Poset a where
    inf :: a -> a -> a
\end{lstlisting}

\begin{lstlisting}
class Poset (Neighborhood a) => Topology a
        where
    type Neighborhood a
    (==) :: a -> a -> (Neighborhood a -> Bool)
\end{lstlisting}

%It is well known that equality testing is almost always a bad idea for floating point numbers.
As mentioned before, floating point numbers are not strictly associative,
but they have a notion of being approximately associative.
In particular, when we call the |associator| function defined above on any three floating point values,
we can guarantee that the result is small relative to the input values.

\subsection{Topological test cases}

\section{Conclusion}

There are two more points worth discussion.

The first is efficiency.
The |stabilize| function is absurdly slow.
If you view the core, you can see two very different outputs.
But the output of |stabilize| is highly structured and ready to be optimized.
It's left as an exercise for the reader to write a GHC plugin to fix the problem.

The second is improved class hierarchies.
Most of the interesting alternative numeric hierarchies involve many non-homogeneous functions.
In particular, they use type classes.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

%\appendix
%\section{Appendix Title}
%
%This is the text of the appendix, if you need one.
%Most notably, it is not generic.
%
%\acks
%
%Acknowledgments, if needed.
%

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%

\bibliographystyle{abbrvnat}
\bibliography{paper}

\end{document}

### 2016-06-06 src/FAlgebra.hs

@@ -116,6 +116,25 @@ unsafeCoerceSigTag _ = unsafeCoerce
 
 ----------------------------------------
 
+data HaskTag {-cxt-} a b where
+    HaskTag ::
+        ( forall (s::[AT]). () --cxt s
+            => Proxy s
+--             -> Proxy cxt
+            -> Proxy a
+            -> Proxy b
+            -> App s a
+            -> App s b
+        ) -> HaskTag a b
+
+apHaskTag :: forall t a b . Proxy (t::[AT]) -> HaskTag a b -> App t a -> App t b
+apHaskTag pt (HaskTag f) = f pt (Proxy::Proxy a) (Proxy::Proxy b)
+
+class FunctorTag (f::[AT]->Type->Type) where
+    fmapTag :: HaskTag a b -> f t a -> f t b
+
+----------------------------------------
+
 data Free (f::[AT]->Type->Type) (t::[AT]) (a::Type) where
     FreeTag  :: TypeConstraints t a => f (s ': t) (Free f t a)  -> Free f (s ': t) a
     Free     :: TypeConstraints t a => f       t  (Free f t a)  -> Free f       t  a
@@ -143,9 +162,102 @@ eval (Pure    a) = a
 eval (Free    s) = runSig    (Proxy::Proxy a) $ mape eval s
 eval (FreeTag s) = runSigTag (Proxy::Proxy a) $ mape eval s
 
+class NoCxt (a::k)
+instance NoCxt a
+
+haskEval :: forall cxt alg t a.
+    ( alg a
+    , FAlgebra alg
+    , FunctorTag (Sig alg)
+    ) => HaskTag (Free (Sig alg) t a) (App t a)
+haskEval = HaskTag go
+    where
+        go :: forall (s::[AT]) .
+            ( alg a
+            , FAlgebra alg
+            , FunctorTag (Sig alg)
+            ) => Proxy s
+--               -> Proxy cxt
+              -> Proxy (Free (Sig alg) t a)
+              -> Proxy (App t a)
+              -> App s (Free (Sig alg) t a)
+              -> App s (App t a)
+        go _ _ _ expr = case appFree @s @t @a @alg expr of
+            (Pure a) -> appApp
+                (Proxy::Proxy s)
+                (Proxy::Proxy t)
+                (Proxy::Proxy a)
+                 a
+            (Free f) -> appApp
[…]

### 2016-06-06 src/FAlgebra98.hs

@@ -3,6 +3,18 @@
 {-# LANGUAGE TypeApplications #-}
 {-# LANGUAGE NoRebindableSyntax #-}
 {-# LANGUAGE TypeInType #-}
+{-# LANGUAGE RankNTypes #-}
+{-# LANGUAGE TypeOperators #-}
+{-# LANGUAGE TupleSections #-}
+{-# LANGUAGE StandaloneDeriving #-}
+{-# LANGUAGE ScopedTypeVariables #-}
+{-# LANGUAGE MultiParamTypeClasses #-}
+{-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE FlexibleInstances #-}
+{-# LANGUAGE GADTs #-}
+{-# LANGUAGE UndecidableInstances #-}
+{-# LANGUAGE ConstraintKinds #-}
+{-# LANGUAGE FlexibleContexts #-}
 
 {-# OPTIONS_GHC -Wno-simplifiable-class-constraints #-}
 
@@ -10,17 +22,17 @@ module FAlgebra98
     (
 
     -- * FAlgebra
-      FAlgebra98 (..)
-    , Free98 (..)
-    , Expr98
-    , eval98
-    , View98 (..)
+      FAlgebra (..)
+    , Free (..)
+    , AST
+    , runHomAST
+    , View (..)
 
     -- ** Template Haskell
-    , mkFAlgebra98
+    , mkFAlgebra
 
     -- * Variety
-    , Variety98 (..)
+    , Variety (..)
     , Law (..)
     , allLaws
     , subVars
@@ -37,6 +49,7 @@ module FAlgebra98
 
     -- ** Variables for generating laws
     , Var
+    , mkExprVar
     , var1
     , var2
     , var3
@@ -47,7 +60,7 @@ module FAlgebra98
     , printLaw
 
     -- ** Evaluating laws
-    , Testable98
+    , Testable
     , runTests
     , runAllTests
 
@@ -84,56 +97,56 @@ import Debug.Trace
 class (alg1 a, alg2 a) => And (alg1 :: Type -> Constraint) (alg2 :: Type -> Constraint) (a:: Type)
 instance (alg1 a, alg2 a) => And alg1 alg2 a
 
-instance (FAlgebra98 alg1, FAlgebra98 alg2) => FAlgebra98 (And alg1 alg2) where
+instance (FAlgebra alg1, FAlgebra alg2) => FAlgebra (And alg1 alg2) where
[…]

### 2016-06-06 src/FAlgebra98Example.hs

@@ -1,7 +1,7 @@
 {-# LANGUAGE PatternSynonyms #-}
 {-# LANGUAGE ViewPatterns #-}
 
-module FAlgebra98Example
+module HomFAlgebraExample
     where
 
 import Prelude hiding (Monoid (..),(-),(+),negate,(==))
@@ -22,9 +22,9 @@ import Lattice
 class Semigroup a where
     (+) :: a -> a -> a
 
-mkFAlgebra98 ''Semigroup
+mkHomFAlgebra ''Semigroup
 
-instance Variety98 Semigroup where
+instance HomVariety Semigroup where
     laws = [ Law
         { lawName = "associative"
         , lhs = (var1+var2)+var3
@@ -42,7 +42,7 @@ instance Approximate98 Semigroup Float where
             }
         ]
 
--- instance FAlgebra98 Semigroup where
+-- instance HomFAlgebra Semigroup where
 --     data Sig98 Semigroup a where
 --         Sig98_plus :: a -> a -> Sig98 Semigroup a
 --     runSig98 (Sig98_plus a1 a2) = a1+a2
@@ -56,26 +56,26 @@ instance Approximate98 Semigroup Float where
 -- instance Show a => Show (Sig98 Semigroup a) where
 --     show (Sig98_plus a1 a2) = show a1++"+"++show a2
 
--- pattern Expr98_plus ::
+-- pattern HomAST_plus ::
 --     ( View98 Semigroup alg
---     ) => Expr98 alg e
---       -> Expr98 alg e
---       -> Expr98 alg e
--- pattern Expr98_plus e1 e2 <- Free98 (unsafeExtractSig -> Sig98_plus e1 e2) where
---     Expr98_plus e1 e2 = Free98 $ embedSig $ Sig98_plus e1 e2
+--     ) => HomAST alg e
+--       -> HomAST alg e
+--       -> HomAST alg e
+-- pattern HomAST_plus e1 e2 <- Free98 (unsafeExtractSig -> Sig98_plus e1 e2) where
+--     HomAST_plus e1 e2 = Free98 $ embedSig $ Sig98_plus e1 e2
 
 -- instance View98 Semigroup alg => Semigroup (Free98 (Sig98 alg) a) where
 --     (+) e1 e2 = Free98 $ embedSig $ Sig98_plus e1 e2
---     (+) = Expr98_plus
+--     (+) = HomAST_plus
 
 --------------------
 
 class Semigroup a => Monoid a where
     zero :: a
 
-mkFAlgebra98 ''Monoid
[…]

### 2016-06-06 src/FAlgebraExample.hs

@@ -15,6 +15,14 @@ import Data.Proxy
 import qualified Prelude as P
 -- import GHC.Exts
 
+import Test.Tasty
+import Test.Tasty.Ingredients.Basic
+import Test.Tasty.Options
+import Test.Tasty.Runners
+import qualified Test.Tasty.SmallCheck as SC
+import qualified Test.Tasty.QuickCheck as QC
+import Test.QuickCheck hiding (Testable)
+
 import FAlgebra
 -- import FAlgebra98
 -- import Topology
@@ -29,6 +37,28 @@ class Poset a where
 
 mkFAlgebra ''Poset
 
+instance cxt '[] => FunctorTag cxt (Sig Poset) where
+    fmapTag f (Sig_inf e1 e2) = Sig_inf
+        (apHaskTag (Proxy::Proxy '[]) f e1)
+        (apHaskTag (Proxy::Proxy '[]) f e2)
+
+type instance ParentClasses '(Poset,t) = '[]
+
+instance Variety Poset where
+    laws =
+        [ Law
+            { lawName = "commutative"
+            , lhs = inf var1 var2
+            , rhs = inf var2 var1
+            }
+        , Law
+            { lawName = "associative"
+            , lhs = (var1 `inf`  var2) `inf` var3
+            , rhs =  var1 `inf` (var2  `inf` var3)
+            }
+        ]
+
+{-
 #define mkPoset(x) \
 instance Poset x where inf = P.min
 
@@ -50,6 +80,23 @@ class Poset a => LowerBounded a where
 
 mkFAlgebra ''LowerBounded
 
+instance FunctorTag (Sig LowerBounded) where
+    fmapTag f (Sig_LowerBounded_Poset_ s) = Sig_LowerBounded_Poset_ $ fmapTag f s
+    fmapTag f (Sig_minBound) = Sig_minBound
+
+type instance ParentClasses '(LowerBounded,t) =
+    '[ '(Poset,t)
+     ]
+
+instance Variety LowerBounded where
+    laws =
+        [ Law
+            { lawName = "minBound"
+            , lhs = inf minBound var1
+            , rhs = minBound
+            }
+        ]
+
 #define mkLowerBounded(x) \
 instance LowerBounded x where minBound = P.minBound
 
[…]

### 2016-06-06 src/HomPrelude.hs

@@ -1,0 +1,213 @@
+{-# LANGUAGE ViewPatterns #-}
+{-# LANGUAGE RankNTypes #-}
+{-# LANGUAGE TemplateHaskell #-}
+{-# LANGUAGE PatternSynonyms #-}
+{-# LANGUAGE DataKinds #-}
+{-# LANGUAGE TypeFamilies #-}
+
+module HomPrelude
+    where
+
+import FAlgebra98
+import Spatial98
+import Prelude
+import Data.Kind
+import Unsafe.Coerce
+
+
+mkFAlgebra ''Num
+mkFAlgebra ''Fractional
+mkFAlgebra ''Floating
+-- mkFAlgebra ''Eq
+instance FAlgebra Eq
+instance Functor (Sig Eq)
+instance Foldable (Sig Eq)
+instance Show (Sig Eq a)
+instance Eq (Sig Eq a)
+mkFAlgebra ''Ord
+
+class (Floating a, Ord a) => FloatingOrd a
+instance {-#OVERLAPPABLE#-} (Floating a, Ord a) => FloatingOrd a
+mkFAlgebra ''FloatingOrd
+
+-- type family DefaultAST a :: Type->Constraint
+-- type instance DefaultAST Double = Floating
+-- type instance DefaultAST Float  = Floating
+-- type instance DefaultAST (a -> b) = DefaultAST b
+-- type instance DefaultAST Int    = Num
+--
+-- type AST a = AST (DefaultAST a) a
+
+-- expLogLogistic1 :: Double -> Double
+-- expLogLogistic1 x = runHomAST $ logexpAST4 $ exp $ log $ 1/(1+exp(-Pure x))
+--
+-- expLogLogistic2 :: Double -> Double
+-- expLogLogistic2 x = 1/(1+exp(-x))
+
+associator :: Num a => a -> a -> a -> a
+associator a1 a2 a3
+    = ((a1+a2)+a3) - (a1+(a2+a3))
+
+--------------------------------------------------------------------------------
+
+toAST1
+    :: ( AST alg x -> y)
+    -> (         x -> y)
+toAST1 f x1 = f (Pure x1)
+
+transform1 :: forall alg x.
+    ( FAlgebra alg
+    , alg x
+    ) => (AST alg x -> AST alg x)
+      -> (AST alg x -> AST alg x)
+      -> (x -> x)
+transform1 go f x = runHomAST $ go $ f (Pure x)
+
+--------------------
+
+toAST2
[…]

### 2016-06-06 src/Lattice.hs

@@ -4,7 +4,7 @@
 module Lattice
     where
 
-import Prelude (fromInteger, Functor(..), Foldable(..))
+import Prelude (fromInteger, Functor(..), Foldable(..), Eq(..))
 import qualified Prelude as P
 import LocalPrelude
 import FAlgebra98
@@ -14,13 +14,13 @@ import FAlgebra98
 class Poset a where
     inf :: a -> a -> a
 
-mkFAlgebra98 ''Poset
+mkFAlgebra ''Poset
 
 infixr 3 &&
 (&&) :: Poset a => a -> a -> a
 (&&) = inf
 
-instance Variety98 Poset where
+instance Variety Poset where
     laws = [ associative (&&), commutative (&&), idempotent (&&) ]
 
 #define mkPoset(x) \
@@ -39,9 +39,9 @@ mkPoset(Char)
 class Poset a => LowerBounded a where
     lowerBound :: a
 
-mkFAlgebra98 ''LowerBounded
+mkFAlgebra ''LowerBounded
 
-instance Variety98 LowerBounded where
+instance Variety LowerBounded where
     laws =
         [ Law
             { lawName = "lowerBounded"
@@ -50,29 +50,24 @@ instance Variety98 LowerBounded where
             }
         ]
 
-# define mkLattice(x) \
-instance Lattice x where sup = P.max
+#define mkLowerBounded(x) \
+instance LowerBounded x where lowerBound = P.minBound
 
-mkLattice(Float)
-mkLattice(Double)
-mkLattice(Rational)
-mkLattice(Integer)
-mkLattice(Int)
-mkLattice(Bool)
-mkLattice(Char)
+mkLowerBounded(Bool)
+mkLowerBounded(Char)
 
 --------------------
 
 class Poset a => Lattice a where
     sup :: a -> a -> a
 
-mkFAlgebra98 ''Lattice
+mkFAlgebra ''Lattice
 
 infixr 3 ||
 (||) :: Lattice a => a -> a -> a
 (||) = sup
 
-instance Variety98 Lattice where
+instance Variety Lattice where
     laws =
         [ associative (||)
         , commutative (||)
@@ -89,20 +84,25 @@ instance Variety98 Lattice where
             }
         ]
 
-#define mkLowerBounded(x) \
-instance LowerBounded x where lowerBound = P.minBound
+# define mkLattice(x) \
+instance Lattice x where sup = P.max
 
-mkLowerBounded(Bool)
-mkLowerBounded(Char)
+mkLattice(Float)
[…]

### 2016-06-06 src/Spatial98.hs

@@ -34,10 +34,10 @@ data EpsLaw (alg :: * -> Constraint) a = EpsLaw
     , epsilon :: [a] -> Neighborhood a
     }
 
-class (Variety98 alg, alg a, Topology a) => Approximate98 alg a where
+class (HomVariety alg, alg a, Topology a) => Approximate98 alg a where
     epsLaws :: [EpsLaw alg a]
 
-instance {-#OVERLAPPABLE#-} (Variety98 alg, alg a, Topology a) => Approximate98 alg a where
+instance {-#OVERLAPPABLE#-} (HomVariety alg, alg a, Topology a) => Approximate98 alg a where
     epsLaws = []
 
 --------------------------------------------------------------------------------
@@ -53,9 +53,9 @@ lookupEpsLaw lawName = case lookup lawName lawmap of
         lawmap = map (\x -> (epsLawName x,x)) (epsLaws::[EpsLaw alg a])
 
 epsLaw2quickcheck :: forall (a :: *) alg.
-    ( Variety98 alg
+    ( HomVariety alg
     , alg a
-    , Testable98 a
+    , HomTestable a
     , Topology a
     ) => Proxy a -> Law alg -> TestTree
 epsLaw2quickcheck p law = QC.testProperty (lawName law) $ do
@@ -65,18 +65,18 @@ epsLaw2quickcheck p law = QC.testProperty (lawName law) $ do
 
 checkApproximate ::
     ( Approximate98 alg a
-    , Testable98 a
+    , HomTestable a
     ) => [a] -> Law alg -> Logic a
 checkApproximate as law
-    = (eval98 $ subVars (lhs law) varmap)
-   == (eval98 $ subVars (rhs law) varmap)
+    = (runHomAST $ subVars (lhs law) varmap)
+   == (runHomAST $ subVars (rhs law) varmap)
     where
         varmap = zip (toList (lhs law) ++ toList (rhs law)) as
 
 approxclass2tasty :: forall alg a.
-    ( Variety98 alg
+    ( HomVariety alg
     , alg a
-    , Testable98 a
+    , HomTestable a
     , Topology a
     ) => Proxy alg -> Proxy a -> TestTree
 approxclass2tasty palg pa = testGroup
@@ -84,9 +84,9 @@ approxclass2tasty palg pa = testGroup
     $ map (epsLaw2quickcheck pa) (laws::[Law alg])
 
 -- runTestsApprox :: forall alg a.
[…]

### 2016-06-06 src/TemplateHaskellUtils.hs

@@ -1,3 +1,4 @@
+{-# LANGUAGE TupleSections #-}
 
 module TemplateHaskellUtils
     where
@@ -30,6 +31,7 @@ renameClassMethod n = concatMap go $ nameBase n
         go '-' = "minus"
         go '.' = "dot"
         go '*' = "mul"
+        go '/' = "div"
         go '=' = "eq"
         go '>' = "gt"
         go '<' = "lt"
@@ -37,6 +39,10 @@ renameClassMethod n = concatMap go $ nameBase n
 
 isOperator :: String -> Bool
 isOperator str = not $ length str == length (renameClassMethod $ mkName $ str)
+
+isVarT :: TH.Type -> Bool
+isVarT (VarT _) = True
+isVarT _        = False
 
 -- | Given a type that stores a function:
 -- return a list of the arguments to that function,
@@ -65,17 +71,6 @@ subForall n (ForallT [v] _ t) = go t
         go (AppT t1 t2) = AppT (go t1) (go t2)
         go (VarT _) = VarT n
         go t = t
-
--- | Given a list of arguments passed to a function, return a list of patterns;
--- the inputs must be all type variables, otherwise an error will be thrown;
--- the Name parameter is only used for better error messages;
--- the output patterns will contain variables e1,e2,...eN where N is the length of the input
-args2pat :: Name -> [TH.Type] -> [Pat]
-args2pat sigName xs = go 0 xs
-    where
-        go i [] = []
-        go i (VarT n:xs) = (VarP $ mkName $ nameBase n ++ show i):go (i+1) xs
-        go i (x:xs) = error $ "function "++nameBase sigName++" has unsupported argument type "++show (ppr x)
 
 -- | Given a class name, find all the superclasses
 listSuperClasses :: Name -> Q [Name]

### 2016-06-06 src/Topology.hs

@@ -36,29 +36,29 @@ class
     isNeighbor :: a -> (a,Neighborhood a) -> Bool
     isNeighbor a1 (a2,n) = (a1==a2) n
 
-    infixr 4 ==
-    (==) :: a -> a -> Logic a
-    (==) a1 a2 n = isNeighbor a1 (a2,n)
-
-    infixr 4 /=
-    (/=) :: a -> a -> Logic a
-    (/=) = not (==)
-
-    infixr 4 <=
-    (<=) :: Poset a => a -> a -> Logic a
-    (<=) a1 a2 = inf a1 a2 == a1
-
-    infixr 4 <
-    (<) :: Poset a => a -> a -> Logic a
-    (<) a1 a2 = inf a1 a2 == a1 && a1 /= a2
-
-    infixr 4 >=
-    (>=) :: Lattice a => a -> a -> Logic a
-    (>=) a1 a2 = sup a1 a2 == a1
-
-    infixr 4 >
-    (>) :: Lattice a => a -> a -> Logic a
-    (>) a1 a2 = sup a1 a2 == a1 && a1 /= a2
+--     infixr 4 ==
+--     (==) :: a -> a -> Logic a
+--     (==) a1 a2 n = isNeighbor a1 (a2,n)
+--
+--     infixr 4 /=
+--     (/=) :: a -> a -> Logic a
+--     (/=) = not (==)
+--
+--     infixr 4 <=
+--     (<=) :: Poset a => a -> a -> Logic a
+--     (<=) a1 a2 = inf a1 a2 == a1
+--
+--     infixr 4 <
+--     (<) :: Poset a => a -> a -> Logic a
+--     (<) a1 a2 = inf a1 a2 == a1 && a1 /= a2
+--
+--     infixr 4 >=
+--     (>=) :: Lattice a => a -> a -> Logic a
+--     (>=) a1 a2 = sup a1 a2 == a1
+--
+--     infixr 4 >
+--     (>) :: Lattice a => a -> a -> Logic a
+--     (>) a1 a2 = sup a1 a2 == a1 && a1 /= a2
 
 -- NOTE:
 -- This law states that (==) is a Lattice and LowerBound homomorphism from (Neighborhood a) to Bool.

### 2016-06-07 paper/paper.tex
some paper changes, but mostly reorganizing the file system

@@
     language=Haskell,
-    basicstyle=\ttfamily,
-    keywordstyle=\ttfamily,
+    basicstyle=\ttfamily\footnotesize,
+    keywordstyle=\ttfamily\footnotesize,
+    %basicstyle=\footnotesize,
     literate={->}{{$\rightarrow$}}2 {>=}{{$\geq$}}2 {<-}{{$\leftarrow$}}2
@@
 \begin{abstract}
-%This pearl uses Haskell's type classes to enable a limited form of \emph{homoiconicity}.
-%%Specifically, we show how to convert any Haskell98 function that is parametric in its return type into an isomorphic abstract syntax tree.
-%Specifically, we show how to convert a certain set of polymorphic functions into isomorphic abstract syntax trees.
-%But this is really a pearl about the story of numerical programming in Haskell.
-
-%Numeric programming in Haskell sucks.
-%Actually, it sucks in every language because it requires floating point numbers,
-%and floating point arithmetic is so difficult to work with.
-%The goal of this pearl is to make numeric programming suck just a little bit less.
-%This statement is tautologically true---numeric programming requires floating point arithmetic, and working with floats sucks in every language.
-%I think this makes numeric programming in Haskell suck just a little bit less.
-
 Floating point numbers are dangerous.
@@
 seemingly correct mathematical formulae can give wildly inaccurate results.
-This pearl uses Haskell's type system to mitigate this danger.
+This pearl shows that Haskell's type system can mitigate this danger.
 
@@
 The construction does not use template Haskell, and so works at both compile and run time.
-We use these ASTs to solve two problems with floating point numbers:
+We will use these ASTs to solve two problems with floating point numbers:
 \begin{enumerate}
 \item
-We show how to write functions that automatically stabilize (and optimize!) mathematical formulae.
-This lets the programmer write naive floating point code;
-the compiler will worry about improving the numeric stability and speed later.
+We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
+%This lets one programmer write naive floating point code;
+%then an expert on numerical analysis can write code that automatically improves the naive code.
 \item
-We show how to specify type class laws for floating point numbers that can be easily and automatically tested.
-This is a difficult unsolved problem because floating point numbers do not obey many of the traditional algebraic laws like associativity.
+We will specify type class laws for floating point arithmetic that can be automatically tested.
+Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
 \end{enumerate}
-We demonstrate these techniques using GHC's builtin numeric hierarchy,
-but these techniques generalize to user defined numeric hierarchies as well.
-
-%First, we show how to convert a certain type of commonly used mathematical function into isomorphic abstract syntax trees.
-%Specifically, we show that a subset of Haskell functions (which we call \emph{homogeneous}) can easily be converted into an isomorphic abstract syntax tree.
-
-%I hate numerical programming in Haskell,
-%but I hate numerical programming in every language.
-%It requires floating point numbers,
-%which
-%
-%I hate floating point numbers.
-
-%This pearl provides a technique that makes working with approximate algebraic structures more convenient.
-%The prototypical example of an approximate structure is floating point arithmetic, which is famously not a semigroup because it is not associative.
-%The technique has two parts.
-%First, it lets us construct an FAlgebra (and hence a syntax tree) for a numerical class hierarchy.
-%Second, we show how to treat topology as an algebraic structure, and that floating point types have a useful topology.
-%This topology can be used to automatically construct properties for approximate algebraic structures that can be easily checked using quickcheck or smallcheck.
+We demonstrate these techniques using GHC's built-in numeric class hierarchy,
+but these techniques generalize to user defined class hierarchies as well.
 \end{abstract}
 
-\category{CR-number}{subcategory}{third-level}
+%\category{CR-number}{subcategory}{third-level}
 
@@
 
-\keywords
-floating point, homoiconic, type classes
+%\keywords
+%floating point, homoiconic, type classes
 
@@
 Imagine you're writing a machine learning library in Haskell.
-At some point, you'll need to implement the log-logistic function
+%You decide to start by implementing logistic regression.
+%The first step is to implement the log-logistic function
+%Logistic regression is one of the most important machine learning algorithms,
+%and to implement it, you'll need to implement the log-logistic function
+To implement the classic logistic regression algorithm,
+you begin by implementing the log-logistic function
 \begin{equation}
@@
 \end{equation}
-This function is used in logistic regression and many families of neural networks.
+%This function is used in logistic regression and many families of neural networks.
 As a first attempt, you might write:
@@
 Therefore, on an input of $-710$,
-|logLogisitic1| will return negative infinity when the true answer is actually very close to $-710$.
+|logLogisitic1| will take the |log| of 0,
+returning negative infinity when the true answer is actually very close to $-710$.
 Such an inaccurate answer is unacceptable,
@@
 
-Writing numerically stable functions is a notoriously difficult black art.
-(See \cite{goldberg1991,higham2002,hamming2012} for some classic introductions.)
-Fortunately, a simple technique can be used to improve the stability of the log-logistic function.
-First, check our input to see if it would overflow the |exp| function.
-If no overflow would occur, then calculate our function as normal;
-but if overflow would occur, then use an approximation that does not rely on a call to |exp|.
-Performing this transformation gives us:
-\begin{lstlisting}
-logLogistic2 :: (Floating x, Ord x) => x->x
-logLogistic2 x = if x < (-709)
-    then x
-    else log(1/1+exp(-x))
-\end{lstlisting}
+With some careful algebra, we can rewrite our log-logistic function in the following numerically stable form.
+\begin{lstlisting}
+logLogistic2 :: (Floating x, Ord x) => x -> x
+logLogistic2 x = m+log(1/(exp(m)+exp(-x+m)))
+    where
+        m = min 0 x
+\end{lstlisting}
+It is hardly obvious that these two functions are meant to calculate the same expression.
+The second implementation not only contains many more terms,
+it has a different type signature due to the use of the |min| function.
+
+Constructing these numerically stable functions is a notoriously difficult black art.\footnote{
+For some classic introductions, see \cite{goldberg1991,higham2002,hamming2012}.
+}
+Wouldn't it be nice to have a tool to automate this procedure?
+We'll construct a simple one in Section 3.
+
+The logistic regression algorithm operates on vector spaces;
+so you're next task is to get a proper vector space implementation.
+Since you plan to use the foreign function interface heavily,
+you decide to represent vectors as a pointer to a memory location.
+\begin{lstlisting}
+newtype Vector (dim::Nat) a = Vector (ForeignPtr a)
+\end{lstlisting}
+The |dim| type parameter encodes the dimension of the vector space.
+It makes sense to store this information at the type level because vector space operations don't change a vector's dimension.
+
+To make our |Vector| type useful,
+we must define how to perform the standard mathematical operations on it.
+You begin by making it an instance of |Num|.
+Since you're a careful programmer, you decide to write test cases to ensure your |Num| instance is correct.
+You begin by writing a QuickCheck property to ensure that your |(+)| function is properly associative.
+It works for |Int|s, but misteriously fails on |Float|s and |Double|s.
+%Unfortunately, the standard Haskell toolkit is not sufficient for the challenge.
+
+\newpage
+%Fortunately, a simple technique can be used to improve the stability of the log-logistic function.
+%First, check our input to see if it would overflow the |exp| function.
+%If no overflow would occur, then calculate our function as normal;
+%but if overflow would occur, then use an approximation that does not rely on a call to |exp|.
+%Performing this transformation gives us:
+%\begin{lstlisting}
+%logLogistic2 :: (Floating x, Ord x) => x->x
+%logLogistic2 x = if x < (-709)
+    %then x
+    %else log(1/1+exp(-x))
+%\end{lstlisting}
 %This new function is numerically stable, but we've introduced two new problems:
+
 We've made our function numerically stable, but we've introduced two new problems:
@@
 
-\begin{problem}[specialization]
-The \lstinline{exp} function when specialized to \lstinline{Float}s overflows on values greater than (approximately) 88.
-This means that |logLogistic2| will give incorrect answers for inputs between 88 and 709.
-To properly handle both |Float|s and |Double|s,
-we must write two separate functions.
-In practice, we may want to support many other numeric types as well.
-Half precision (16 bit) floats and fixed point arithmetic are becoming increasingly popular in machine learning applications due to their increased speed, smaller memory footprint, and lower energy consumption \cite{gupta2015}.
-We need to manually reimplement the log-logistic function specialized to every numeric type we want our library to support.
-\end{problem}
+%\begin{problem}[specialization]
+%The \lstinline{exp} function when specialized to \lstinline{Float}s overflows on values greater than (approximately) 88.
+%This means that |logLogistic2| will give incorrect answers for inputs between 88 and 709.
+%To properly handle both |Float|s and |Double|s,
+%we must write two separate functions.
+%In practice, we may want to support many other numeric types as well.
+%Half precision (16 bit) floats and fixed point arithmetic are becoming increasingly popular in machine learning applications due to their increased speed, smaller memory footprint, and lower energy consumption \cite{gupta2015}.
+%We need to manually reimplement the log-logistic function specialized to every numeric type we want our library to support.
+%\end{problem}
 
@@
 and all such manual transformations suffer from these problems.
-%(In fact, the log-logistic transformation described above might be considered an exceptionally simple case.)
-%There is a large body of literature on the theory of automating these transformations,
-%but relatively little on improving the programmer's experience of dealing with floating point numbers.
-%This pearl focuses on the latter.
 This pearl presents a method for alleviating these problems.
 Specifically, we will see how to write a function
-\begin{lstlisting}
-stabilize :: (Floating x, Ord x)
-    => (x -> x) -> (x -> x)
+%\begin{lstlisting}
+%stabilize :: (Floating x, Ord x)
+    %=> (x -> x) -> (x -> x)
+%\end{lstlisting}
+\begin{lstlisting}
+stabilize :: (Floating x,Ord x) => (x->x) -> (x->x)
 \end{lstlisting}
@@
 
-\section{Homogeneous Functions are Homoiconic}
+\section{Homogeneous Functions are homoiconic}
 Consider the simple numeric class hierarchy shown in Figure \ref{code:group}.
@@
 See for example \cite{swierstra2008}.
-The heart of our construction is the |HomFAlgebra| type class, defined as
+The heart of our construction is the |FAlgebra| type class, defined as
 \begin{lstlisting}
 class
-    ( Functor (HomSig alg)
-    , Foldable (HomSig alg)
-    ) => HomFAlgebra (alg::Type->Constraint)
-        where
-    data HomSig alg a
-    runHomSig :: alg a => HomSig alg a -> a
-\end{lstlisting}
-The |HomFAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
-This means that instances of |HomFAlgebra| will be other type classes (instead of types or type constructors).
-The |Hom| in |HomFAlgebra| stands for homogeneous.
-Most of the names used in this section will be prefixed with |Hom| to differentiate them from related names used in the later heterogeneous constructions.
-
-The easiest way to explain the |HomFAlgebra| class is by example.
+    ( Functor (Sig alg)
+    , Foldable (Sig alg)
+    ) => FAlgebra (alg::Type->Constraint)
+        where
+    data Sig alg a
+    runSig :: alg a => Sig alg a -> a
+\end{lstlisting}
+The |FAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
+This means that instances of |FAlgebra| will be other type classes (instead of types or type constructors).
+The || in |FAlgebra| stands for homogeneous.
+Most of the names used in this section will be prefixed with || to differentiate them from related names used in the later heterogeneous constructions.
+
+The easiest way to explain the |FAlgebra| class is by example.
 We'll start with the |Semigroup| type class defined above.
 \begin{lstlisting}
-instance HomFAlgebra Semigroup where
-    data HomSig Semigroup a = HomSig_plus a a
-    runHomSig (HomSig_plus a1 a2) = a1+a2
-\end{lstlisting}
-The |HomSig| associated data family is meant to encode what mathematicians call the \emph{signature} of the FAlgebra.
-That is, for every class method, |HomSig| should have a constructor with the same number of parameters.
+instance FAlgebra Semigroup where
+    data Sig Semigroup a = Sig_plus a a
+    runSig (Sig_plus a1 a2) = a1+a2
+\end{lstlisting}
+The |Sig| associated data family is meant to encode what mathematicians call the \emph{signature} of the FAlgebra.
+That is, for every class method, |Sig| should have a constructor with the same number of parameters.
 Since |Semigroup| has a single function that takes two parameters,
-the |HomSig Semigroup| data instance has a single constructor that takes two parameters.
-The |runHomSig| function evaluates the signature.
-For each constructor in |HomSig|, |runHomSig| should apply the corresponding function to the constructor's parameters.
-
-To complete the |HomFAlgebra Semigroup| instance, we must also define the following superclass instances.
-\begin{lstlisting}
-instance Functor (HomSig Semigroup) where
-    fmap f (HomSig_plus a1 a2)
-        = HomSig_plus (f a1) (f a2)
-
-instance Foldable (HomSig Semigroup) where
-    foldr f b (HomSig_plus a1 a2)
+the |Sig Semigroup| data instance has a single constructor that takes two parameters.
+The |runSig| function evaluates the signature.
+For each constructor in |Sig|, |runSig| should apply the corresponding function to the constructor's parameters.
+
+To complete the |FAlgebra Semigroup| instance, we must also define the following superclass instances.
+\begin{lstlisting}
+instance Functor (Sig Semigroup) where
+    fmap f (Sig_plus a1 a2)
+        = Sig_plus (f a1) (f a2)
+
+instance Foldable (Sig Semigroup) where
+    foldr f b (Sig_plus a1 a2)
         = f a2 (f a1 b)
@@
 Instances get slightly more complicated when the parameter class has superclasses.
-In this case, the |HomSig| data instance must also have a constructor for each superclass that stores the superclass's signature.
+In this case, the |Sig| data instance must also have a constructor for each superclass that stores the superclass's signature.
 This can be seen in the |Monoid| and |Group| instances below.
-%The |HomFAlgebra| instances of |Monoid| and |Group| demonstrate this.
-\begin{lstlisting}
-instance HomFAlgebra Monoid where
-    data HomSig Monoid a
-        = HomSig_zero
-        | HomSig_Semigroup_Monoid
-            (HomSig Semigroup a)
-    runSig HomSig_zero = zero
-    runSig (HomSig_Semigroup_Monoid s)
-        = runHomSig s
-
-instance HomFAlgebra Group where
-    data HomSig Group a
-        = HomSig_negate a
-        | HomSig_minus a a
-        | HomSig_Monoid
-            (HomSig_Monoid_Group a)
-    runSig (HomSig_negate a) = a
-    runSig (HomSig_minus a1 a2) = a1-a2
-    runSig (HomSig_Monoid_Group s)
-        = runHomSig s
-\end{lstlisting}
-Notice that the data instance |HomSig Group| does not need a constructor for the |Semigroup| signature because this is included in the |Monoid| signature.
+%The |FAlgebra| instances of |Monoid| and |Group| demonstrate this.
+\begin{lstlisting}
+instance FAlgebra Monoid where
+    data Sig Monoid a
+        = Sig_zero
+        | Sig_Semigroup_Monoid
+            (Sig Semigroup a)
+    runSig Sig_zero = zero
+    runSig (Sig_Semigroup_Monoid s)
+        = runSig s
+
+instance FAlgebra Group where
+    data Sig Group a
+        = Sig_negate a
+        | Sig_minus a a
+        | Sig_Monoid
+            (Sig_Monoid_Group a)
+    runSig (Sig_negate a) = a
+    runSig (Sig_minus a1 a2) = a1-a2
+    runSig (Sig_Monoid_Group s)
+        = runSig s
+\end{lstlisting}
+Notice that the data instance |Sig Group| does not need a constructor for the |Semigroup| signature because this is included in the |Monoid| signature.
 The corresponding |Functor| and |Foldable| instances have been omitted for brevity,
@@
 
-%The procedure for generating |HomFAlgebra| instances is purely mechanical.
+%The procedure for generating |FAlgebra| instances is purely mechanical.
 %The ?? library automates the process with the |mkHomFAlgebra| template Haskell function.
@@
 \begin{lstlisting}
-type HomAST alg a = HomFree (HomSig alg) a
-
-data HomFree f a
-    = HomPure a
-    | HomFree (f (HomFree f a))
+type AST alg a = Free (Sig alg) a
+
+data Free f a
+    = Pure a
+    | Free (f (Free f a))
 \end{lstlisting}
 Intuitively,
-the |HomPure| constructor represents a leaf in the AST,
-and the |HomFree| constructor represents a branch.
-The parameter to |HomFree| is filled by the appropriate |HomSig| data instance.
-The constructor used for the |HomSig| instance corresponds to an operation,
+the |Pure| constructor represents a leaf in the AST,
+and the |Free| constructor represents a branch.
+The parameter to |Free| is filled by the appropriate |Sig| data instance.
+The constructor used for the |Sig| instance corresponds to an operation,
 and the parameters to the constructor will contain ASTs nested recursively.
@@
 For example, consider the expression |(1+2)+3|.
-We can create a corresponding AST using |Semigroup|'s |HomFAlgebra| instance.
-\begin{lstlisting}
-HomFree
-  (HomSig_plus
-    (HomFree
-      (HomSig_plus
-        (HomPure 1)
-        (HomPure 2)
-      )
-    )
-    (HomPure 3)
+We can create a corresponding AST using |Semigroup|'s |FAlgebra| instance.
+\begin{lstlisting}
+Free
+  (Sig_plus
+    (Free
+      (Sig_plus
+        (Pure 1)
+        (Pure 2)
+      )
+    )
+    (Pure 3)
   )
@@
 \begin{lstlisting}
-HomFree
-  (HomSig_Monoid_Group
-    (HomSig_Semigroup_Monoid
-      (HomSig_plus
-        (HomFree
-          (HomSig_Monoid_Group
-            (HomSig_Semigroup_Monoid
-              (HomSig_plus
-                (HomPure 1)
-                (HomPure 2)
-              )
-            )
-          )
-        )
-        (HomPure 3)
-      )
-    )
-  )
-\end{lstlisting}
-Notice that the |HomSig_plus| constructor must be embedded into the |Group|'s |HomSig| with calls to |HomSig_Monoid_Group| and |HomSig_Semigroup_Monoid|.
+Free
+  (Sig_Monoid_Group
+    (Sig_Semigroup_Monoid
+      (Sig_plus
+        (Free
+          (Sig_Monoid_Group
+            (Sig_Semigroup_Monoid
+              (Sig_plus
+                (Pure 1)
+                (Pure 2)
+              )
+            )
+          )
+        )
+        (Pure 3)
+      )
+    )
+  )
+\end{lstlisting}
+Notice that the |Sig_plus| constructor must be embedded into the |Group|'s |Sig| with calls to |Sig_Monoid_Group| and |Sig_Semigroup_Monoid|.
 
 Constructing these ASTs by hand is a tedious process.
-There is a considerable amount of boilerplate in the |Group| AST above as a result of the need to embed the |HomSig_plus| constructor.
+There is a considerable amount of boilerplate in the |Group| AST above as a result of the need to embed the |Sig_plus| constructor.
 We can avoid this boilerplate with the following type class.
@@
 class
-    ( HomFAlgebra alg1
-    , HomFAlgebra alg2
-    ) => HomView alg1 alg2
+    ( FAlgebra alg1
+    , FAlgebra alg2
+    ) => View alg1 alg2
         where
     embedHomSig
-        :: HomSig alg1 a -> HomSig alg2 a
+        :: Sig alg1 a -> Sig alg2 a
     unsafeExtractHomSig
-        :: HomSig alg2 a -> HomSig alg1 a
-\end{lstlisting}
-Instances of the |HomView| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
+        :: Sig alg2 a -> Sig alg1 a
+\end{lstlisting}
+Instances of the |View| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
 all such relationships should have a corresponding instance.
 For example, |Semigroup| is a superclass of |Group|,
-so we should have a |HomView Semigroup Group| instance.
-\begin{lstlisting}
-instance HomView Semigroup Group
+so we should have a |View Semigroup Group| instance.
+\begin{lstlisting}
+instance View Semigroup Group
     embedHomSig s
-        = HomSig_Monoid_Group (embedHomSig s)
-    unsafeExtractHomSig (HomSig_Monoid_Group s)
+        = Sig_Monoid_Group (embedHomSig s)
+    unsafeExtractHomSig (Sig_Monoid_Group s)
         = unsafeExtractHomSig s
 \end{lstlisting}
-Because of this property, we know that the |HomSig alg1 a| data instance can always be embedded into a |HomSig alg2 a| data instance.
-But it is only sometimes true that we can extract a |HomSig alg1 a|,
+Because of this property, we know that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance.
+But it is only sometimes true that we can extract a |Sig alg1 a|,
 which is why the corresponding function is labeled |unsafe|.
 
-Armed with the |HomView| class, we can create a uniform representation for all class's ASTs.
+Armed with the |View| class, we can create a uniform representation for all class's ASTs.
 The expression |(1+2)+3| can now be represented as:
 \begin{lstlisting}
-HomFree
+Free
   (embedHomSig
-    (HomSig_plus
-      (HomFree
+    (Sig_plus
+      (Free
         (embedHomSig
-          (HomSig_plus
-            (HomPure 1)
-            (HomPure 2)
-          )
-        )
-      )
-      (HomPure 3)
+          (Sig_plus
+            (Pure 1)
+            (Pure 2)
+          )
+        )
+      )
+      (Pure 3)
     )
@@
 %\begin{lstlisting}
-%instance HomView alg alg where
+%instance View alg alg where
     %embedSig = id
@@
 Again, we'll approach the procedure by example.
-We can write a |Semigroup| instance for |HomAST Semigroup a| as:
+We can write a |Semigroup| instance for |AST Semigroup a| as:
 \begin{lstlisting}
 instance
-    ( HomView Semigroup alg
-    ) => Semigroup (HomFree (HomSig alg) a)
-        where
-    (+) e1 e2 = HomFree $ embedHomSig $
-            HomSig_plus e1 e2
+    ( View Semigroup alg
+    ) => Semigroup (Free (Sig alg) a)
+        where
+    (+) e1 e2 = Free $ embedHomSig $
+            Sig_plus e1 e2
 \end{lstlisting}
 For each function in the type class,
-we simply embed the function's constructor into the AST by prepending the expression |HomFree $ embedHomSig $|.
+we simply embed the function's constructor into the AST by prepending the expression |Free $ embedHomSig $|.
 The |Monoid| and |Group| instances follow a similar pattern.
@@
 instance
-    ( HomView Semigroup alg
-    , HomView Monoid alg
-    ) => Monoid (HomFree (HomSig alg) a)
-        where
-    zero = HomFree $ embedHomSig $
-        HomSig_zero
+    ( View Semigroup alg
+    , View Monoid alg
+    ) => Monoid (Free (Sig alg) a)
+        where
+    zero = Free $ embedHomSig $
+        Sig_zero
 
 instance
-    ( HomView Semigroup alg
-    , HomView Monoid alg
-    , HomView Group alg
-    ) => Group (HomFree (HomSig alg) a)
-        where
-    negate e  = HomFree $ embedHomSig $
-        HomSig_negate e
-    (-) e1 e2 = HomFree $ embedHomSig $
-        HomSig_minus e1 e2
+    ( View Semigroup alg
+    , View Monoid alg
+    , View Group alg
+    ) => Group (Free (Sig alg) a)
+        where
+    negate e  = Free $ embedHomSig $
+        Sig_negate e
+    (-) e1 e2 = Free $ embedHomSig $
+        Sig_minus e1 e2
 \end{lstlisting}
@@
 \begin{lstlisting}
-(HomPure 1+HomPure 2)+HomPure 3
-    :: HomAST Semigroup a
+(Pure 1+Pure 2)+Pure 3
+    :: AST Semigroup a
 \end{lstlisting}
@@
 \begin{lstlisting}
-(HomPure 1+HomPure 2)+HomPure 3
-    :: HomAST Group a
+(Pure 1+Pure 2)+Pure 3
+    :: AST Group a
 \end{lstlisting}
@@
 The simplest task is to convert them into a |String| for display.
-This is an easy procedure of writing |Show| instances for the |HomSig| and |HomFree| data types.
-
-The are no tricks in the |Show| instance for |Semigroup|'s |HomSig|.
+This is an easy procedure of writing |Show| instances for the |Sig| and |Free| data types.
+
+The are no tricks in the |Show| instance for |Semigroup|'s |Sig|.
 \begin{lstlisting}
 instance Show a =>
-    Show (HomSig Semigroup a)
-        where
-    show (HomSig_plus a1 a2)
+    Show (Sig Semigroup a)
+        where
+    show (Sig_plus a1 a2)
         = show a1++"+"++show a2
@@
 instance Show a =>
-    Show (HomSig Monoid a)
-        where
-    show (HomSig_Semigroup_Monoid s)
+    Show (Sig Monoid a)
+        where
+    show (Sig_Semigroup_Monoid s)
         = show s
-    show (HomSig_zero) = "zero"
+    show (Sig_zero) = "zero"
 
 instance Show a =>
-    Show (HomSig Group a)
-        where
-    show (HomSig_Monoid_Group s)
+    Show (Sig Group a)
+        where
+    show (Sig_Monoid_Group s)
         = show s
-    show (HomSig_negate a)
+    show (Sig_negate a)
         = "negate "++show a
-    show (HomSig_minus a1 a2)
+    show (Sig_minus a1 a2)
         = show a1++"-"++show a2
 \end{lstlisting}
-The |HomFree| type's |Show| instance is similarly straightforward..
-\begin{lstlisting}
-instance (Show a, Show (f (HomFree f a)))
-    => Show (HomFree f a)
-        where
-    show (HomPure a) = show a
-    show (HomFree f) = "("++show f++")"
+The |Free| type's |Show| instance is similarly straightforward..
+\begin{lstlisting}
+instance (Show a, Show (f (Free f a)))
+    => Show (Free f a)
+        where
+    show (Pure a) = show a
+    show (Free f) = "("++show f++")"
 \end{lstlisting}
@@
 This data type is called the free monad because as long as the parameter |f| is a |Functor|,
-then |HomFree| has a valid |Monad| instance.
+then |Free| has a valid |Monad| instance.
 Since all |Monad|s are by necessity also |Functor|s,
-|HomFree| has a valid |Functor| instance as well.
-\begin{lstlisting}
-instance Functor f => Functor (HomFree f) where
-    fmap g (HomPure a) = HomPure (g a)
-    fmap g (HomFree f)
-        = HomFree (fmap (fmap g) f)
+|Free| has a valid |Functor| instance as well.
+\begin{lstlisting}
+instance Functor f => Functor (Free f) where
+    fmap g (Pure a) = Pure (g a)
+    fmap g (Free f)
+        = Free (fmap (fmap g) f)
 \end{lstlisting}
@@
 \begin{lstlisting}
-runHomAST :: (HomFAlgebra alg, alg a)
-    => HomAST alg a -> a
-runHomAST (HomPure a) = a
-runHomAST (HomFree f) = runHomSig (fmap evalHom f)
-\end{lstlisting}
-If our syntax tree already consists of just a single value (i.e. it is the |HomePure| constructor),
+runAST :: (FAlgebra alg, alg a)
+    => AST alg a -> a
+runAST (Pure a) = a
+runAST (Free f) = runSig (fmap evalHom f)
+\end{lstlisting}
+If our syntax tree already consists of just a single value (i.e. it is the |ePure| constructor),
 then we just return that value.
-Otherwise, we use the |Functor| instance to recursively convert the |HomSig|s into types of |a|, then a call to |runHomSig| combines the results.
+Otherwise, we use the |Functor| instance to recursively convert the |Sig|s into types of |a|, then a call to |runSig| combines the results.
 
@@
 \begin{lstlisting}
-ghci> let ast=associator (HomPure a1) (HomPure a2) (HomPure a3)
+ghci> let ast=associator (Pure a1) (Pure a2) (Pure a3)
 \end{lstlisting}
@@
 \begin{lstlisting}
-ghci> runHomAST ast
+ghci> runAST ast
 
@@
       -> (x -> x)
-transform1 go f x = runHomAST $ go $ f (Pure x)
+transform1 go f x = runAST $ go $ f (Pure x)
 \end{lstlisting}
@@
 
-ghci> fixAST stabilizeAST (testFunc2 var1 :: AST Floating Var)
+ghci> fixAST stabilizeAST
+    (testFunc2 var1 :: AST Floating Var)
 ((fromInteger 1)+var1)
@@
 
+Many type classes in Haskell come with an associated set of class laws.
+Usually, these laws are stated informally via haddock documentation.
+For example, the haddock documentation for the |Functor| class
+\begin{lstlisting}
+class Functor f where
+    fmap :: (a -> b) -> f a -> f b
+\end{lstlisting}
+states that all instances should satisfy the following two laws.
+\begin{lstlisting}
+fmap id  ==  id
+fmap (f . g)  ==  fmap f . fmap g
+\end{lstlisting}
+The type classes in the numeric hierarchy, however, do not have any laws.
+%For example, the haddock documentation for the |Monoid| class
+
+In this section, we'll show how to encode the laws for any |FAlgebra| directly in Haskell.
+This makes verifying the laws easy.
+Then we'll show how to make these laws topological.
+
 \subsection{Varieties and exact laws}
 
-A class law can be defined by a pair of ASTs that substituted for each other without changing the semantics of a program.
+A \emph{variety} is an algebraic structure that has a set of associated laws.
+A law can be defined by a pair of ASTs.
+These ASTs are equivalent in the sense that they can be substituted for each other without changing the program's semantics.
 For convenience, we assign each law a name and put it in the following data type.
+We will represent a law with the following data type.
 \begin{lstlisting}
@@
     { lawName :: String
-    , lhs     :: HomAST alg Var
-    , rhs     :: HomAST alg Var
+    , lhs     :: AST alg Var
+    , rhs     :: AST alg Var
     }
 \end{lstlisting}
-A \emph{variety} is an algebraic structure that has a set of associated laws.
-We can encode a variety in Haskell as:
-\begin{lstlisting}
-class HomFAlgebra alg => HomVariety alg where
+%The records |lhs| and |rhs| are meant to indicate the left and right hand sides of the equality the law is encoding.
+%Specifically, in any expression, we should be able to substitute the these two expressions for each other without changing the expression's semantics.
+Now we can encode a variety in Haskell as a list of associated laws.
+\begin{lstlisting}
+class FAlgebra alg => Variety alg where
     laws :: [Law alg]
 \end{lstlisting}
-Again, notice that instances of this class will be other type classes.
-
-To encode the laws of a type class, we create an appropriate |HomVariety| instance.
-The |Semigroup| class has the following instance defining what it means to be associative.
-\begin{lstlisting}
-instance HomVariety Semigroup where
-    laws = [ Law
+To encode the laws of a particular type class,
+we create an appropriate |Variety| instance.
+There are many possible laws we might want a |Num| instance to obey.
+We will use small, deliberately inadequate set of laws to illustrate the point.
+The instance below encodes that the |(+)| function should be both associative and commutative.
+\begin{lstlisting}
+instance Variety Num where
+    laws =
+        [ Law
             { lawName = "associative"
@@
             , rhs = var1+(var2+var3)
-        } ]
-\end{lstlisting}
-
-\subsection{Listing the laws}
-
+            }
+        , Law
+            { lawName = "commutative"
+            , lhs = var1*var2
+            , rhs = var2*var1
+            }
+        ]
+\end{lstlisting}
+
+\subsubsection{Listing the laws}
 When working with a large numerical hierarchy,
@@
 
-With this encoding, there is.
+Since we've encoded our class laws directly in Haskell, we can.
 First, we define a function that pretty prints a |Law|.
 \begin{lstlisting}
-printLaw :: Show (HomAST alg Var)
+printLaw :: Show (AST alg Var)
     => Law alg -> IO ()
@@
 printLaws :: forall alg.
-    ( HomVariety alg
-    , Show (HomAST alg Var)
+    ( Variety alg
+    , Show (AST alg Var)
     ) => Proxy alg -> IO ()
@@
 \end{lstlisting}
-Now, if we enter the command |printLaws @Semigroup| into the REPL,
-we get a nice reminder that every instance of |Semigroup| should be associative.
+Now we can ask at the REPL for our class laws as follows
+\begin{lstlisting}
+ghci> printLaws (Proxy::Proxy Num)
+\end{lstlisting}
+which gives the following output
+\begin{lstlisting}
+  associative:
+    lhs: ((var1+var2)+var3)
+    rhs: (var1+(var2+var3))
+  commutative:
+    lhs: (var1*var2)
+    rhs: (var2*var1)
+\end{lstlisting}
+
+\subsubsection{Testing the laws (Part I)}
+We can also use the first class nature of our laws to automatically test that instances obey them.
+
+First, we need to be able to convert an AST of |Var|s into an AST of arbitrary types by substituting.
+\begin{lstlisting}
+subVars :: FAlgebra alg => AST alg Var -> [(Var,a)] -> AST alg a
+subVars expr varmap = fmap go expr
+    where
+        go var = case lookup var varmap of
+            Just a -> a
+\end{lstlisting}
+
+Armed with this function, we can convert any law into a quickcheck property.
+\begin{lstlisting}
+law2property :: forall (a :: Type) alg.
+    ( FAlgebra alg
+    , alg a
+    , Eq a
+    , Arbitrary a
+    ) => Proxy a -> Law alg -> Gen Bool
+law2property p law = do
+    as <- infiniteListOf (arbitrary::Gen a)
+    let varmap = zip (toList (lhs law) ++ toList (rhs law)) as
+    return $ (runAST $ subVars (lhs law) varmap)
+          == (runAST $ subVars (rhs law) varmap)
+\end{lstlisting}
+%stopzone
+And we can test all the laws for a given class instance at once.
+\begin{lstlisting}
+class2quickcheck :: forall a alg.
+    ( Variety alg
+    , alg a
+    , Eq a
+    , Arbitrary a
+    ) => Proxy a -> Proxy alg -> IO ()
+class2quickcheck pa _ = forM_ (laws::[Law alg]) $ \law -> do
+    putStr $ lawName law++": "
+    quickCheck $ law2property pa law
+\end{lstlisting}
+%stopzone
+Now in GHCi:
+\begin{lstlisting}
+ghci> class2quickcheck (Proxy::Proxy Int) (Proxy::Proxy Num)
+associative: +++ OK, passed 100 tests.
+commutative: +++ OK, passed 100 tests.
+\end{lstlisting}
+But as mentioned before, |Float|s don't strictly obey the associative property.
+\begin{lstlisting}
+ghci> > class2quickcheck (Proxy::Proxy Float) (Proxy::Proxy Num)
+associative: *** Failed! Falsifiable (after 2 tests):
+commutative: +++ OK, passed 100 tests
+\end{lstlisting}
+We'll need something a bit more subtle for them.
 

### 2016-06-07 Bug1.hs

@@ -1,8 +1,0 @@
-{-# LANGUAGE PatternSynonyms #-}
-{-# LANGUAGE ViewPatterns #-}
-
-data Foo a = Foo a a
-
-pattern A a1 a2 = Foo a1 a2
-pattern B a1 a2 = A a1 a2
-

### 2016-06-07 old/Bug1.hs

@@ -1,0 +1,14 @@
+-- {-# LANGUAGE PatternSynonyms #-}
+-- {-# LANGUAGE ViewPatterns #-}
+--
+-- data Foo a = Foo a a
+--
+-- pattern A a1 a2 = Foo a1 a2
+-- pattern B a1 a2 = A a1 a2
+
+{-# UndecidableSuperClasses #-}
+{-# AllowAmbiguousTypes #-}
+{-# TypeFamilies #-}
+
+class A (T a) => A a where
+    type T a

### 2016-06-07 old/Bug2.hs

@@ -1,0 +1,12 @@
+{-# LANGUAGE UndecidableSuperClasses #-}
+{-# LANGUAGE AllowAmbiguousTypes #-}
+{-# LANGUAGE TypeFamilies #-}
+
+class A (T a) => A a where
+    type T a
+
+-- test1 :: forall a. A a => ()
+-- test1 = ()
+
+test2 :: A a => proxy a -> ()
+test2 _ = ()

### 2016-06-07 old/Bug3.hs

@@ -1,0 +1,5 @@
+import Data.Typeable
+import Data.Proxy
+import Prelude
+
+x = typeRep (Proxy::Proxy '(Int,Int))

### 2016-06-07 src/Heterogeneous/Topology.hs

@@ -36,10 +36,10 @@ class
     isNeighbor :: a -> (a,Neighborhood a) -> Bool
     isNeighbor a1 (a2,n) = (a1==a2) n
 
---     infixr 4 ==
---     (==) :: a -> a -> Logic a
---     (==) a1 a2 n = isNeighbor a1 (a2,n)
---
+    infixr 4 ==
+    (==) :: a -> a -> Logic a
+    (==) a1 a2 n = isNeighbor a1 (a2,n)
+
 --     infixr 4 /=
 --     (/=) :: a -> a -> Logic a
 --     (/=) = not (==)

### 2016-06-07 src/Homogeneous/FAlgebra.hs

@@ -18,14 +18,14 @@
 
 {-# OPTIONS_GHC -Wno-simplifiable-class-constraints #-}
 
-module FAlgebra98
+module Homogeneous.FAlgebra
     (
 
     -- * FAlgebra
       FAlgebra (..)
     , Free (..)
     , AST
-    , runHomAST
+    , runAST
     , View (..)
 
     -- ** Template Haskell
@@ -61,11 +61,21 @@ module FAlgebra98
 
     -- ** Evaluating laws
     , Testable
+
+    -- *** Exact
     , runTests
     , runAllTests
 
+    , law2property
+    , class2quickcheck
+
+    , law2tasty
     , class2tasty
     , superclasses2tasty
+
+    -- *** Approximate
+    , isNeighbor
+    , Approximate (..)
     )
     where
 
@@ -193,10 +203,10 @@ instance (Functor f, Foldable f) => Foldable (Free f) where
 
 type AST (alg :: Type -> Constraint) a = Free (Sig alg) a
 
-{-# INLINABLE runHomAST #-}
-runHomAST :: (FAlgebra alg, alg a) => AST alg a -> a
-runHomAST (Pure a) = a
-runHomAST (Free f) = runSig (fmap runHomAST f)
+{-# INLINABLE runAST #-}
+runAST :: (FAlgebra alg, alg a) => AST alg a -> a
+runAST (Pure a) = a
+runAST (Free f) = runSig (fmap runAST f)
 
 class (FAlgebra alg1, FAlgebra alg2) => View alg1 alg2 where
     embedSig         :: Sig alg1 a -> Sig alg2 a
@@ -352,16 +362,39 @@ subVars expr varmap = fmap go expr
         go var = case lookup var varmap of
             Just a -> a
 
-law2quickcheck :: forall (a :: Type) alg.
+law2property :: forall (a :: Type) alg.
+    ( FAlgebra alg
+    , alg a
+    , Eq a
+    , Arbitrary a
+    ) => Proxy a -> Law alg -> Gen Bool
+law2property p law = do
+    as <- infiniteListOf (arbitrary::Gen a)
+    let varmap = zip (toList (lhs law) ++ toList (rhs law)) as
+    return $ (runAST $ subVars (lhs law) varmap)
+          == (runAST $ subVars (rhs law) varmap)
+
+class2quickcheck :: forall a alg.
+    ( Variety alg
+    , alg a
+    , Eq a
+    , Arbitrary a
+    ) => Proxy a -> Proxy alg -> IO ()
[…]

### 2016-06-07 src/Homogeneous/PreludeInstances.hs

@@ -5,15 +5,11 @@
 {-# LANGUAGE DataKinds #-}
 {-# LANGUAGE TypeFamilies #-}
 
-module HomPrelude
+module Homogeneous.PreludeInstances
     where
 
-import FAlgebra98
-import Spatial98
+import Homogeneous.FAlgebra
 import Prelude
-import Data.Kind
-import Unsafe.Coerce
-
 
 mkFAlgebra ''Num
 mkFAlgebra ''Fractional
@@ -30,23 +26,43 @@ class (Floating a, Ord a) => FloatingOrd a
 instance {-#OVERLAPPABLE#-} (Floating a, Ord a) => FloatingOrd a
 mkFAlgebra ''FloatingOrd
 
--- type family DefaultAST a :: Type->Constraint
--- type instance DefaultAST Double = Floating
--- type instance DefaultAST Float  = Floating
--- type instance DefaultAST (a -> b) = DefaultAST b
--- type instance DefaultAST Int    = Num
---
--- type AST a = AST (DefaultAST a) a
-
--- expLogLogistic1 :: Double -> Double
--- expLogLogistic1 x = runHomAST $ logexpAST4 $ exp $ log $ 1/(1+exp(-Pure x))
---
--- expLogLogistic2 :: Double -> Double
--- expLogLogistic2 x = 1/(1+exp(-x))
-
 associator :: Num a => a -> a -> a -> a
 associator a1 a2 a3
     = ((a1+a2)+a3) - (a1+(a2+a3))
+
+--------------------------------------------------------------------------------
+
+instance Variety Num where
+    laws =
+        [ Law
+            { lawName = "associative"
+            , lhs = (var1+var2)+var3
+            , rhs = var1+(var2+var3)
+            }
+        , Law
+            { lawName = "commutative"
+            , lhs = var1*var2
+            , rhs = var2*var1
+            }
+        ]
+
+instance Approximate Num Float where
+    lawError (Law "associative" _ _) [a1,a2,a3] = abs $ ((a1+a2)+a3)-(a1+(a2+a3))
+
+----------------------------------------
+
+data Vec3 a = Vec3 {a::a, b::a, c::a}
+
+instance Num a => Num (Vec3 a) where
+    (Vec3 a1 a2 a3)+(Vec3 b1 b2 b3) = Vec3 (a1+b1) (a2+b2) (a3+b3)
[…]

### 2016-06-08 paper/paper.tex
restructured paper; lots of changes to the heterogeneous functions for better pattern matching

@@
 %\title{What's the point of homoiconicity?}
-\title{The (floating) point of homoiconicity}
+%\title{The (floating) point of homoiconicity}
+\title{Homoiconicity Without Template Haskell}
 \subtitle{Functional Pearl}
@@
 \begin{abstract}
-Floating point numbers are dangerous.
-Due to their limited precision,
-seemingly correct mathematical formulae can give wildly inaccurate results.
-This pearl shows that Haskell's type system can mitigate this danger.
-
-The main observation is that many numerical Haskell functions are \emph{homoiconic}.
-That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
-The construction does not use template Haskell, and so works at both compile and run time.
-We will use these ASTs to solve two problems with floating point numbers:
-\begin{enumerate}
-\item
-We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
-%This lets one programmer write naive floating point code;
-%then an expert on numerical analysis can write code that automatically improves the naive code.
-\item
-We will specify type class laws for floating point arithmetic that can be automatically tested.
-Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
-\end{enumerate}
-We demonstrate these techniques using GHC's built-in numeric class hierarchy,
-but these techniques generalize to user defined class hierarchies as well.
+We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST),
+and the resulting AST can be converted back into the original function.
+Homoiconicity is a useful property because it lets libraries implement arbitrary program transformations on these functions by manipulating the AST.
+This pearl shows that many Haskell functions are homoiconic.
+The technique does not rely on template Haskell,
+and so works on expressions available at either compile or run time.
+To motivate this homoiconicity,
+we construct a small library that automatically transforms numerically unstable code into stable code.
+
+%This pearl shows that many Haskell functions are \emph{homoiconic}.
+%That is, we can convert these functions into isomorphic abstract syntax trees (ASTs),
+%and these ASTs can be converted back into their original Haskell functions.
+%%These ASTs are ordinary Haskell data types,
+%%so they can be manipulated arbitrarily to perform program transformations.
+%The construction does not require template Haskell,
+%and so works on expressions available at either compile or run time.
+%To motivate this homoiconicity,
+%we construct a small library that automatically transforms numerically unstable code into stable code.
+
+%Floating point numbers are dangerous.
+%Due to their limited precision,
+%seemingly correct mathematical formulae can give wildly inaccurate results.
+%This pearl shows that Haskell's type system can mitigate this danger.
+%
+%The main observation is that many numerical Haskell functions are \emph{homoiconic}.
+%That is, we can easily convert them into isomorphic abstract syntax trees (ASTs).
+%The construction does not use template Haskell, and so works at both compile and run time.
+%We will use these ASTs to solve two problems with floating point numbers:
+%\begin{enumerate}
+%\item
+%We will write higher order functions that automatically stabilize (or optimize!) mathematical formulae.
+%%This lets one programmer write naive floating point code;
+%%then an expert on numerical analysis can write code that automatically improves the naive code.
+%\item
+%We will specify type class laws for floating point arithmetic that can be automatically tested.
+%Existing test frameworks do not work on floating point numbers because floats do not obey many traditional algebraic laws like associativity.
+%\end{enumerate}
+%We demonstrate these techniques using GHC's built-in numeric class hierarchy,
+%but these techniques generalize to user defined class hierarchies as well.
 \end{abstract}
@@
 %and to implement it, you'll need to implement the log-logistic function
-To implement the classic logistic regression algorithm,
-you begin by implementing the log-logistic function
+%To implement the classic logistic regression algorithm,
+%you begin by implementing the log-logistic function
+At some point, you'll need to implement the log-logistic function
 \begin{equation}
@@
 \end{equation}
-%This function is used in logistic regression and many families of neural networks.
+This function is used in logistic regression and many families of neural networks.
 As a first attempt, you might write:
@@
 Therefore, on an input of $-710$,
-|logLogisitic1| will take the |log| of 0,
+|logLogisitic| will take the |log| of 0,
 returning negative infinity when the true answer is actually very close to $-710$.
@@
 
-With some careful algebra, we can rewrite our log-logistic function in the following numerically stable form.
-\begin{lstlisting}
-logLogistic2 :: (Floating x, Ord x) => x -> x
-logLogistic2 x = m+log(1/(exp(m)+exp(-x+m)))
-    where
-        m = min 0 x
-\end{lstlisting}
-It is hardly obvious that these two functions are meant to calculate the same expression.
-The second implementation not only contains many more terms,
-it has a different type signature due to the use of the |min| function.
-
-Constructing these numerically stable functions is a notoriously difficult black art.\footnote{
-For some classic introductions, see \cite{goldberg1991,higham2002,hamming2012}.
-}
-Wouldn't it be nice to have a tool to automate this procedure?
-We'll construct a simple one in Section 3.
-
-The logistic regression algorithm operates on vector spaces;
-so you're next task is to get a proper vector space implementation.
-Since you plan to use the foreign function interface heavily,
-you decide to represent vectors as a pointer to a memory location.
-\begin{lstlisting}
-newtype Vector (dim::Nat) a = Vector (ForeignPtr a)
-\end{lstlisting}
-The |dim| type parameter encodes the dimension of the vector space.
-It makes sense to store this information at the type level because vector space operations don't change a vector's dimension.
-
-To make our |Vector| type useful,
-we must define how to perform the standard mathematical operations on it.
-You begin by making it an instance of |Num|.
-Since you're a careful programmer, you decide to write test cases to ensure your |Num| instance is correct.
-You begin by writing a QuickCheck property to ensure that your |(+)| function is properly associative.
-It works for |Int|s, but misteriously fails on |Float|s and |Double|s.
-%Unfortunately, the standard Haskell toolkit is not sufficient for the challenge.
-
-\newpage
-%Fortunately, a simple technique can be used to improve the stability of the log-logistic function.
-%First, check our input to see if it would overflow the |exp| function.
-%If no overflow would occur, then calculate our function as normal;
-%but if overflow would occur, then use an approximation that does not rely on a call to |exp|.
-%Performing this transformation gives us:
-%\begin{lstlisting}
-%logLogistic2 :: (Floating x, Ord x) => x->x
-%logLogistic2 x = if x < (-709)
-    %then x
-    %else log(1/1+exp(-x))
-%\end{lstlisting}
-%This new function is numerically stable, but we've introduced two new problems:
-
-We've made our function numerically stable, but we've introduced two new problems:
-
-\begin{problem}[obfuscation]
-The purpose of the if statement in |logLogistic2| is non-obvious.
-To understand its purpose,
-the reader must carefully consider the limits of floating point arithmetic.
-Anyone who doesn't regularly work with floating point numbers is unlikely to discover this purpose on their own.
-By transforming our function to have the correct floating point semantics,
-we've obfuscated the function's purpose and made it harder for others to understand and contribute to our library.
-\end{problem}
-
-%\begin{problem}[specialization]
-%The \lstinline{exp} function when specialized to \lstinline{Float}s overflows on values greater than (approximately) 88.
-%This means that |logLogistic2| will give incorrect answers for inputs between 88 and 709.
-%To properly handle both |Float|s and |Double|s,
-%we must write two separate functions.
-%In practice, we may want to support many other numeric types as well.
-%Half precision (16 bit) floats and fixed point arithmetic are becoming increasingly popular in machine learning applications due to their increased speed, smaller memory footprint, and lower energy consumption \cite{gupta2015}.
-%We need to manually reimplement the log-logistic function specialized to every numeric type we want our library to support.
-%\end{problem}
-
-These problems are not unique to the log-logistic function.
-All functions over the real or complex numbers need to be transformed for floating point stability,
-and all such manual transformations suffer from these problems.
-This pearl presents a method for alleviating these problems.
-Specifically, we will see how to write a function
-%\begin{lstlisting}
-%stabilize :: (Floating x, Ord x)
-    %=> (x -> x) -> (x -> x)
-%\end{lstlisting}
-\begin{lstlisting}
-stabilize :: (Floating x,Ord x) => (x->x) -> (x->x)
-\end{lstlisting}
-that automatically takes an input function like |logLogistic1| and outputs a numerically stable version of that function.
-Unfortunately, this introduces a third problem:
-
-\begin{problem}[unlawfulness]
-%How do we know what transformations are valid?
-The standard way to encode valid program transformations in Haskell is through type class laws.
-A classic example is the |Semigroup| type class
-\begin{lstlisting}
-class Semigroup a where
-    (+) :: a -> a -> a
-\end{lstlisting}
-which has the law that |(+)| should be associative.
-%That is, for every type |a| that is a valid instance of |Semigroup|,
-%|(a1+a2)+a3::a| should be equal to |a1+(a2+a3)::a|.
-That is, if the type |a| is an instance of |Semigroup|,
-and the variables |a1|, |a2|, and |a3| all have type |a|,
-then |(a1+a2)+a3| should be equal to |a1+(a2+a3)|.
-This is not true of floating point numbers, however.
-The expression |(0.1+0.2)+0.3| is famously not equal to |0.1+(0.2+0.3)| for |Double|s.
-But there is clearly a sense in which |Double|s are approximately associative,
-and performing an associativity transformation cannot change our answer ``too much.''
-%How can we formalize this notion?
-In order for our |stabilize| method to be generic and useful,
-we need a way to formalize this notion.
-\end{problem}
-
-This pearl attempts to solve all three problems.
-First, we provide a mechanical procedure for constructing FAlgebras from what we call \emph{homogeneous} type classes.
-(Both terms will be defined formally in the next section.)
-These FAlgebras make our numeric functions \emph{homoiconic};
-that is, we can easily get a syntax tree representing the function.
-Second, we show how these syntax trees let us easily manipulate our numeric functions to implement the |stabilize| function described above.
-Third, we introduce a notion of \emph{topological class laws}.
-These class laws let us formalize the intuition that floating point numbers approximately obey class laws.
-Finally, we extend our construction of FAlgebras to the \emph{heterogeneous} case.
-Heterogeneous type classes let us represent the complicated numeric type class hierarchies that arise in practice when working with numeric types.
-
-\section{Homogeneous Functions are homoiconic}
-Consider the simple numeric class hierarchy shown in Figure \ref{code:group}.
-With this class hierarchy, we can write a function that measures the degree to which the |(+)| operation obeys the associative rule.
-\begin{lstlisting}
-associator :: Group a => a -> a -> a -> a
-associator a1 a2 a3
-    = ((a1+a2)+a3) - (a1+(a2+a3))
-\end{lstlisting}
-We call the |associator| function \emph{homogeneous} because:
+But like all good programmers, you're lazy.
+Your machine learning library will contain thousands of numeric functions,
+all of which will need to be carefully written to avoid numeric instabilities.
+What a maintenance nightmare!
+Soon you start daydreaming about a function
+\begin{lstlisting}
+stabilize :: Floating a => (a -> a) -> (a -> a)
+\end{lstlisting}
+that takes a naively written floating point function as input and outputs a numerically stable version.
+Armed with such a function, you could write a stable implementation of the log-logistic function as
+\begin{lstlisting}
+logLogistic2 :: Floating x => x -> x
+logLogistic2 x = stabilize $ log(1/(1+exp(-x)))
+\end{lstlisting}
+%stopzone
+
+In this pearl, we see how to write functions like |stabilize|.
+The key observation is that certain Haskell functions can be converted into an isomorphic abstract syntax tree (AST).
+The |stabilize| function uses this fact to convert its input function into an AST;
+manipulates that AST to remove numerical stabilities;
+then converts the AST back into a callable function.
+
+In Section \ref{sec:homogeneous}, we introduce the idea of a \emph{homogeneous} function and show that these functions are homoiconic.
+In Section \ref{sec:stabilize}, we use the construction above to implement several versions of the |stabilize| function.
+In Section \ref{sec:heterogeneous}, we introduce the idea of \emph{heterogeneous} functions and show that they are also homoiconic.
+Finally, we conclude by introducing a Haskell package called {\ttfamily homoiconic} that makes it easy for you to .
+
+\section{Homogeneous functions are homoiconic}
+\label{sec:homogeneous}
+We call a function \emph{homogeneous} if:
 \begin{enumerate}
 \item
-The function's return type and all of its parameters are the same type variable.
+%The function's return type and all of its parameters are the same type variable.
+The function is polymorphic in its return type.
+\item
+The function's parameters are either the same as the return type or they are concrete.
 \item
 The constraints contain exactly one type class, and
-that type class has exactly one parameter.
+that type class has exactly one parameter,
+which is the function's return type.
 \end{enumerate}
-Notice that each of the class methods for |Semigroup|, |Monoid|, and |Group| are also homogeneous.
-Therefore, we will also say these classes are homogeneous.
-
-In this section, we will see how to convert any homogeneous function into an isomorphic abstract syntax tree (AST).
-This ability is called \emph{homoiconicity},
-and is usually associated with Lisps.
-We will then see several methods for evaluating these ASTs.
-In subsequent sections, we will manipulate these ASTs to make floating point operations more convenient.
-
-\begin{figure}
-\begin{lstlisting}
-class Semigroup a where
-    (+) :: a -> a -> a
-
-class Semigroup a => Monoid a where
-    zero :: a
-
-class Monoid a => Group a where
-    (-) :: a -> a -> a
-    negate :: a -> a
-\end{lstlisting}
-\caption{A simple numeric hierarchy that obeys the restrictions for homogeneous FAlgebras.}
-\label{code:group}
-\end{figure}
-
-%\begin{figure}
-%\begin{lstlisting}
-%class Num a where
-    %(+), (-), (*)       :: a -> a -> a
-    %negate              :: a -> a
-    %abs, signum         :: a -> a
-    %fromInteger         :: Integer -> a
-%
-%class Num a => Fractional a where
-    %(/)                 :: a -> a -> a
-    %recip               :: a -> a
-    %fromRational        :: Rational -> a
-%
-%class Fractional a => Floating a where
-    %pi                  :: a
-    %exp, log, sqrt      :: a -> a
-    %(**), logBase       :: a -> a -> a
-    %sin, cos, tan       :: a -> a
-    %asin, acos, atan    :: a -> a
-    %sinh, cosh, tanh    :: a -> a
-    %asinh, acosh, atanh :: a -> a
-%
-%class Eq a where
-    %(==), (/=)          :: a -> a -> Bool
-%
-%class Eq a => Ord a where
-    %compare         :: a -> a -> Ordering
-    %(<),(<=),(>),(>=)   :: a -> a -> Bool
-    %max, min            :: a -> a -> a
-%\end{lstlisting}
-%\caption{
-    %A portion of the numeric and comparison hierarchies defined in GHC's Prelude.
-    %Many class methods satisfy the requirements for homogeneous FAlgebras, except:
-    %%|fromInteger| and |fromRational| have non-variable parameters;
-    %%and |(==)|, |(/=)|, |(<)|, |(<=)|, |(>)|, and |(>=)| have non-variable return types.
-%}
-%\label{code:ghc}
-%\end{figure}
-
-\subsection{A type class for FAlgebras}
+The |logLogistic1| function is homogeneous.
+Figure \ref{code:ghc} shows many more examples from GHC's numeric and ordering hierarchies.
+
+In this section, we will show that every type class has an associated monad representing the abstract syntax tree of the class's homogeneous functions.
 Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
 See for example \cite{swierstra2008}.
-The heart of our construction is the |FAlgebra| type class, defined as
-\begin{lstlisting}
-class
-    ( Functor (Sig alg)
-    , Foldable (Sig alg)
-    ) => FAlgebra (alg::Type->Constraint)
-        where
+
+\begin{figure}
+\begin{lstlisting}
+class Num a where
+    (+), (-), (*)       :: a -> a -> a
+    negate              :: a -> a
+    abs, signum         :: a -> a
+    fromInteger         :: Integer -> a
+
+class Num a => Fractional a where
+    (/)                 :: a -> a -> a
+    recip               :: a -> a
+    fromRational        :: Rational -> a
+
+class Fractional a => Floating a where
+    pi                  :: a
+    exp, log, sqrt      :: a -> a
+    (**), logBase       :: a -> a -> a
+    sin, cos, tan       :: a -> a
+    asin, acos, atan    :: a -> a
+    sinh, cosh, tanh    :: a -> a
+    asinh, acosh, atanh :: a -> a
+
+class Eq a where
+    (==), (/=)          :: a -> a -> Bool
+
+class Eq a => Ord a where
+    compare             :: a -> a -> Ordering
+    (<),(<=),(>),(>=)   :: a -> a -> Bool
+    max, min            :: a -> a -> a
+\end{lstlisting}
+\caption{
+    A portion of the numeric and comparison hierarchies defined in GHC's Prelude.
+    %Functions highlighted in bold are homogeneous,
+    %and the remainder are not.
+}
+\label{code:ghc}
+\end{figure}
+
+\subsection{A type class for FAlgebras}
+FAlgebras are a technique used in the mathematical fields of universal algebra and category theory for defining algebraic structures.
+We represent an FAlgebra with the following type class.
+\begin{lstlisting}
+class Functor (Sig alg) => FAlgebra alg where
     data Sig alg a
@@
 The |FAlgebra| class is unusual in that its parameter has kind |Type->Constraint|.
-This means that instances of |FAlgebra| will be other type classes (instead of types or type constructors).
-The || in |FAlgebra| stands for homogeneous.
-Most of the names used in this section will be prefixed with || to differentiate them from related names used in the later heterogeneous constructions.
-
-The easiest way to explain the |FAlgebra| class is by example.
-We'll start with the |Semigroup| type class defined above.
-\begin{lstlisting}
-instance FAlgebra Semigroup where
-    data Sig Semigroup a = Sig_plus a a
-    runSig (Sig_plus a1 a2) = a1+a2
-\end{lstlisting}
-The |Sig| associated data family is meant to encode what mathematicians call the \emph{signature} of the FAlgebra.
-That is, for every class method, |Sig| should have a constructor with the same number of parameters.
-Since |Semigroup| has a single function that takes two parameters,
-the |Sig Semigroup| data instance has a single constructor that takes two parameters.
-The |runSig| function evaluates the signature.
-For each constructor in |Sig|, |runSig| should apply the corresponding function to the constructor's parameters.
-
-To complete the |FAlgebra Semigroup| instance, we must also define the following superclass instances.
-\begin{lstlisting}
-instance Functor (Sig Semigroup) where
-    fmap f (Sig_plus a1 a2)
-        = Sig_plus (f a1) (f a2)
-
-instance Foldable (Sig Semigroup) where
-    foldr f b (Sig_plus a1 a2)
-        = f a2 (f a1 b)
+This means that instances of |FAlgebra| will be other type classes
+(instead of types or type constructors).
+
+It is easiest to understand instances of |FAlgebra| by walking through an example.
+Below is the instance for |Fractional|.
+\begin{lstlisting}
+instance FAlgebra Fractional where
+    data Sig Fractional a
+        = Sig_div a a
+        | Sig_recip a
+        | Sig_fromRational Rational
+        | Sig_Fractional_Num (Sig Num a)
+    runSig (Sig_div a1 a2)        = a1/a2
+    runSig (Sig_recip a)          = recip a
+    runSig (Sig_fromRational r)   = fromRational r
+    runSig (Sig_Fractional_Num s) = runSig s
+\end{lstlisting}
+
+The data family |Sig| encodes what mathematicians call the \emph{signature} of the FAlgebra.
+The signature defines all the operations that can be performed on an FAlgebra.
+Each |Sig| data instance can have many constructors,
+which can be divided into two categories.
+First, there should be one constructor for each homogeneous class function.
+This constructor should have the same parameters as the function.
+For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three member functions |(/)|, |recip|, and |fromRational|.
+Second, there should be a constructor for each superclass.
+This constructor should store the |Sig| of the corresponding superclass.
+For the |Fractional| example, we have the |Sig_Fractional_Num| constructor corresponding to the |Num| superclass.
+
+The |runSig| class method gives us a way to evaluate any |Sig|.
+For each function constructor, |runSig| should call the corresponding function.
+For each superclass constructor, |runSig| recursively calls |runSig| on the superclass's |Sig|.
+
+The F in FAlgebra comes from the fact that every signature is actually a functor.
+In Haskell, we encode this by enforcing that |Sig alg| must be an instance of the |Functor| class in order to create a valid |FAlgebra| instance for |alg|.
+As usual, there is only a single valid |Functor| instance.
+For |Fractional|, it is:
+
+\begin{lstlisting}
+instance Functor (Sig Fractional) where
+    fmap f (Sig_div a1 a2) = Sig_div (f a1) (f a2)
+    fmap f (Sig_recip a) = Sig_recip (f a)
+    fmap f (Sig_fromRational r) = Sig_fromRational r
+    fmap f (Sig_Fractional_Num s) = fmap f s
 \end{lstlisting}
@@
 
-Instances get slightly more complicated when the parameter class has superclasses.
-In this case, the |Sig| data instance must also have a constructor for each superclass that stores the superclass's signature.
-This can be seen in the |Monoid| and |Group| instances below.
-%The |FAlgebra| instances of |Monoid| and |Group| demonstrate this.
-\begin{lstlisting}
-instance FAlgebra Monoid where
-    data Sig Monoid a
-        = Sig_zero
-        | Sig_Semigroup_Monoid
-            (Sig Semigroup a)
-    runSig Sig_zero = zero
-    runSig (Sig_Semigroup_Monoid s)
-        = runSig s
-
-instance FAlgebra Group where
-    data Sig Group a
-        = Sig_negate a
-        | Sig_minus a a
-        | Sig_Monoid
-            (Sig_Monoid_Group a)
-    runSig (Sig_negate a) = a
-    runSig (Sig_minus a1 a2) = a1-a2
-    runSig (Sig_Monoid_Group s)
-        = runSig s
-\end{lstlisting}
-Notice that the data instance |Sig Group| does not need a constructor for the |Semigroup| signature because this is included in the |Monoid| signature.
-The corresponding |Functor| and |Foldable| instances have been omitted for brevity,
-but can be found in the paper's git repo.
-
-%The procedure for generating |FAlgebra| instances is purely mechanical.
-%The ?? library automates the process with the |mkHomFAlgebra| template Haskell function.
-
 \subsection{Constructing the AST}
 
-We can convert the signature of an FAlgebra into an AST as follows:
-\begin{lstlisting}
-type AST alg a = Free (Sig alg) a
-
+For every FAlgebra, there is an associated AST with the corresponding algebraic structure.
+This AST is sometimes called an \emph{initial algebra},
+and has a standard construction via the \emph{free monad}.
+
+In Haskell, the free monad is defined as
+\begin{lstlisting}
 data Free f a
@@
     | Free (f (Free f a))
+\end{lstlisting}
+which lets us define our AST as
+\begin{lstlisting}
+type AST alg a = Free (Sig alg) a
 \end{lstlisting}
@@
 For example, consider the expression |(1+2)+3|.
-We can create a corresponding AST using |Semigroup|'s |FAlgebra| instance.
+We can create a corresponding AST using |Num|'s |FAlgebra| instance.
 \begin{lstlisting}
@@
 \end{lstlisting}
-We can also create a corresponding AST for any class that has |Semigroup| as a superclass.
-Below is the |Group| AST.
+We can also create a corresponding AST for any class that has |Num| as a superclass.
+Below is the same expression encoded using |Floating|'s |FAlgebra| instance.
 \begin{lstlisting}
 Free
-  (Sig_Monoid_Group
-    (Sig_Semigroup_Monoid
+  (Sig_Floating_Fractional
+    (Sig_Fractional_Num
       (Sig_plus
         (Free
-          (Sig_Monoid_Group
-            (Sig_Semigroup_Monoid
+          (Sig_Floating_Fractional
+            (Sig_Fractional_Num
               (Sig_plus
@@
 \end{lstlisting}
-Notice that the |Sig_plus| constructor must be embedded into the |Group|'s |Sig| with calls to |Sig_Monoid_Group| and |Sig_Semigroup_Monoid|.
+Notice that the |Sig_plus| constructor must be embedded into |Floating|'s |Sig| with calls to |Sig_Floating_Fractional| and |Sig_Fractional_Num|.
 
 Constructing these ASTs by hand is a tedious process.
-There is a considerable amount of boilerplate in the |Group| AST above as a result of the need to embed the |Sig_plus| constructor.
+There is a considerable amount of boilerplate in the |Floating| AST above as a result of the need to embed the |Sig_plus| constructor.
 We can avoid this boilerplate with the following type class.
@@
         where
-    embedHomSig
-        :: Sig alg1 a -> Sig alg2 a
-    unsafeExtractHomSig
-        :: Sig alg2 a -> Sig alg1 a
+    embedSig         :: Sig alg1 a -> Sig alg2 a
+    unsafeExtractSig :: Sig alg2 a -> Sig alg1 a
 \end{lstlisting}
@@
 all such relationships should have a corresponding instance.
-For example, |Semigroup| is a superclass of |Group|,
-so we should have a |View Semigroup Group| instance.
-\begin{lstlisting}
-instance View Semigroup Group
-    embedHomSig s
-        = Sig_Monoid_Group (embedHomSig s)
-    unsafeExtractHomSig (Sig_Monoid_Group s)
-        = unsafeExtractHomSig s
+For example, |Num| is a superclass of |Floating|,
+so we should have a |View Num Floating| instance.
+\begin{lstlisting}
+instance View Num Floating
+    embedSig s
+        = Sig_Fractional_Floating (embedSig s)
+    unsafeExtractSig (Sig_Fractional_Floating s)
+        = unsafeExtractSig s
 \end{lstlisting}
@@
 Free
-  (embedHomSig
+  (embedSig
     (Sig_plus
       (Free
-        (embedHomSig
+        (embedSig
           (Sig_plus
@@
 
-%Every FAlgebra can be embedded into itself.
-%\begin{lstlisting}
-%instance View alg alg where
-    %embedSig = id
-    %unsafeExtractSig = id
-%\end{lstlisting}
-
-%\begin{lstlisting}
-%instance View Semigroup Monoid where
-    %embedSig = Sig_Monoid_Semigroup
-    %unsafeExtractSig (Sig_Monoid_Semigroup s) = s
-
-%instance View Monoid Group where
-    %embedSig s = Sig_Group_Monoid (embedSig s)
-    %unsafeExtractSig (Sig_Group_Monoid s) = unsafeExtractSig s
-%\end{lstlisting}
-
-%\begin{lstlisting}
-%instance View Semigroup Group where
-    %embedSig s = Sig_Group_Monoid (embedSig s)
-    %unsafeExtractSig (Sig_Group_Monoid s) = unsafeExtractSig s
-%\end{lstlisting}
-
 Using our uniform representation of an AST,
-we can make the AST itself an instance of any homogeneous type class.
+we can make the AST itself an instance of any |FAlgebra| type class.
 Again, we'll approach the procedure by example.
-We can write a |Semigroup| instance for |AST Semigroup a| as:
-\begin{lstlisting}
-instance
-    ( View Semigroup alg
-    ) => Semigroup (Free (Sig alg) a)
+We can write a |Num| instance for |AST Num a| as:
+\begin{lstlisting}
+instance View Num alg => Num (Free (Sig alg) a)
         where
-    (+) e1 e2 = Free $ embedHomSig $
-            Sig_plus e1 e2
+    (+) e1 e2 = Free (embedSig (Sig_plus e1 e2))
 \end{lstlisting}
 For each function in the type class,
-we simply embed the function's constructor into the AST by prepending the expression |Free $ embedHomSig $|.
-The |Monoid| and |Group| instances follow a similar pattern.
-\begin{lstlisting}
-instance
-    ( View Semigroup alg
-    , View Monoid alg
-    ) => Monoid (Free (Sig alg) a)
-        where
-    zero = Free $ embedHomSig $
-        Sig_zero
-
-instance
-    ( View Semigroup alg
-    , View Monoid alg
-    , View Group alg
-    ) => Group (Free (Sig alg) a)
-        where
-    negate e  = Free $ embedHomSig $
-        Sig_negate e
-    (-) e1 e2 = Free $ embedHomSig $
-        Sig_minus e1 e2
-\end{lstlisting}
+we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.
+The |Fractional| and |Floating| instances follow a similar pattern.
+
 We can easily create ASTs using these instances.
-The |Semigroup| AST can be created with the expression
-\begin{lstlisting}
-(Pure 1+Pure 2)+Pure 3
-    :: AST Semigroup a
-\end{lstlisting}
-And the |Group| AST can be created by just changing the type signature.
-\begin{lstlisting}
-(Pure 1+Pure 2)+Pure 3
-    :: AST Group a
+The |Num| AST can be created with the expression
+\begin{lstlisting}
+(1+2)+3 :: AST Num a
+\end{lstlisting}
+And the |Floating| AST can be created by just changing the type signature.
+\begin{lstlisting}
+(1+2)+3 :: AST Floating a
 \end{lstlisting}
@@
 
-The are no tricks in the |Show| instance for |Semigroup|'s |Sig|.
-\begin{lstlisting}
-instance Show a =>
-    Show (Sig Semigroup a)
-        where
-    show (Sig_plus a1 a2)
-        = show a1++"+"++show a2
-\end{lstlisting}
-For each operation constructor, we simply show the parameters and the operation connecting them.
-For type classes with superclasses, we also need to recursively show the superclasses as can be seen in the |Monoid| and |Group| instances below.
-\begin{lstlisting}
-instance Show a =>
-    Show (Sig Monoid a)
-        where
-    show (Sig_Semigroup_Monoid s)
-        = show s
-    show (Sig_zero) = "zero"
-
-instance Show a =>
-    Show (Sig Group a)
-        where
-    show (Sig_Monoid_Group s)
-        = show s
-    show (Sig_negate a)
-        = "negate "++show a
-    show (Sig_minus a1 a2)
-        = show a1++"-"++show a2
-\end{lstlisting}
-The |Free| type's |Show| instance is similarly straightforward..
+The are no tricks in the |Show| instance for the |Sig| instance.
+The instance for |Fractional| is shown below.
+\begin{lstlisting}
+instance Show a => Show (Sig Fractional a) where
+    show (Sig_div a1 a2) = show a1++"/"++show a2
+    show (Sig_recip a) = "recip "++show a
+    show (Sig_fromRational r) = "fromRational "++show r
+    show (Sig_Fractional_Num s) = show s
+\end{lstlisting}
+For each function constructor, we simply show the corresponding function and its parameters.
+If the function is an operator, we'll display it infix for convenience.
+For each superclass constructor, we simply recursively call show on the superclass's |Sig|.
+
+The |Free| type's |Show| instance is similarly straightforward.
 \begin{lstlisting}
@@
 
+Now, when we type an expression into ghci,
+\begin{lstlisting}
+ghci> (1+2)+3
+6
+ghci> (1+2)+3 :: AST Floating Int
+(1+2)+3
+\end{lstlisting}
+
 \subsection{Evaluating the syntax tree}
@@
 generating the value they represent.
-This data type is called the free monad because as long as the parameter |f| is a |Functor|,
+To do this, we will take advantage of the free monad's structure.
+The free monad is so called because as long as the parameter |f| is a |Functor|,
 then |Free| has a valid |Monad| instance.
@@
     fmap g (Pure a) = Pure (g a)
-    fmap g (Free f)
-        = Free (fmap (fmap g) f)
+    fmap g (Free f) = Free (fmap (fmap g) f)
 \end{lstlisting}
@@
 \begin{lstlisting}
-runAST :: (FAlgebra alg, alg a)
-    => AST alg a -> a
+runAST :: (FAlgebra alg, alg a) => AST alg a -> a
 runAST (Pure a) = a
@@
 \section{Implementing the stabilize function}
+\label{sec:stabilize}
 
@@
 \begin{lstlisting}
-stabAST :: Eq a => AST FloatingOrd a -> AST FloatingOrd a
+stabAST :: Eq a =>
+    AST FloatingOrd a -> AST FloatingOrd a
 stabAST
@@
 
-\section{Making floats lawful}
-
-Many type classes in Haskell come with an associated set of class laws.
-Usually, these laws are stated informally via haddock documentation.
-For example, the haddock documentation for the |Functor| class
-\begin{lstlisting}
-class Functor f where
-    fmap :: (a -> b) -> f a -> f b
-\end{lstlisting}
-states that all instances should satisfy the following two laws.
-\begin{lstlisting}
-fmap id  ==  id
-fmap (f . g)  ==  fmap f . fmap g
-\end{lstlisting}
-The type classes in the numeric hierarchy, however, do not have any laws.
-%For example, the haddock documentation for the |Monoid| class
-
-In this section, we'll show how to encode the laws for any |FAlgebra| directly in Haskell.
-This makes verifying the laws easy.
-Then we'll show how to make these laws topological.
-
-\subsection{Varieties and exact laws}
-
-A \emph{variety} is an algebraic structure that has a set of associated laws.
-A law can be defined by a pair of ASTs.
-These ASTs are equivalent in the sense that they can be substituted for each other without changing the program's semantics.
-For convenience, we assign each law a name and put it in the following data type.
-We will represent a law with the following data type.
-\begin{lstlisting}
-data Law alg = Law
-    { lawName :: String
-    , lhs     :: AST alg Var
-    , rhs     :: AST alg Var
+\section{Heterogeneous functions are homoiconic}
+
+%\begin{lstlisting}
+%data Vec3 a = Vec3 a a a
+    %deriving (Eq,Show)
+%\end{lstlisting}
+
+One of the main limitations of GHC's numeric hierarchy is that it was designed for scalar numbers.
+A logistic regression implementation needs vectors.
+There's been many proposals to extend GHC's built-in hierarchy to handle vectors.
+A very simple such extension is shown in Figure \ref{code:vector}.
+This extension uses type families.
+
+\begin{figure}
+\begin{lstlisting}
+type family Scalar a
+
+class (Num a, Fractional (Scalar a)) => Vector a where
+    (.*) :: Scalar a -> a -> a
+
+class Vector a => Hilbert a where
+    dotproduct :: a -> a -> Scalar a
+\end{lstlisting}
+\caption{
+    A simple extension to GHC's numeric hierarchy for vector arithmetic.
 }
-\end{lstlisting}
-%The records |lhs| and |rhs| are meant to indicate the left and right hand sides of the equality the law is encoding.
-%Specifically, in any expression, we should be able to substitute the these two expressions for each other without changing the expression's semantics.
-Now we can encode a variety in Haskell as a list of associated laws.
-\begin{lstlisting}
-class FAlgebra alg => Variety alg where
-    laws :: [Law alg]
-\end{lstlisting}
-To encode the laws of a particular type class,
-we create an appropriate |Variety| instance.
-There are many possible laws we might want a |Num| instance to obey.
-We will use small, deliberately inadequate set of laws to illustrate the point.
-The instance below encodes that the |(+)| function should be both associative and commutative.
-\begin{lstlisting}
-instance Variety Num where
-    laws =
-        [ Law
-            { lawName = "associative"
-            , lhs = (var1+var2)+var3
-            , rhs = var1+(var2+var3)
-            }
-        , Law
-            { lawName = "commutative"
-            , lhs = var1*var2
-            , rhs = var2*var1
-            }
-        ]
-\end{lstlisting}
-
-\subsubsection{Listing the laws}
-When working with a large numerical hierarchy,
-it is easy to forget the exact laws that a class must obey.
-Wouldn't it be nice if there were an easy way to ask the repl what the class laws are?
-
-Since we've encoded our class laws directly in Haskell, we can.
-First, we define a function that pretty prints a |Law|.
-\begin{lstlisting}
-printLaw :: Show (AST alg Var)
-    => Law alg -> IO ()
-printLaw law = do
-    putStrLn ( "  "++lawName law++":"      )
-    putStrLn ( "    lhs: "++show (lhs law) )
-    putStrLn ( "    rhs: "++show (rhs law) )
-\end{lstlisting}
-Then we define a function that calls |printLaw| on each class law.
-\begin{lstlisting}
-printLaws :: forall alg.
-    ( Variety alg
-    , Show (AST alg Var)
-    ) => Proxy alg -> IO ()
-printLaws palg = do
-    forM_ (laws::[Law alg]) printLaw
-\end{lstlisting}
-Now we can ask at the REPL for our class laws as follows
-\begin{lstlisting}
-ghci> printLaws (Proxy::Proxy Num)
-\end{lstlisting}
-which gives the following output
-\begin{lstlisting}
-  associative:
-    lhs: ((var1+var2)+var3)
-    rhs: (var1+(var2+var3))
-  commutative:
-    lhs: (var1*var2)
-    rhs: (var2*var1)
-\end{lstlisting}
-
-\subsubsection{Testing the laws (Part I)}
-We can also use the first class nature of our laws to automatically test that instances obey them.
-
-First, we need to be able to convert an AST of |Var|s into an AST of arbitrary types by substituting.
-\begin{lstlisting}
-subVars :: FAlgebra alg => AST alg Var -> [(Var,a)] -> AST alg a
-subVars expr varmap = fmap go expr
+\label{code:vector}
+\end{figure}
+
+\begin{lstlisting}
+logLoss :: Vector a => a -> a -> Scalar a
+logLoss a1 a2 = logLogistic1 (dotproduct a1 a2)
+\end{lstlisting}
+
+The |logLoss|, |(.*)|, and |dotProduct| functions are not homogeneous because of the type families.
+Heterogeneous functions are a relaxation of homogeneous functions.
+We call a function \emph{heterogeneous} if:
+\begin{enumerate}
+\item
+The function is polymorphic in its return type.
+\item
+The function's parameters are either the same as the return type, they are concrete.
+\item
+The constraints contain exactly one type class, and
+that type class has exactly one parameter,
+which is the function's return type.
+\end{enumerate}
+
+\subsection{Dealing with type families}
+
+\begin{lstlisting}
+type AT = Type
+\end{lstlisting}
+
+\begin{lstlisting}
+type family App (t::k) (a::Type) ::  Type
+type instance App '[]       a = a
+type instance App (x ': xs) a = App x (App xs a)
+\end{lstlisting}
+
+\subsection{Heterogeneous FAlgebras}
+
+\begin{lstlisting}
+class FunctorTag (f::[AT]->Type->Type) where
+    fmapTag :: HaskTag a b -> f t a -> f t b
+\end{lstlisting}
+
+\begin{lstlisting}
+class HetFunctor (Sig alg) => FAlgebra (alg::Type->Constraint) where
+    data Sig alg (t::[AT]) a
+
+    runSigTag :: alg a => proxy a -> Sig alg (s ':  t) (App t a) -> App (s ':  t) a
+    runSig    :: alg a => proxy a -> Sig alg        t  (App t a) -> App        t  a
+\end{lstlisting}
+
+\subsection{Constructing the AST}
+
+\begin{lstlisting}
+type AST alg t a = Free (Sig alg) t a
+
+data Free (f::[AT]->Type->Type) (t::[AT]) (a::Type) where
+    FreeTag  :: f (s ': t) (Free f t a)  -> Free f (s ': t) a
+    Free     :: f       t  (Free f t a)  -> Free f       t  a
+    Pure     :: App t a -> Free f t a
+\end{lstlisting}
+
+\begin{lstlisting}
+class (FAlgebra alg1, FAlgebra alg2) => View alg1 t1 alg2 t2 where
+    embedSig          :: Sig alg1       t1  a -> Sig alg2       t2  a
+    unsafeExtractSig  :: Sig alg2       t2  a -> Sig alg1       t1  a
+\end{lstlisting}
+
+\begin{lstlisting}
+embedSigTag :: View alg1 (t ': t1) alg2 (t ': t2) => Sig alg1 (t ': t1) a -> Sig alg2 (t ': t2) a
+embedSigTag = embedSig
+\end{lstlisting}
+
+\subsection{Showing the AST}
+
+\begin{lstlisting}
+instance
+    ( Show      (App t a)
+    , Show      (f t (Free f t a))
+    , ShowUntag (f t (Free f t a))
+    ) => Show (Free f t a)
         where
-        go var = case lookup var varmap of
-            Just a -> a
-\end{lstlisting}
-
-Armed with this function, we can convert any law into a quickcheck property.
-\begin{lstlisting}
-law2property :: forall (a :: Type) alg.
+    show (FreeTag     f) = "("++show f++")"
+    show (Free        f) = "("++show f++")"
+    show (Pure        a) = show a
+
+type family ShowUntag (f::Type) :: Constraint where
+    ShowUntag (f (s ':  t) (Free f (s ':  t) a))  = Show (f (s ':  t) (Free f          t  a))
+    ShowUntag a = ()
+\end{lstlisting}
+
+\subsection{Evaluating the AST}
+
+\begin{lstlisting}
+eval :: forall alg t a.
     ( FAlgebra alg
     , alg a
-    , Eq a
-    , Arbitrary a
-    ) => Proxy a -> Law alg -> Gen Bool
-law2property p law = do
-    as <- infiniteListOf (arbitrary::Gen a)
-    let varmap = zip (toList (lhs law) ++ toList (rhs law)) as
-    return $ (runAST $ subVars (lhs law) varmap)
-          == (runAST $ subVars (rhs law) varmap)
-\end{lstlisting}
-%stopzone
-And we can test all the laws for a given class instance at once.
-\begin{lstlisting}
-class2quickcheck :: forall a alg.
-    ( Variety alg
-    , alg a
-    , Eq a
-    , Arbitrary a
-    ) => Proxy a -> Proxy alg -> IO ()
-class2quickcheck pa _ = forM_ (laws::[Law alg]) $ \law -> do
-    putStr $ lawName law++": "
-    quickCheck $ law2property pa law
-\end{lstlisting}
-%stopzone
-Now in GHCi:
-\begin{lstlisting}
-ghci> class2quickcheck (Proxy::Proxy Int) (Proxy::Proxy Num)
-associative: +++ OK, passed 100 tests.
-commutative: +++ OK, passed 100 tests.
-\end{lstlisting}
-But as mentioned before, |Float|s don't strictly obey the associative property.
-\begin{lstlisting}
-ghci> > class2quickcheck (Proxy::Proxy Float) (Proxy::Proxy Num)
-associative: *** Failed! Falsifiable (after 2 tests):
-commutative: +++ OK, passed 100 tests
-\end{lstlisting}
-We'll need something a bit more subtle for them.
-
-\subsection{The floating point topology}
-
-Checking if two floating point numbers are equal is almost always a bad idea.
-Better practice is to check if the two numbers are ``close enough'' for your application.
-We can formalize this notion via topology.
-
-In mathematics, a \emph{topology} is a set $X$ equipped with a family of subsets of $X$ called $\tau$ such that:
-%In mathematics, a \emph{topology} is a set $X$ and a set of neighborhoods $\tau \subseteq 2^X$ such that:
-\begin{enumerate}
-\item
-$\tau$ contains the empty set and the set $X$.
-\item
-$\tau$ is closed under an arbitrary number of unions.
-\item
-$\tau$ is closed under finitely many intersections.
-\end{enumerate}
-We call an element of $\tau$ a neighborhood.
-
-To encode the notion of a topology in Haskell,
-we first must encode the idea of a set.
-
-\begin{lstlisting}
-class Poset a where
-    inf :: a -> a -> a
-\end{lstlisting}
-
-\begin{lstlisting}
-class Poset (Neighborhood a) => Topology a
-        where
-    type Neighborhood a
-    (==) :: a -> a -> (Neighborhood a -> Bool)
-\end{lstlisting}
-
-%It is well known that equality testing is almost always a bad idea for floating point numbers.
-As mentioned before, floating point numbers are not strictly associative,
-but they have a notion of being approximately associative.
-In particular, when we call the |associator| function defined above on any three floating point values,
-we can guarantee that the result is small relative to the input values.
-
-\subsection{Topological test cases}
+    ) => Free (Sig alg) t a -> App t a
+eval (Pure    a) = a
+eval (Free    s) = runSig    (Proxy::Proxy a) $ mape eval s
+eval (FreeTag s) = runSigTag (Proxy::Proxy a) $ mape eval s
+\end{lstlisting}
+
+\subsection{Heterogeneous homoiconicity at last!}
 

### 2016-06-08 src/Heterogeneous/Example.hs

@@ -4,7 +4,7 @@
 {-# LANGUAGE TypeInType #-}
 {-# LANGUAGE CPP #-}
 
-module FAlgebraExample
+module Heterogeneous.Example
     where
 
 import Prelude hiding (Monoid (..),(-),(+),negate,(==),minBound)
@@ -23,7 +23,7 @@ import qualified Test.Tasty.SmallCheck as SC
 import qualified Test.Tasty.QuickCheck as QC
 import Test.QuickCheck hiding (Testable)
 
-import FAlgebra
+import Heterogeneous.FAlgebra
 -- import FAlgebra98
 -- import Topology
 -- import Lattice
@@ -37,7 +37,7 @@ class Poset a where
 
 mkFAlgebra ''Poset
 
-instance cxt '[] => FunctorTag cxt (Sig Poset) where
+instance FunctorTag (Sig Poset) where
     fmapTag f (Sig_inf e1 e2) = Sig_inf
         (apHaskTag (Proxy::Proxy '[]) f e1)
         (apHaskTag (Proxy::Proxy '[]) f e2)
@@ -58,7 +58,6 @@ instance Variety Poset where
             }
         ]
 
-{-
 #define mkPoset(x) \
 instance Poset x where inf = P.min
 
@@ -573,4 +572,3 @@ z :: Free (Sig Space) '[TFloo] (Int,Int)
 z = Pure 2
 
 type instance TypeConstraints t a = ()
--}

### 2016-06-08 src/Heterogeneous/ExamplePrelude.hs

@@ -1,0 +1,142 @@
+module Heterogeneous.ExamplePrelude
+    where
+
+import Prelude hiding (Ord(..))
+import Heterogeneous.FAlgebra
+
+import Data.Proxy
+
+--------------------------------------------------------------------------------
+
+type instance TypeConstraints t a = ()
+
+mkFAlgebra ''Num
+mkFAlgebra ''Fractional
+mkFAlgebra ''Floating
+-- mkFAlgebra ''Eq
+-- instance FAlgebra Eq
+-- instance Functor (Sig Eq)
+-- instance Foldable (Sig Eq)
+-- instance Show (Sig Eq a)
+-- instance Eq (Sig Eq a)
+-- mkFAlgebra ''Ord
+
+class Ord a where
+    min :: a -> a -> a
+    max :: a -> a -> a
+mkFAlgebra ''Ord
+
+class (Floating a, Ord a) => FloatingOrd a
+instance {-#OVERLAPPABLE#-} (Floating a, Ord a) => FloatingOrd a
+mkFAlgebra ''FloatingOrd
+
+-- pattern Expr_plus ::
+--     ( ViewPoset alg t
+--     , TypeConstraints t a
+--     ) => Free (Sig alg) t a
+--       -> Free (Sig alg) t a
+--       -> Free (Sig alg) t a
+
+pattern Expr_log :: forall t alg a.
+    ( View Floating '[] alg t
+    ) => Free (Sig alg) t a
+      -> Free (Sig alg) t a
+pattern Expr_log e <- Free ((unsafeExtractSig :: Sig alg      t   (Free (Sig alg) t a)
+                                              -> Sig Floating '[] (Free (Sig alg) t a)
+                            ) -> Sig_log e) where
+-- pattern Expr_log e <- Free (unsafeExtractSig -> Sig_log e) --where
+    Expr_log e = Free $ embedSig $ Sig_log e
+
+--------------------------------------------------------------------------------
+
+type family Scalar a
+mkAT ''Scalar
+
+type instance Scalar Float = Float
+type instance Scalar (Free (Sig alg) t a) = Free (Sig alg) (TScalar ': t) a
+
+----------------------------------------
+
+class (Num a, Fractional (Scalar a)) => Vector a where
+    (.*) :: Scalar a -> a -> a
+
[…]

### 2016-06-08 src/Heterogeneous/FAlgebra.hs

@@ -7,7 +7,7 @@
 
 {-# OPTIONS_GHC -Wno-simplifiable-class-constraints #-}
 
-module FAlgebra
+module Heterogeneous.FAlgebra
     where
 
 import Prelude
@@ -135,6 +135,8 @@ class FunctorTag (f::[AT]->Type->Type) where
 
 ----------------------------------------
 
+type AST (alg::Type->Constraint) t a = Free (Sig alg) t a
+
 data Free (f::[AT]->Type->Type) (t::[AT]) (a::Type) where
     FreeTag  :: TypeConstraints t a => f (s ': t) (Free f t a)  -> Free f (s ': t) a
     Free     :: TypeConstraints t a => f       t  (Free f t a)  -> Free f       t  a
@@ -212,7 +214,8 @@ instance
     ( View alg1 (s++t) alg2 (s++t)
     ) => ValidView alg1 alg2 t s
 
-embedExpr :: forall alg1 alg2 cxt (s::[AT]) t a.
+{-
+embedAST :: forall alg1 alg2 cxt (s::[AT]) t a.
     ( FAlgebra alg1
 --     , FunctorTag (ValidView alg1 alg2 t) (Sig alg1)
     , FunctorTag (Sig alg1)
@@ -221,7 +224,7 @@ embedExpr :: forall alg1 alg2 cxt (s::[AT]) t a.
 --         (ValidView alg1 alg2 t)
         (Free (Sig alg1) t a)
         (Free (Sig alg2) t a)
-embedExpr = HaskTag go
+embedAST = HaskTag go
     where
         go :: forall (s::[AT]).
             ( FAlgebra alg1
@@ -239,12 +242,13 @@ embedExpr = HaskTag go
             (Free f) -> appFree' @s @t @a @alg2
                 $ Free
                 $ embedSig
---                 $ _ embedExpr f
-                $ fmapTag  embedExpr f
+--                 $ _ embedAST f
+                $ fmapTag  embedAST f
 -- --             (FreeTag f) -> appFree' @s @t @a @alg2
 -- --                 $ FreeTag
 -- --                 $ embedSig
--- --                 $ fmapTag @(ValidView alg1 alg2 t) embedExpr f
+-- --                 $ fmapTag @(ValidView alg1 alg2 t) embedAST f
+-}
 
[…]

### 2016-06-08 src/Homogeneous/FAlgebra.hs

@@ -390,11 +390,6 @@ law2tasty :: forall (a :: Type) alg.
     , Testable a
     ) => Proxy a -> Law alg -> TestTree
 law2tasty p law = QC.testProperty (lawName law) $ law2property p law
--- law2tasty p law = QC.testProperty (lawName law) $ do
---     as <- infiniteListOf (arbitrary::Gen a)
---     let varmap = zip (toList (lhs law) ++ toList (rhs law)) as
---     return $ (runAST $ subVars (lhs law) varmap)
---           == (runAST $ subVars (rhs law) varmap)
 
 class2tasty :: forall alg a.
     ( Variety alg

### 2016-06-08 src/TemplateHaskellUtils.hs

@@ -44,6 +44,12 @@ isVarT :: TH.Type -> Bool
 isVarT (VarT _) = True
 isVarT _        = False
 
+isConcrete :: TH.Type -> Bool
+isConcrete (VarT _) = False
+isConcrete (ConT _) = True
+isConcrete (AppT a1 a2) = isConcrete a1 && isConcrete a2
+isConcrete t = error ("isConcrete: t="++show t)
+
 -- | Given a type that stores a function:
 -- return a list of the arguments to that function,
 -- and discard the return value.

### 2016-06-09 paper/paper.tex
more paper edits

@@
 
-\usepackage{amsmath}
+\usepackage{amsmath,amsfonts}
 
@@
 \newtheorem{problem}{Problem}
+\newtheorem{defn}{Definition}
 
@@
 
+@inproceedings{jones2001playing,
+  title={Playing by the rules: rewriting as a practical optimisation technique in GHC},
+  author={Jones, Simon Peyton and Tolmach, Andrew and Hoare, Tony},
+  booktitle={Haskell workshop},
+  volume={1},
+  pages={203--233},
+  year={2001}
+}
+
 @book{higham2002,
@@
 }
+
+@misc{viewpatterns,
+  author = {GHC Wiki},
+  title = {View patterns: lightweight views for Haskell},
+  url = {https://ghc.haskell.org/trac/ghc/wiki/ViewPatterns},
+  note = {Accessed: 2016-06-09}
+}
+
+@misc{patternsynonyms,
+  author = {GHC Wiki},
+  title = {Pattern Synonyms},
+  url = {https://ghc.haskell.org/trac/ghc/wiki/PatternSynonyms},
+  note = {Accessed: 2016-06-09}
+}
 \end{filecontents}
@@
 \begin{abstract}
-We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST),
-and the resulting AST can be converted back into the original function.
-Homoiconicity is a useful property because it lets libraries implement arbitrary program transformations on these functions by manipulating the AST.
-This pearl shows that many Haskell functions are homoiconic.
-The technique does not rely on template Haskell,
-and so works on expressions available at either compile or run time.
-To motivate this homoiconicity,
+We say a function is \emph{homoiconic} if it can be converted into an isomorphic abstract syntax tree (AST).
+This AST can then be manipulated arbitrarily and converted back into a function.
+Template Haskell gives a \emph{compile time} form of homoiconicity to Haskell programs;
+but in this pearl, we demonstrate another form of homoiconicity that relies on polymorphism and GHC's type programming capabilities.
+To motivate our technique,
 we construct a small library that automatically transforms numerically unstable code into stable code.
@@
 Such an inaccurate answer is unacceptable,
-so our function must be rewritten.
+so the function must be rewritten.
 
 But like all good programmers, you're lazy.
-Your machine learning library will contain thousands of numeric functions,
+Your machine learning library will contain hundreds of numeric functions,
 all of which will need to be carefully written to avoid numeric instabilities.
@@
 
-In this pearl, we see how to write functions like |stabilize|.
-The key observation is that certain Haskell functions can be converted into an isomorphic abstract syntax tree (AST).
-The |stabilize| function uses this fact to convert its input function into an AST;
-manipulates that AST to remove numerical stabilities;
-then converts the AST back into a callable function.
-
-In Section \ref{sec:homogeneous}, we introduce the idea of a \emph{homogeneous} function and show that these functions are homoiconic.
-In Section \ref{sec:stabilize}, we use the construction above to implement several versions of the |stabilize| function.
-In Section \ref{sec:heterogeneous}, we introduce the idea of \emph{heterogeneous} functions and show that they are also homoiconic.
-Finally, we conclude by introducing a Haskell package called {\ttfamily homoiconic} that makes it easy for you to .
+In this pearl, we see how to write functions like |stabilize| that perform program transformations.
+These functions follow the following pattern.
+First, convert the input function into an abstract syntax tree (AST).
+Next, manipulate the AST to remove the numerical instabilities.
+Finally, convert the AST back into a callable function.
+%The key observation is that certain Haskell functions are \emph{homoiconic}.
+%That is, we can convert them into isomorphic abstract syntax trees (ASTs).
+%The key observation is that certain Haskell functions (which we call \emph{homogeneous} and \emph{heterogeneous}) can be converted into equivalent abstract syntax trees (ASTs).
+%The |stabilize| function converts its input function into an AST,
+%then manipulate that AST to remove numerical stabilities,
+%and finally convert the AST back into a callable function.
+
+The main idea of the |homoiconic| library is that we can associate with every type class an FAlgebra that captures the algebraic structure of that class.
+
+In Section \ref{sec:homogeneous},
+%In Section \ref{sec:homogeneous}, we introduce the idea of a \emph{homogeneous} function and show that these functions are homoiconic.
+%In Section \ref{sec:stabilize}, we use the construction above to implement several versions of the |stabilize| function.
+%In Section \ref{sec:heterogeneous}, we introduce the idea of \emph{heterogeneous} functions and show that they are also homoiconic.
+%Finally, we conclude by introducing a Haskell package called {\ttfamily homoiconic} that makes it easy for you to .
 
@@
 \label{sec:homogeneous}
-We call a function \emph{homogeneous} if:
+
+In this section we first define homogeneous type classes and functions.
+There are two intuitive ideas for homogeneous type classes.
+First, they have ASTs that contain only a single type variable.
+Second, we can construct an FAlgebra for these type classes.
+
+\begin{defn}
+We call a type class homogeneous if:
 \begin{enumerate}
 \item
-%The function's return type and all of its parameters are the same type variable.
-The function is polymorphic in its return type.
-\item
-The function's parameters are either the same as the return type or they are concrete.
-\item
-The constraints contain exactly one type class, and
-that type class has exactly one parameter,
-which is the function's return type.
+it has a single parameter; and
+\item
+the class's constraints contain only homogeneous type classes applied directly to the parameter.
 \end{enumerate}
-The |logLogistic1| function is homogeneous.
-Figure \ref{code:ghc} shows many more examples from GHC's numeric and ordering hierarchies.
-
-In this section, we will show that every type class has an associated monad representing the abstract syntax tree of the class's homogeneous functions.
-Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
-See for example \cite{swierstra2008}.
+For example, all of the classes in the Prelude are homogeneous.
+Some of these classes are shown in Figure \ref{code:ghc}.
+Figure 2 (near the end of the paper) shows type classes which are not homogeneous due to the presence of type families in the constraints.
+\end{defn}
+
+\begin{defn}
+We call a function homogeneous if:
+\begin{enumerate}
+\item
+there is exactly one type variable in the signature, which we denote by |a|;
+\item
+each of the function's parameters is either
+\begin{enumerate}
+\item
+|a|, or
+\item
+concrete;
+\end{enumerate}
+\item
+the function's return type is |a|; and
+\item
+the constraints contain only homogeneous type classes applied to |a|.
+\end{enumerate}
+The |logLogistic1| function is homogeneous,
+and Figure \ref{code:ghc} shows many more examples taken from the Prelude's numeric hierarchy.
+\end{defn}
+
+%In this section, we will show that every type class has an associated monad representing the abstract syntax tree of the class's homogeneous functions.
+%Our construction is closely related to other constructions involving initial algebras, FAlgebras, and free monads.
+%See for example \cite{swierstra2008}.
+
+%Our goal in this section is to convert homogeneous functions into ASTs.
+%These ASTs will contain only a single parameter type, and so are relatively simple.
+%The construction will use FAlgebras\cite{swierstra2008}.
 
@@
 \subsection{A type class for FAlgebras}
-FAlgebras are a technique used in the mathematical fields of universal algebra and category theory for defining algebraic structures.
-We represent an FAlgebra with the following type class.
+Every homogeneous type class has an associated FAlgebra.
+We represent this FAlgebra with the following type class.
 \begin{lstlisting}
@@
 
-It is easiest to understand instances of |FAlgebra| by walking through an example.
+It is easiest to understand the |FAlgebra| class by walking through an example instance.
 Below is the instance for |Fractional|.
@@
 Each |Sig| data instance can have many constructors,
-which can be divided into two categories.
-First, there should be one constructor for each homogeneous class function.
-This constructor should have the same parameters as the function.
-For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three member functions |(/)|, |recip|, and |fromRational|.
-Second, there should be a constructor for each superclass.
-This constructor should store the |Sig| of the corresponding superclass.
+and these constructors come in two flavors.
+First is the \emph{function constructor}.
+There should be one function constructor for each homogeneous class function, and
+this constructor should have the same parameters as the function.
+For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three member functions |(/)|, |recip|, and |fromRational| respectively.
+The second constructor flavor is the \emph{superclass constructor}.
+There should be one superclass constructor for each superclass, and
+this constructor should store the |Sig| of the corresponding superclass.
 For the |Fractional| example, we have the |Sig_Fractional_Num| constructor corresponding to the |Num| superclass.
@@
 The |runSig| class method gives us a way to evaluate any |Sig|.
+Mathematicians sometimes call this function simply an \emph{algebra}.
 For each function constructor, |runSig| should call the corresponding function.
 For each superclass constructor, |runSig| recursively calls |runSig| on the superclass's |Sig|.
+This is exactly what the |Fractional| instance above does.
 
 The F in FAlgebra comes from the fact that every signature is actually a functor.
-In Haskell, we encode this by enforcing that |Sig alg| must be an instance of the |Functor| class in order to create a valid |FAlgebra| instance for |alg|.
+In Haskell, we encode this by enforcing that |Sig alg| must be an instance of the |Functor| class.
 As usual, there is only a single valid |Functor| instance.
 For |Fractional|, it is:
-
 \begin{lstlisting}
@@
 \end{lstlisting}
-As usual, there is only a single valid |Functor| instance.
-There are multiple valid |Foldable| instances depending on the order in which |foldr| considers the constructors' parameters,
-but for our purposes the order is not important.
 
@@
 
-For every FAlgebra, there is an associated AST with the corresponding algebraic structure.
+For every FAlgebra, there is an associated AST.
 This AST is sometimes called an \emph{initial algebra},
 and has a standard construction via the \emph{free monad}.
-
 In Haskell, the free monad is defined as
@@
 \begin{lstlisting}
-Free
+expr1 :: AST Num Double
+expr1 = Free
   (Sig_plus
@@
 \end{lstlisting}
+The type signature is not necessary, but shown for clarity.
+
 We can also create a corresponding AST for any class that has |Num| as a superclass.
@@
 \begin{lstlisting}
-Free
+expr2 :: AST Floating Double
+expr2 = Free
   (Sig_Floating_Fractional
@@
 all such relationships should have a corresponding instance.
-For example, |Num| is a superclass of |Floating|,
-so we should have a |View Num Floating| instance.
+This property ensures that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance (via a chain of superclass constructors).
+It is only sometimes true that we can extract a |Sig alg1 a| from a |Sig alg2 a|,
+so the function is partial and hence labeled |unsafe|.
+For example,
+we have the following |View Num Floating| instance because |Num| is a superclass of |Floating| via |Fractional|.
 \begin{lstlisting}
@@
 \end{lstlisting}
-Because of this property, we know that the |Sig alg1 a| data instance can always be embedded into a |Sig alg2 a| data instance.
-But it is only sometimes true that we can extract a |Sig alg1 a|,
-which is why the corresponding function is labeled |unsafe|.
+The |embedSig| function shows us how to embed a |Sig Num| into a |Sig Floating|;
+notice that the inner call to |embedSig| relies on a |View Num Fractional| instance also existing.
+The |unsafeExtractSig| function extracts the |Sig Num| instance only if the input contains one.
 
@@
 \begin{lstlisting}
-Free
+expr3 :: View Floating alg => AST alg Double
+expr3 = Free
   (embedSig
@@
 we can make the AST itself an instance of any |FAlgebra| type class.
-Again, we'll approach the procedure by example.
-We can write a |Num| instance for |AST Num a| as:
-\begin{lstlisting}
-instance View Num alg => Num (Free (Sig alg) a)
-        where
-    (+) e1 e2 = Free (embedSig (Sig_plus e1 e2))
+As an example, the |Fractional| instance is shown below.
+\begin{lstlisting}
+instance View Fractional alg
+    => Fractional (AST alg a) where
+    (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
+    recip e   = Free $ embedSig $ Sig_recip e
+    fromRational r
+        = Free $ embedSig $ Sig_fromRational r
 \end{lstlisting}
@@
 we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.
-The |Fractional| and |Floating| instances follow a similar pattern.
 
@@
 \begin{lstlisting}
-(1+2)+3 :: AST Num a
+expr4 :: AST Num Double
+expr4 = (1+2)+3
 \end{lstlisting}
@@
 \begin{lstlisting}
-(1+2)+3 :: AST Floating a
-\end{lstlisting}
+expr5 = AST Floating Double
+expr5 = (1+2)+3
+\end{lstlisting}
+Notice that in these examples we didn't embed the numbers using the |Pure| constructor.
+Since |AST Num Double| is an instance of |Num|,
+GHC will internally call the |fromInteger| to embed the number.
 
@@
 
-The are no tricks in the |Show| instance for the |Sig| instance.
+There are no tricks in the |Show| instance for |Sig|.
 The instance for |Fractional| is shown below.
@@
 instance (Show a, Show (f (Free f a)))
-    => Show (Free f a)
-        where
+    => Show (Free f a) where
     show (Pure a) = show a
@@
 If the syntax tree contains a branch,
-then put parenthesis around the branch so that we don't lose information about the order of operations, and show the functor in the middle.
-
-Now, when we type an expression into ghci,
+then put parenthesis around the branch,
+and show the functor (i.e. the |Sig|) in the middle.
+
+Now, when we type a numeric expression into ghci without a type signature,
+we evaluate the expression like normal.
 \begin{lstlisting}
@@
 6
-ghci> (1+2)+3 :: AST Floating Int
-(1+2)+3
-\end{lstlisting}
+\end{lstlisting}
+But when we add a type signature,
+we get the full AST.
+\begin{lstlisting}
+ghci> (1+2)+3 :: AST Floating Double
+(((fromInteger 1)+(fromInteger 2))+(fromInteger 3))
+\end{lstlisting}
+Again, the way ghci works, all integer literals automatically have |fromInteger| applied to them,
+which is why this function appears in the AST above.
+It would be easy to create a pretty printer that removes these function calls and excess parentheses,
+but this is left as an exercise for the reader.
+
+\subsection{Showing functions}
+Reconsider our |logLogistic1| function.
+It has type
+\begin{lstlisting}
+logLogistic1 :: Floating a => a -> a
+\end{lstlisting}
+What happens if we specialize |a| to |AST Floating Double|?
+We get a peek at the functions internals.
+\begin{lstlisting}
+ghci> logLogistic1 1 :: AST Floating Double
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate (fromInteger 1))))))
+\end{lstlisting}
+We don't just have to input single numbers;
+we can also input full expressions.
+These expressions will also get displayed,
+intermixed with the original function.
+\begin{lstlisting}
+ghci> logLogistic1 (1*3+4) :: AST Floating Double
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate (((fromInteger 1)*(fromInteger 3))
+    +(fromInteger 4)))))))
+\end{lstlisting}
+%\begin{lstlisting}
+%ghci> logLogistic1 (logLogistic1 1) :: AST Floating Double
+%(log ((fromInteger 1)/((fromInteger 1)+(exp
+    %(negate (log ((fromInteger 1)/((fromInteger 1)
+    %+(exp (negate (fromInteger 1)))))))))))
+%\end{lstlisting}
+To more conveniently display functions,
+we need a variable type.
+\begin{lstlisting}
+newtype Var = Var String
+
+var1 :: AST alg Var
+var1 = Pure "var1"
+\end{lstlisting}
+Now we can show the function properly.
+\begin{lstlisting}
+ghci> logLogistic1 var1 :: AST Floating Var
+(log ((fromInteger 1)/((fromInteger 1)+(exp (negate
+    var1)))))
+\end{lstlisting}
+%At this point, you might be tempted to try to write a |Show| instance for functions.
+%Unfortunately, you can't quite do that with this technique.
 
 \subsection{Evaluating the syntax tree}
-Another useful task for ASTs is to evaluate them,
-generating the value they represent.
+Our last task is to evaluate the ASTs.
+That is, we need to run them to generate the value the tree represents.
+
 To do this, we will take advantage of the free monad's structure.
@@
 
-First, we construct a function that ``runs'' the syntax tree, converting it into a single value.
+The following function evaluates the syntax tree.
 \begin{lstlisting}
@@
 
-\subsection{Homoiconicity at last!}
-We are finally ready to see that our construction makes homogeneous functions homoiconic.
-If we open GHCi and enter the following expression
-\begin{lstlisting}
-ghci> let ast=associator (Pure a1) (Pure a2) (Pure a3)
-\end{lstlisting}
-the display shows our AST
-\begin{lstlisting}
-\end{lstlisting}
-We can also evaluate our AST and see that doing so gives the same results as running the original expression.
-\begin{lstlisting}
-ghci> runAST ast
-
-ghci> associator a1 a2 a3
-
-\end{lstlisting}
-We are now ready to use our ASTs to make working with numeric functions easier.
-%\subsection{The Prelude's Numeric Hierarchy}
+When we evaluate an AST for an expression,
+we get the same result as if we had just evaluated the expression directly.
+\begin{lstlisting}
+ghci> logLogistic1 10 :: Double
+-4.539889921682063e-5
+
+ghci> runAST (logLogistic1 10 :: AST Floating Double)
+-4.539889921682063e-5
+\end{lstlisting}
+And this demonstrates that homogeneous functions are in fact homoiconic.
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
 
@@
 
-We've seen how to easily construct the ASTs,
-now it's time to perform computations on them.
-The standard way to deconstruct an AST is via pattern matching.
-We'll use the |ViewPatterns| and |PatternSynonyms| language extensions to simplify this task.
-Then we'll see three examples of improving floating point accuracy.
+We're now ready to modify our programs by manipulating the ASTs.
+The standard way to deconstruct an AST Haskell is via pattern matching.
+In this section, we'll first demonstrate how to pattern match our ASTs using the |ViewPatterns| and |PatternSynonyms| language extensions.
+Then we'll see three simple examples of program transformations implemented in this way.
+
+Automatically stabilizing floating point expressions is an active area of research.
+One way of doing this is to create a library of program transformations that fix many different types of expressions.
+Then combining everything in the library on your problem.
+In this section, we'll create a very small library for these floating point issues,
+and don't claim to have completely solved the floating point problem.
+This is simply a motivating example for homoiconicity.
 
@@
 
-We'll begin with a very simple example.
-Consider the following simple, but numerically unstable function.
+Consider the following simple,
+but numerically unstable function.
 \begin{lstlisting}
 testFunc1 :: Floating a => a -> a
-testFunc1 = log(exp a)
+testFunc1 = log (exp a)
 \end{lstlisting}
@@
 But clearly this is not the correct answer.
-Mathematically, the $\log$ is the inverse of $\exp$, so the final result should be 711.
-
-First, we'll see how to properly manipulate the AST to fix this problem.
-%We need a function of type |AST Floating a -> AST Floating a|.
-We're trying to convert the expression |log (exp x)| into simply |x|.
+Mathematically, the $\log$ is the inverse of $\exp$, so the final result should be 710.
+We'll write a program transformation that fixes this function,
+then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.
+
+Our goal is to convert the expression |log (exp x)| into just |x|.
 To do this, we can just pattern match on the AST for the left hand side of the expression,
@@
 \begin{lstlisting}
-logexpAST1 ::
-    AST Floating a -> AST Floating a
-logexpAST1
-    (Free (Sig_log (Free (Sig_exp a)))) = a
-logexpAST1 a = a
-\end{lstlisting}
-This function has two problems.
-First, our pattern matching expression is large and obnoxious.
-Second, it's not generic.
-
-We can fix the second problem using the |ViewPatterns| language extension.
+logexpAST1 :: AST Floating a -> AST Floating a
+logexpAST1 (Free (Sig_log (Free (Sig_exp a)))) = a
+\end{lstlisting}
+This function works, but has some warts.
+First, our pattern contains noise from the need to alternate between the |Free| and |Sig_| constructors.
+Second, it is specialized to the |Floating| type class,
+so it won't work on any subclasses of |Floating| when it should be able to.
+
+We'll start by fixing the second problem using the |ViewPatterns| language extension.
+We will provide a very brief description of this GHC extension here;
+for a more comprehensive description, see the GHC wiki \cite{viewpatterns}.
+|ViewPatterns| lets us use functions within a pattern,
+and we pattern match on the return value of the function.
+Because functions can be polymorphic,
+we can use the |ViewPatterns| extension to make our pattern accept multiple constructors.
+Fortunately, we've already defined a function that extracts a |Sig| from its subclasses.
+\begin{lstlisting}
+unsafeExtractSig :: View alg1 alg2
+    => Sig alg2 a -> Sig alg1 a
+\end{lstlisting}
+By using this function in the pattern, we can write a polymorphic version of our program transformation.
 \begin{lstlisting}
@@
     = a
-logexpAST2 a = a
-\end{lstlisting}
-But this made our pattern matches even uglier!
-Fortunately, we can use the |PatternSynonyms| extension to greatly simplify these patterns.
-\footnote{
-    %These patterns can be made bidirectional using the |embedSig| function.
-    For ease of presentation, only the unidirectional case is shown.
-}
-\begin{lstlisting}
-pattern AST_log ::
-    ( View Floating alg
-    ) => AST alg a -> AST alg a
+\end{lstlisting}
+This function now has the correct type signature (so we've solved problem 2),
+but the pattern matches are even uglier (so problem 1 is worse).
+Fortunately, we can use the |PatternSynonyms| extension to simplify the patterns.
+Again, see the GHC wiki for a complete description of this language extension \cite{patternsynonyms}.
+
+We define a pattern synonym for each function contained in an FAlgebra.
+Our naming convention is to prefix |AST_| to the name of the function.
+These pattern synonyms combine the |Free| constructor with the view pattern from the previous example.
+The synonyms for the |log| and |exp| functions are shown below.
+\begin{lstlisting}
+pattern AST_log :: View Floating alg
+    => AST alg a -> AST alg a
 pattern AST_log e
@@
 
-pattern AST_exp ::
-    ( View Floating alg
-    ) => AST alg a -> AST alg a
+pattern AST_exp :: View Floating alg
+    => AST alg a -> AST alg a
 pattern AST_exp e
@@
 \end{lstlisting}
-Armed with these pattern synonyms, we can rewrite our function in the much simpler form below.
+Armed with these pattern synonyms,
+we can rewrite our function in the much simpler form below.
 \begin{lstlisting}
@@
 logexpAST3 (AST_log (AST_exp a)) = a
-logexpAST3 a = a
-\end{lstlisting}
-Now that we can manipulate the AST, let's apply it to our function.
-\begin{lstlisting}
-transform1 :: forall alg x.
-    ( FAlgebra alg
-    , alg x
-    ) => (AST alg x -> AST alg x)
-      -> (AST alg x -> AST alg x)
-      -> (x -> x)
-transform1 go f x = runAST $ go $ f (Pure x)
-\end{lstlisting}
-
-But is our function really correct?
+\end{lstlisting}
+This simple function works in the simple motivating test case:
+\begin{lstlisting}
+ghci> logexpAST3 $ testFunc1 var1 :: AST Floating Var
+var1
+\end{lstlisting}
+%stopzone
+but it fails to work in this slightly more complex function:
 \begin{lstlisting}
 testFunc2 :: Floating a => a -> a
-testFunc2 = 1+log(exp a)
-\end{lstlisting}
-What if we call
-
+testFunc2 a = 1+exp(log a)
+\end{lstlisting}
+We haven't told our program transformations what to do when they don't pattern match!
+The correct answer is that we want them to recurse into the subexpressions,
+applying their program transformations there as well.
+In the free monad, we handle these recursions using our |Functor| instance and calls to |fmap|.
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-\begin{lstlisting}
-testFunc :: Floating a => a -> a
-testFunc a = log(exp a)
-
-ghci> logexpAST4 (testFunc 2) :: AST Floating Double
-((fromInteger 1)+(fromInteger 2))
-
-testFunc2 :: Floating a => a -> a
-testFunc2 a = 1+exp(exp(exp(log(log(log a)))))
-
-ghci> fixAST stabilizeAST
-    (testFunc2 var1 :: AST Floating Var)
+Now we correctly recurse into subexpressions.
+\begin{lstlisting}
+ghci> logexpAST4 $ testFunc2 var1 :: AST Floating Var
 ((fromInteger 1)+var1)
 \end{lstlisting}
-
-\begin{lstlisting}
-fixAST :: (Eq (Sig alg (Free (Sig alg) a)), Eq a) => (AST alg a -> AST alg a) -> AST alg a -> AST alg a
+%stopzone
+There's still one last wrinkle.
+What if we need to apply the program transformation multiple times?
+\begin{lstlisting}
+ghci> logexpAST4 $ testFunc3 var1
+    :: AST Floating Var
+((fromInteger 1)+(log (log (exp (exp var1)))))
+\end{lstlisting}
+%stopzone
+We want to apply the program transformation repeatedly until it stops changing the AST.
+This is called the fixed point of the transformation.
+The |fixAST| function below does this.
+\begin{lstlisting}
+fixAST ::
+    ( Eq (Sig alg (Free (Sig alg) a))
+    , Eq a
+    ) => (AST alg a -> AST alg a)
+      -> (AST alg a -> AST alg a)
 fixAST f ast = if ast==ast'
@@
 \end{lstlisting}
+Now we can correctly modify the |testFunc3| function.
+\begin{lstlisting}
+ghci> fixAST logexpAST4 $ testFunc3 var1
+    :: AST Floating Var
+((fromInteger 1)+var1)
+\end{lstlisting}
+%stopzone
 
@@
 The rules in in the previous section are relatively simple.
-They could just as easily have been written using GHC's rewrite rules syntax.
+They could just as easily have been written using GHC's rewrite rules syntax \cite{jones2001playing}.
 Rewrite rules are the primary mechanism for optimising source code in GHC.
@@
 
-\begin{lstlisting}
-foldConstants :: View Num alg => AST alg a -> AST alg a
-foldConstants (AST_plus   (AST_fromInteger a1) (AST_fromInteger a2)) = AST_fromInteger (a1+a2)
-foldConstants (AST_minus  (AST_fromInteger a1) (AST_fromInteger a2)) = AST_fromInteger (a1-a2)
-foldConstants (AST_mul    (AST_fromInteger a1) (AST_fromInteger a2)) = AST_fromInteger (a1*a2)
-foldConstants (AST_negate (AST_fromInteger a1)) = AST_fromInteger (negate a1)
-foldConstants (Free sig) = Free (fmap foldConstants sig)
-foldConstants (Pure a  ) = Pure a
-\end{lstlisting}
-
-\begin{lstlisting}
-constExpr :: Num a => a
-constExpr = 4+2*(8-2)-1
-
--- ghci> fixAST foldConstants constExpr
--- (fromInteger 18)
-
+Our program transformations, however, are powerful enough to fold constants.
+The following function performs constant folding on expressions containing addition and multiplication of integers.
+\begin{lstlisting}
+foldConstants :: View Num alg
+    => AST alg a -> AST alg a
+foldConstants (AST_plus
+    (AST_fromInteger a1)
+    (AST_fromInteger a2))
+    = AST_fromInteger (a1+a2)
+foldConstants (AST_mul
+    (AST_fromInteger a1)
+    (AST_fromInteger a2))
+    = AST_fromInteger (a1*a2)
+foldConstants (Free sig)
+    = Free (fmap foldConstants sig)
+foldConstants (Pure a) = Pure a
+\end{lstlisting}
+It is straightforward to extend this function to work for other operations,
+but for space reasons these were omitted.
+
+Here's an example of our constant folding in action.
+First define a simple function.
+\begin{lstlisting}
 constFunc :: Num a => a -> a -> a
 constFunc x1 x2 = x1*2-(7-2)*x2
-
--- ghci> fixAST foldConstants (constFunc 2 3)
--- (fromInteger -11)
--- ghci> fixAST foldConstants (constFunc var1 var2)
--- ((var1*(fromInteger 2))-((fromInteger 5)*var2))
+\end{lstlisting}
+Then, if we call this function with two known parameters,
+our |foldConstants| function is able to take it down to a single constant.
+\begin{lstlisting}
+ghci> fixAST foldConstants (constFunc 2 3)
+    :: AST Floating Double
+(fromInteger 31)
+\end{lstlisting}
+If we instead invoke the function with only
+\begin{lstlisting}
+ghci> fixAST foldConstants (constFunc var1 var2)
+    :: AST Floating Var
+((var1*(fromInteger 2))+((fromInteger 9)*var2))
+\end{lstlisting}
+We can even invoke the function with only a single known parameter,
+and have that constant get folded into the final expression.
+\begin{lstlisting}
+ghci> fixAST foldConstants (constFunc var1 3)
+    :: AST Floating Var
+((var1*(fromInteger 2))+(fromInteger 27))
 \end{lstlisting}
@@
 
-\begin{lstlisting}
-stabAST :: Eq a =>
+Our original motivation was to stabilize the |logLogistic1| function.
+Stabilizing this function is a bit harder because it actually requires changing the type signature of the function by adding an |Ord| instance.
+One of the requirements of a homogeneous function is that it have only a single class in the context.
+To accommodate this condition, we'll create a new type class that has our desired types in the constraint.
+\begin{lstlisting}
+class (Floating a, Ord a) => FloatingOrd a
+instance {-#OVERLAPPABLE#-} (Floating a, Ord a) => FloatingOrd a
+\end{lstlisting}
+Note that the |OVERLAPPABLE| pragma is required so that we can create an |Ord| instance for |AST FloatingOrd a|.
+
+Now we can define our transformation as
+\begin{lstlisting}
+stabilize :: Eq a =>
     AST FloatingOrd a -> AST FloatingOrd a
-stabAST
+stabilize
     (AST_log
@@
     )
-    = logLogistic2 x
+    = m+log(1/(exp(m)+exp(-x+m)))
+        where
+            m = min 0 x
 \end{lstlisting}
@@
 
-%\begin{lstlisting}
-%data Vec3 a = Vec3 a a a
-    %deriving (Eq,Show)
-%\end{lstlisting}
-
-One of the main limitations of GHC's numeric hierarchy is that it was designed for scalar numbers.
-A logistic regression implementation needs vectors.
-There's been many proposals to extend GHC's built-in hierarchy to handle vectors.
-A very simple such extension is shown in Figure \ref{code:vector}.
-This extension uses type families.
+The next step in implementing logistic regression is implementing the logistic loss function
+\begin{align}
+l &: \mathbb{R}^n \times \mathbb{R}^n \to \mathbb{R} \\
+l &(v_1,v_2) = f(<v_1,v_2>)
+\end{align}
+This function takes two vectors as inputs and returns a scalar.
+Unfortunately, GHC's numeric hierarchy has no notion of vectors.
+So you implement a simple extension to the hierarchy shown in Figure \ref{code:vector}.
 
@@
 
-class (Num a, Fractional (Scalar a)) => Vector a where
+class (Num a, Fractional (Scalar a))
+    => Vector a where
     (.*) :: Scalar a -> a -> a
@@
 
+You can now implement the logistic loss as:
 \begin{lstlisting}
@@
 \end{lstlisting}
-
-The |logLoss|, |(.*)|, and |dotProduct| functions are not homogeneous because of the type families.
-Heterogeneous functions are a relaxation of homogeneous functions.
-We call a function \emph{heterogeneous} if:
+Unfortunately, this function is not homogeneous due to the application of a type family in the return type.
+So we cannot use the construction of Section \ref{sec:homogeneous} to generate an AST for this function.
+In this section, we'll introduce the more general notion of heterogeneous functions and show how to construct ASTs for them.
+
+Informally, the key difference between a homogeneous and heterogeneous FAlgebra is that homogeneous FAlgebras contain functions only over a single type; whereas heterogeneous FAlgebras contain functions over many different types.
+Formally, we call a function \emph{heterogeneous} if:
 \begin{enumerate}
 \item
-The function is polymorphic in its return type.
-\item
-The function's parameters are either the same as the return type, they are concrete.
+The function's signature contains a single type variable, which we'll call |a|.
+\item
+The function returns either |a| or a type family applied to |a|.
+\item
+%The function's parameters are either the same as the return type, they are concrete.
+Each of the function's parameters is either:
+\begin{enumerate}
+\item
+|a|,
+\item
+a type family applied to |a|, or
+\item
+concrete.
+\end{enumerate}
 \item
@@
 \end{enumerate}
+The construction of ASTs follows for heterogeneous types follows a similar pattern as for homogeneous types, but involves considerably more type hackery.
 
 \subsection{Dealing with type families}
-
-\begin{lstlisting}
-type AT = Type
-\end{lstlisting}
-
+Before we're ready to construct the ASTs, we need a short diversion to discuss type families.
+Type families are functions at the type level in Haskell.
+Unfortunately, they're not as convenient to work with.
+In particular, type families cannot be partially applied.
+
+To simulate partial application, we will create a new data type for each type family.
+The naming convention is to put a |T| in front of the family's name.
+For example, we declare
+\begin{lstlisting}
+data TScalar
+\end{lstlisting}
+for the type family |Scalar|.
 \begin{lstlisting}
 type family App (t::k) (a::Type) :: Type
+\end{lstlisting}
+
+Repeated applications of type families will be stored in a promoted list.
+\begin{lstlisting}
 type instance App '[]       a = a
@@
 
+Since |TScalar| is an ordinary Haskell type, it has kind |Type|.
+Ideally, we would like all of these shadow types to inhabit the same kind.
+But this would require open kind families,
+an extension that has been proposed\footnote{} but not implemented.
+To make our code easier to read, we will create the following kind synonym.
+\begin{lstlisting}
+type AT = Type
+\end{lstlisting}
+
+\subsection{The problems}
+
+There are three problems with our homogeneous FAlgebra encoding that make it unsuitable for heterogeneous FAlgebras.
+
+\begin{problem}[the superclass problem]
+The |Vector| class has a superclass constraint on a type family.
+If we follow the formulation for making a homogeneous FAlgebra, we get
+\begin{lstlisting}
+instance FAlgebra Vector where
+    data Sig Vector a
+        = Sig_dotmul (Scalar a) a
+        | Sig_Vector_Num (Sig Num a)
+        | Sig_Vector_Floating (Sig Floating (Scalar a))
+    runSig (Sig_dotmul s a) = s.* a
+    runSig (Sig_Vector_Num s) = runSig s
+    runSig (Sig_Vector_Floating s) = runSig s
+\end{lstlisting}
+This definition does not type check.
+In the last line of |runSig|'s definition,
+the type checker will try to match the |a| type with |Scalar a| and fail.
+\end{problem}
+
+\begin{problem}[the parameter problem]
+Now consider what happens when you try to write the |Functor| instance
+\begin{lstlisting}
+instance Functor (Sig Vector) where
+    fmap f (Sig_dotmul s a) = Sig_dotmul (f s) (f a)
+    fmap f (Sig_Vector_Num s) = Sig_Vector_Num (f s)
+    fmap f (Sig_Vector_Floating s) = Sig_Vector_Floating (f s)
+\end{lstlisting}
+Here type checking fails on the last line for the same reason as above,
+but also fails on the first line due to the presence of a type family application in the function's parameter.
+\end{problem}
+
+\begin{problem}[the return type problem]
+Consider for a moment the |Hilbert| class in Figure \ref{code:vector}.
+The |dotProduct| function is not homogeneous because of the type family in the return type.
+What happens if we try to make |Hilbert| and instance of |FAlgebra| anyways?
+We would get something like:
+\begin{lstlisting}
+instance FAlgebra Hilbert where
+    data Sig Hilbert a
+        = Sig_Hilbert_Module (Sig Module a)
+        | Sig_dotProduct a a
+    runSig (Sig_Hilbert_Module s) = runSig s
+    runSig (Sig_dotProduct a1 a2) = dotProduct a1 a2
+\end{lstlisting}
+This code doesn't type check.
+|runSig| expects its return type to be |a|,
+but |dotProduct| returns a |Scalar a|.
+\end{problem}
+
 \subsection{Heterogeneous FAlgebras}
 
-\begin{lstlisting}
-class FunctorTag (f::[AT]->Type->Type) where
-    fmapTag :: HaskTag a b -> f t a -> f t b
-\end{lstlisting}
-
-\begin{lstlisting}
-class HetFunctor (Sig alg) => FAlgebra (alg::Type->Constraint) where
+Our new |FAlgebra| class has a number of modifications to solve the problems mentioned above.
+It is defined as:
+\begin{lstlisting}
+class HetFunctor (Sig alg) => FAlgebra alg where
     data Sig alg (t::[AT]) a
 
-    runSigTag :: alg a => proxy a -> Sig alg (s ':  t) (App t a) -> App (s ':  t) a
-    runSig    :: alg a => proxy a -> Sig alg        t  (App t a) -> App        t  a
+    runSig    :: alg a => proxy a
+        -> Sig alg        t  (App t a) -> App        t  a
+    runSigTag :: alg a => proxy a
+        -> Sig alg (s ':  t) (App t a) -> App (s ':  t) a
+\end{lstlisting}
+
+The most important change is that the |Sig| data family has been given a new type parameter |t|.
+This parameter indexes the return type of the operation stored in |Sig|.
+Notice that |t| has kind |[AT]|.
+If there is no type family application, then |t| will be the empty list |'[]|;
+if there is a single application by data constructor |A|, |t| will be |'[A]|;
+and if the return type is |A (B a)|, then |t| will be |'[A,B]|.
+
+When defining instances of the |Sig| family, we will need to use GADTs to enforce that the |t| parameter is set correctly.
+|Sig|'s constructors can be classified into the same categories as before.
+For superclass constructors, if the class constraint does not involve a type family,
+then we get defined as:
+\begin{lstlisting}
+Sig_Vector_Num :: Sig Num t a -> Sig Vector t a
+\end{lstlisting}
+if the class constraint does involve a type family,
+then we prepend the type family to |t| in the constructor's return type.
+\begin{lstlisting}
+Sig_Vector_Floating_Scalar
+    :: Sig Floating t a -> Sig Vector (TScalar ': t) a
+\end{lstlisting}
+Function constructors are likewise divided into two cases.
+If the function's return type does not use a type family,
+then
+\begin{lstlisting}
+Sig_dotmul :: Scalar a -> a -> Sig Vector '[] a
+\end{lstlisting}
+If the function's return type does use a type family,
+then
+\begin{lstlisting}
+Sig_dotProduct :: a -> a -> Sig Hilbert '[TScalar] a
+\end{lstlisting}
+These changes to the |Sig| type are the core of the new addition.
+There will be many changes to all of the other functions,
+but they all stem from adapting the same techniques used in the homogeneous case to the updated |Sig|.
+
+The |runSig| function has been split into two functions, |runSig| and |runSigTag|.
+First we'll explain the |runSig| function, then see why its deficiencies necessitate the |runSigTag| function.
+Recall that the |t| type variable represents the type families that have been applied to the content of the |Sig|.
+So when we run the |Sig|, the content needs to reflect those applications.
+Thus, the parameter to |runSig| is |Sig alg t (App t a)|.
+Similarly, the return type needs to reflect those applications plus whatever applications are present in the function.
+And so the return type is |Sig alg t (App t a)|.
+Because the |a| type variable appears only inside type family applications,
+we need the |proxy a| argument for type resolution.
+The |runSig| function is defined exactly as before.
+\begin{lstlisting}
+runSig _ (Sig_Vector_Num s)
+    = runSig (Proxy::Proxy a) s
+runSig _ (Sig_Vector_Floating s)
+    = runSig (Proxy::Proxy (Scalar a)) s
+runSig _ (Sig_dotmul s a) = s.*a
+\end{lstlisting}
+If the function's return type involves a type family application,
+then it is impossible to use the |runSig| function on it.
+The signature is wrong.
+That is why we introduce a second function |runSigTag|.
+\begin{lstlisting}
+runSigTag p (Sig_Hilbert_Vector s) = runSigTag p s
+runSigTag p (Sig_dotproduct a1 a2) = dotproduct a1 a2
 \end{lstlisting}
@@
 
-\begin{lstlisting}
-type AST alg t a = Free (Sig alg) t a
-
+To construct an AST out of the |Sig| functor requires a more complex version of the free monad.
+\begin{lstlisting}
 data Free (f::[AT]->Type->Type) (t::[AT]) (a::Type) where
@@
 \end{lstlisting}
-
-\begin{lstlisting}
-class (FAlgebra alg1, FAlgebra alg2) => View alg1 t1 alg2 t2 where
-    embedSig          :: Sig alg1       t1  a -> Sig alg2       t2  a
-    unsafeExtractSig  :: Sig alg2       t2  a -> Sig alg1       t1  a
-\end{lstlisting}
-
-\begin{lstlisting}
-embedSigTag :: View alg1 (t ': t1) alg2 (t ': t2) => Sig alg1 (t ': t1) a -> Sig alg2 (t ': t2) a
-embedSigTag = embedSig
-\end{lstlisting}
+There are now three constructors.
+The |Free| and |Pure| constructors work exactly as they did before.
+The only difference is that |Pure| does not store the |a| type inside of it,
+but instead it applies the |t| type families to |a|.
+The |FreeTag| constructor represents a branch like the |Free| constructor,
+but it also represents a change in return type.
+When embedding a |Sig| into an AST,
+we will use the |FreeTag| constructor whenever the function's return type uses a type family.
+
+The AST is then defined as
+\begin{lstlisting}
+type AST alg t a = Free (Sig alg) t a
+\end{lstlisting}
+
+We can make the AST type instances of the type classes similarly to how we did for homogeneous FAlgebras.
+The main difference is that we will need a new |View| class for converting between |Sig|s.
+This |View| class not only converts the |alg| parameter, but also the |t| parameter.
+\begin{lstlisting}
+class (FAlgebra alg1, FAlgebra alg2)
+    => View alg1 t1 alg2 t2 where
+    embedSig
+        :: Sig alg1 t1 a -> Sig alg2 t2 a
+    unsafeExtractSig
+        :: Sig alg2 t2 a -> Sig alg1 t1 a
+\end{lstlisting}
+
+%\begin{lstlisting}
+%embedSigTag :: View alg1 (t ': t1) alg2 (t ': t2) => Sig alg1 (t ': t1) a -> Sig alg2 (t ': t2) a
+%embedSigTag = embedSig
+%\end{lstlisting}
 
@@
 
+To convert the heterogeneous AST into a |String|,
+we need to write a |Show| instance for |Free| and a show instance for each |Sig|.
+Both of these instances are considerably more complicated than before.
+
+The |Show| instance for the |Sig| type follows the same pattern as before.
+For each function constructor,
+simply show the constructor's arguments connected in an appropriate way by the function's name.
+For each superclass constructor, show the contained constructor.
+\begin{lstlisting}
+instance
+    ( Show a
+    , Show (Scalar a)
+    ) => Show (Sig Vector t a)
+        where
+    show (Sig_Vector_Num s) = show s
+    show (Sig_Vector_Floating s) = show s
+    show (Sig_dotmul a1 a2) = show a1++".*"++show a2
+\end{lstlisting}
+We had to add the |Show (Scalar a)| constraint above because the |a1| variable in the last line has type |Scalar a|.
+
+This instance can be defined, but it can't yet be used.
+Recall that we defined
+\begin{lstlisting}
+type Scalar (AST alg t a) = AST alg (TScalar ': t) a
+\end{lstlisting}
+This means that the constraint |Show (Scalar a)| will use the same |Show| instance as the one displayed above.
+In order to satisfy that constraint, we need |Show (Show (Scalar a))|.
+This nesting of |Show| constraints continues forever!
+In order to stop this explosion of |Show| constraints, we add the following overlapping instance.
+\begin{lstlisting}
+instance {-#OVERLAPS#-} Show
+    (Sig Module (t1 ': t2 ': t3 ': ts) a) where
+    show _ = "<<overflow>>"
+\end{lstlisting}
+Once there are at least three |Scalar|s applies,
+this instance becomes more specific,
+and is thus selected by GHC.
+Since this instance has no constraints,
+that ends the recursion.
+
+We must perform a related type hack to get the |Show| instance for |Free|.
+The |Show| instance for |Free| is shown below.
 \begin{lstlisting}
@@
 \end{lstlisting}
+The |FreeTag| constructor must be undressed in order to satisfy the |Show| requirement.
 
@@
 
+Running the heterogeneous AST is almost the same as the homogeneous case.
 \begin{lstlisting}
@@
 \end{lstlisting}
+The only difference is we need to add a line for the |FreeTag| parameter,
+and we need to manually specify the type |a| with the |Proxy|.
 

### 2016-06-09 src/Heterogeneous/ExamplePrelude.hs

@@ -1,3 +1,6 @@
+{-# LANGUAGE ViewPatterns #-}
+{-# LANGUAGE PatternSynonyms #-}
+
 module Heterogeneous.ExamplePrelude
     where
 
@@ -37,28 +40,55 @@ mkFAlgebra ''FloatingOrd
 --       -> Free (Sig alg) t a
 --       -> Free (Sig alg) t a
 
-pattern Expr_log :: forall t alg a.
-    ( View Floating '[] alg t
-    ) => Free (Sig alg) t a
-      -> Free (Sig alg) t a
-pattern Expr_log e <- Free ((unsafeExtractSig :: Sig alg      t   (Free (Sig alg) t a)
-                                              -> Sig Floating '[] (Free (Sig alg) t a)
-                            ) -> Sig_log e) where
--- pattern Expr_log e <- Free (unsafeExtractSig -> Sig_log e) --where
-    Expr_log e = Free $ embedSig $ Sig_log e
+-- pattern Expr_negate :: forall t alg a.
+--     ( View Num '[] alg t
+--     ) => Free (Sig alg) t a
+--       -> Free (Sig alg) t a
+-- pattern Expr_negate e <- Free (unsafeExtractSigTag0 -> Sig_negate e) where
+--     Expr_negate e = Free $ embedSig $ Sig_negate e
+
+-- pattern Expr_log :: forall t alg a.
+--     ( View Floating '[] alg t
+--     ) => Free (Sig alg) t a
+--       -> Free (Sig alg) t a
+-- pattern Expr_log e <- Free ((unsafeExtractSig :: Sig alg      t   (Free (Sig alg) t a)
+--                                               -> Sig Floating '[] (Free (Sig alg) t a)
+--                             ) -> Sig_log e) where
+-- -- pattern Expr_log e <- Free (unsafeExtractSig -> Sig_log e) --where
+--     Expr_log e = Free $ embedSig $ Sig_log e
 
 --------------------------------------------------------------------------------
 
 type family Scalar a
 mkAT ''Scalar
 
+type instance Scalar Var = Var
+
+scalar1 :: AST alg '[TScalar] Var
+scalar1 = Pure $ Var "scalar1"
+
 type instance Scalar Float = Float
 type instance Scalar (Free (Sig alg) t a) = Free (Sig alg) (TScalar ': t) a
 
 ----------------------------------------
 
[…]

### 2016-06-09 src/Heterogeneous/FAlgebra.hs

@@ -133,6 +133,8 @@ apHaskTag pt (HaskTag f) = f pt (Proxy::Proxy a) (Proxy::Proxy b)
 class FunctorTag (f::[AT]->Type->Type) where
     fmapTag :: HaskTag a b -> f t a -> f t b
 
+    fmapTag' :: Proxy s -> Proxy a -> Proxy b -> App s a -> App s b
+
 ----------------------------------------
 
 type AST (alg::Type->Constraint) t a = Free (Sig alg) t a
@@ -272,6 +274,11 @@ class Typeable alg => FAlgebra (alg::Type->Constraint) where
          -> Sig alg t (Free (Sig alg') t' a)
          -> Sig alg t (App t' a)
 
+--     mapf :: (TypeConstraints t' a)
+--          => (forall s. Free (Sig alg') s a -> Free (Sig alg') s a)
+--          -> Sig alg t (Free (Sig alg') t' a)
+--          -> Sig alg t (Free (Sig alg') t' a)
+
 ----------------------------------------
 
 class (FAlgebra alg1, FAlgebra alg2) => View alg1 t1 alg2 t2 where
@@ -288,6 +295,11 @@ unsafeExtractSigTag ::
     ( View alg1 '[s] alg2 (s:t)
     ) => Sig alg2 (s ': t) (Free (Sig alg2) t a) -> Sig alg1 '[s] (Free (Sig alg2) t a)
 unsafeExtractSigTag = unsafeExtractSig
+
+unsafeExtractSigTag0 ::
+    ( View alg1 '[] alg2 t
+    ) => Sig alg2 t (Free (Sig alg2) t a) -> Sig alg1 '[] (Free (Sig alg2) t a)
+unsafeExtractSigTag0 = unsafeExtractSig
 
 instance FAlgebra alg => View alg t alg t where
     embedSig = id
@@ -533,9 +545,9 @@ mkFAlgebra algName = do
 
     -- remove functions from decsraw that we can't handle
     let go x = case x of
-            SigD _ sigType -> if not $ isVarT $ getReturnType sigType
+            SigD _ sigType -> if isConcrete $ getReturnType sigType
                 then False
-                else True -- and $ map isVarT $ getArgs sigType
+                else True
     let decs = filter go rawdecs
 
     -- common variables we'll need later
@@ -760,15 +772,16 @@ mkFAlgebra algName = do
                             (VarT _) -> AppE
                                 ( AppE
                                     ( VarE $ mkName "runSigTag" )
[…]

### 2016-06-09 src/Homogeneous/PreludeInstances.hs

@@ -4,6 +4,7 @@
 {-# LANGUAGE PatternSynonyms #-}
 {-# LANGUAGE DataKinds #-}
 {-# LANGUAGE TypeFamilies #-}
+{-# LANGUAGE NoRebindableSyntax #-}
 
 module Homogeneous.PreludeInstances
     where
@@ -124,14 +125,17 @@ logexpAST4 (AST_log (AST_exp a)) = a
 logexpAST4 (Free f) = Free $ fmap logexpAST4 f
 logexpAST4 (Pure a) = Pure a
 
-testFunc :: Floating a => a -> a
-testFunc a = log(exp a)
+testFunc1 :: Floating a => a -> a
+testFunc1 a = log(exp a)
+
+testFunc2 :: Floating a => a -> a
+testFunc2 a = 1+log(exp a)
 
 -- ghci> logexpAST4 $ testFunc 2 :: AST Floating Double
 -- ((fromInteger 1)+(fromInteger 2))
 
-testFunc2 :: Floating a => a -> a
-testFunc2 a = 1+exp(exp(exp(log(log(log a)))))
+testFunc3 :: Floating a => a -> a
+testFunc3 a = 1+log(log(log(exp(exp(exp a)))))
 
 -- ghci> fixAST stabilizeAST (testFunc2 var1 :: AST Floating Var)
 -- ((fromInteger 1)+var1)
@@ -192,13 +196,34 @@ constExpr = 4+2*(8-2)-1
 -- (fromInteger 18)
 
 constFunc :: Num a => a -> a -> a
-constFunc x1 x2 = x1*2-(7-2)*x2
+constFunc x1 x2 = x1*2+(7+2)*x2
 
--- ghci> fixAST foldConstants (constFunc 2 3)
--- (fromInteger -11)
--- ghci> fixAST foldConstants (constFunc var1 var2)
--- ((var1*(fromInteger 2))-((fromInteger 5)*var2))
 
+--------------------------------------------------------------------------------
+
+-- type family Scalar a
+-- -- mkAT ''Scalar
+--
+-- class (Num a, Floating (Scalar a)) => Vector a where
+--     (.*) :: Scalar a -> a -> a
+--
+-- instance FAlgebra Vector where
+--     data Sig Vector a
+--         = Sig_dotmul (Scalar a) a
+--         | Sig_Vector_Num (Sig Num a)
+--         | Sig_Vector_Floating (Sig Floating (Scalar a))
+--     runSig (Sig_dotmul s a) = s.* a
[…]

### 2016-06-09 paper/paper.tex
more paper edits

@@
 \usepackage{amsmath,amsfonts}
+\usepackage{url}
 
@@
 \usepackage{courier}
+%\usepackage{pxfonts}
 \usepackage{listings}
 \lstset{
-    language=Haskell,
+    %language=Haskell,
     basicstyle=\ttfamily\footnotesize,
-    keywordstyle=\ttfamily\footnotesize,
+    keywordstyle=\bfseries\ttfamily\footnotesize,
+    showstringspaces=false,
+    %morekeywords={class,data,type,family,instance, where, ghci},
+    morekeywords={ghci},
     %basicstyle=\footnotesize,
@@
 \newcommand{\cL}{{\cal L}}
+
+%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
+
+\newcommand{\homoiconic}{{\ttfamily homoiconic}~}
 
@@
 
+@inproceedings{schrijvers2009complete,
+  title={Complete and decidable type inference for GADTs},
+  author={Schrijvers, Tom and Peyton Jones, Simon and Sulzmann, Martin and Vytiniotis, Dimitrios},
+  booktitle={ACM Sigplan Notices},
+  volume={44},
+  number={9},
+  pages={341--352},
+  year={2009},
+  organization={ACM}
+}
+
 @book{hamming2012,
@@
   year={2015}
+}
+
+@article{eisenberg2015promoting,
+  title={Promoting functions to type families in Haskell},
+  author={Eisenberg, Richard A and Stolarek, Jan},
+  journal={ACM SIGPLAN Notices},
+  volume={49},
+  number={12},
+  pages={95--106},
+  year={2015},
+  publisher={ACM}
 }
@@
 
+%Imagine you've just downloaded and installed a machine learning library for Haskell.
+%You open up ghci and import the corresponding modules.
+%In the documentation, you see a function called |logLogistic|,
+%so you ask ghci its type.
+%\begin{lstlisting}
+%ghci> :t logLogistic
+%logLogistic :: Floating a => a -> a
+%\end{lstlisting}
+%Now imagine you want to know how the function is implemented.
+%\begin{lstlisting}
+%ghci> logLogistic var1 :: AST Floating Var
+%\end{lstlisting}
+%Oh no!
+%The |logLogistic| function has a numerical instability in it.
+
 Imagine you're writing a machine learning library in Haskell.
-%You decide to start by implementing logistic regression.
-%The first step is to implement the log-logistic function
-%Logistic regression is one of the most important machine learning algorithms,
-%and to implement it, you'll need to implement the log-logistic function
-%To implement the classic logistic regression algorithm,
-%you begin by implementing the log-logistic function
 At some point, you'll need to implement the log-logistic function
@@
 In this section we first define homogeneous type classes and functions.
-There are two intuitive ideas for homogeneous type classes.
+There are two intuitions for homogeneous type classes.
 First, they have ASTs that contain only a single type variable.
@@
 \item
-it has a single parameter; and
+it has a single parameter of kind |Type|; and
 \item
@@
 
-The |runSig| class method gives us a way to evaluate any |Sig|.
+The |runSig| class method evaluates an FAlgebra's signature.
 Mathematicians sometimes call this function simply an \emph{algebra}.
@@
 
-We can also create a corresponding AST for any class that has |Num| as a superclass.
+We can also create an AST for the same expression using any subclass of |Num|.
 Below is the same expression encoded using |Floating|'s |FAlgebra| instance.
@@
 Notice that the |Sig_plus| constructor must be embedded into |Floating|'s |Sig| with calls to |Sig_Floating_Fractional| and |Sig_Fractional_Num|.
-
-Constructing these ASTs by hand is a tedious process.
-There is a considerable amount of boilerplate in the |Floating| AST above as a result of the need to embed the |Sig_plus| constructor.
-We can avoid this boilerplate with the following type class.
-\begin{lstlisting}
-class
-    ( FAlgebra alg1
-    , FAlgebra alg2
-    ) => View alg1 alg2
-        where
+These embeddings create considerable boilerplate,
+making the construction of ASTs by hand a tedious process.
+
+We can avoid this boilerplate with a type class that performs these embeddings for us.\footnote{
+    The name {\ttfamily View} stems from the fact that the class methods are actually what GHC calls ``view patterns.''
+    See Section \ref{sec:viewpatterns} for details on how the {\ttfamily ViewPatterns} language extension lets us use the {\ttfamily View} type class for easier pattern matching on our ASTs.
+}
+\begin{lstlisting}
+class (FAlgebra alg1, FAlgebra alg2)
+    => View alg1 alg2 where
     embedSig         :: Sig alg1 a -> Sig alg2 a
@@
 \end{lstlisting}
-Instances of the |View| class should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
+Instances of |View| should satisfy the property that |alg1| is either equal to |alg2| or is a superclass of |alg2|, and
 all such relationships should have a corresponding instance.
@@
 \end{lstlisting}
-The |embedSig| function shows us how to embed a |Sig Num| into a |Sig Floating|;
-notice that the inner call to |embedSig| relies on a |View Num Fractional| instance also existing.
-The |unsafeExtractSig| function extracts the |Sig Num| instance only if the input contains one.
-
-Armed with the |View| class, we can create a uniform representation for all class's ASTs.
-The expression |(1+2)+3| can now be represented as:
+The |embedSig| function shows us how to embed a |Sig Num| into a |Sig Floating| by calling the appropriate superclass constructors.
+Notice that the recursive call to |embedSig| relies on a |View Num Fractional| instance also existing.
+Similarly, the |unsafeExtractSig| function extracts the |Sig Num| instance by pattern matching on the appropriate superclass constructors.
+
+Armed with this |View| class, we can create a uniform representation for |FAlgebra|'s ASTs.
+The expression |(1+2)+3| can now be represented polymorphically as:
 \begin{lstlisting}
@@
 Using our uniform representation of an AST,
-we can make the AST itself an instance of any |FAlgebra| type class.
+we can make the |AST| type an instance of the homogeneous type classes.
 As an example, the |Fractional| instance is shown below.
@@
 \end{lstlisting}
-Notice that in these examples we didn't embed the numbers using the |Pure| constructor.
-Since |AST Num Double| is an instance of |Num|,
-GHC will internally call the |fromInteger| to embed the number.
+Notice that in these last two examples we did not embed the numbers into the AST with the |Pure| constructor.
+Since |AST| is an instance of |Num|,
+GHC will call |fromInteger| to embed the number for us automatically.
 
@@
 Now that we can create our ASTs, we want to actually do something with them!
-The simplest task is to convert them into a |String| for display.
+In this section we will convert them into a |String| for display.
 This is an easy procedure of writing |Show| instances for the |Sig| and |Free| data types.
@@
 Now, when we type a numeric expression into ghci without a type signature,
-we evaluate the expression like normal.
+ghci will evaluate the expression like normal.
 \begin{lstlisting}
@@
 \end{lstlisting}
-But when we add a type signature,
-we get the full AST.
+But when we add the appropriate type signature,
+the full AST is displayed.
 \begin{lstlisting}
@@
 which is why this function appears in the AST above.
-It would be easy to create a pretty printer that removes these function calls and excess parentheses,
+It would be easy to create a pretty printer that does not display |fromInteger| or the excess parentheses,
 but this is left as an exercise for the reader.
@@
 \end{lstlisting}
-What happens if we specialize |a| to |AST Floating Double|?
-We get a peek at the functions internals.
+What happens if we specialize the |a| parameter to |AST Floating Double|?
+We get a peek at the function's internals.
 \begin{lstlisting}
@@
 \end{lstlisting}
-%\begin{lstlisting}
-%ghci> logLogistic1 (logLogistic1 1) :: AST Floating Double
-%(log ((fromInteger 1)/((fromInteger 1)+(exp
-    %(negate (log ((fromInteger 1)/((fromInteger 1)
-    %+(exp (negate (fromInteger 1)))))))))))
-%\end{lstlisting}
-To more conveniently display functions,
-we need a variable type.
+And we can even display the resulting AST of nested function calls.
+\begin{lstlisting}
+ghci> logLogistic1 (logLogistic1 1)
+    :: AST Floating Double
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate (log ((fromInteger 1)/((fromInteger 1)
+    +(exp (negate (fromInteger 1)))))))))))
+\end{lstlisting}
+It would be nice if the displayed function would display the actual input variables,
+rather than just substituting numbers.
+To accomplish this, we need a variable type.
 \begin{lstlisting}
 newtype Var = Var String
-
+\end{lstlisting}
+Values of type |Var| cannot be plugged directly into the |logLogistic1| function because |Var| is not an instance of |Floating|.
+Fortunately, we have a way to make any type an instance of |Floating| via the |AST Floating| type.
+More generically, we define the following variables.
+\begin{lstlisting}
 var1 :: AST alg Var
 var1 = Pure "var1"
-\end{lstlisting}
-Now we can show the function properly.
+
+var2 :: AST alg Var
+var2 = Pure "var2"
+
+var3 :: AST alg Var
+var3 = Pure "var3"
+\end{lstlisting}
+And we can use them to recover the exact AST of a function.
 \begin{lstlisting}
 ghci> logLogistic1 var1 :: AST Floating Var
-(log ((fromInteger 1)/((fromInteger 1)+(exp (negate
-    var1)))))
-\end{lstlisting}
-%At this point, you might be tempted to try to write a |Show| instance for functions.
-%Unfortunately, you can't quite do that with this technique.
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate var1)))))
+\end{lstlisting}
+At this point, you might be tempted to write a |Show| instance for functions.
+\begin{lstlisting}
+instance
+    ( FAlgebra alg
+    , Show (Sig alg (Free (Sig alg) Var))
+    ) => Show (AST alg Var -> AST alg Var) where
+    show f = show (f var1)
+\end{lstlisting}
+Unfortunately, if you try to use the instance naively
+\begin{lstlisting}
+ghci> logLogistic1
+\end{lstlisting}
+then ghci will complain about ambiguous type variables.
+GHC's defaulting mechanism is not powerful enough for it to properly specialize the function without a type signature.
+So unfortunately this technique doesn't give us a convenient |Show| instance,
+it does give us a slightly more verbose alternative for showing a function.
 
@@
 To do this, we will take advantage of the free monad's structure.
+(Recall that the |AST| type we've been using is a synonym for the free monad.)
 The free monad is so called because as long as the parameter |f| is a |Functor|,
@@
 
-The following function evaluates the syntax tree.
+The following simple function evaluates the syntax tree.
 \begin{lstlisting}
@@
 \end{lstlisting}
-If our syntax tree already consists of just a single value (i.e. it is the |ePure| constructor),
+If our syntax tree already consists of just a single value (i.e. it is the |Pure| constructor),
 then we just return that value.
-Otherwise, we use the |Functor| instance to recursively convert the |Sig|s into types of |a|, then a call to |runSig| combines the results.
+Otherwise, we use the |Functor| instance to recursively convert the |Sig|s into types of |a|,
+combining the results with |runSig|.
 
@@
 \end{lstlisting}
-And this demonstrates that homogeneous functions are in fact homoiconic.
+We've finally demonstrated that homogeneous functions are in fact homoiconic.
 
@@
 
-\section{Implementing the stabilize function}
+\section{Writing program transformations}
 \label{sec:stabilize}
 
-We're now ready to modify our programs by manipulating the ASTs.
+We're now ready to write program transformations that manipulate the ASTs.
 The standard way to deconstruct an AST Haskell is via pattern matching.
@@
 
-Automatically stabilizing floating point expressions is an active area of research.
-One way of doing this is to create a library of program transformations that fix many different types of expressions.
-Then combining everything in the library on your problem.
-In this section, we'll create a very small library for these floating point issues,
-and don't claim to have completely solved the floating point problem.
-This is simply a motivating example for homoiconicity.
+%Automatically stabilizing floating point expressions is an active area of research.
+%One way of doing this is to create a library of program transformations that fix many different types of expressions.
+%Then combining everything in the library on your problem.
+%In this section, we'll create a very small library for these floating point issues,
+%and don't claim to have completely solved the floating point problem.
+%This is simply a motivating example for homoiconicity.
 
@@
 \end{lstlisting}
-As in our |logLogistic1| function, |testFunc| is unstable when specialized to |Double|s
-because input values greater than (approximately) 710 cause the |exp| function to overflow to infinity;
+As in our |logLogistic1| function, |testFunc| gives incorrect answers when specialized to |Double|.
+Inputs greater than (approximately) 710 cause the |exp| function to overflow to infinity;
 the log of infinity is still infinity;
 so the returned answer is infinity.
-But clearly this is not the correct answer.
-Mathematically, the $\log$ is the inverse of $\exp$, so the final result should be 710.
-We'll write a program transformation that fixes this function,
+But this wrong.
+Mathematically, the $\log$ function is defined to be the inverse of $\exp$,
+so the final result should be exactly the same as the input.
+We'll write a program transformation that stabilizes the |testFunc1| function,
 then we'll demonstrate how the |ViewPatterns| and |PatternSynonyms| language extensions simplify this transformation.
-
-Our goal is to convert the expression |log (exp x)| into just |x|.
-To do this, we can just pattern match on the AST for the left hand side of the expression,
-and replace it with the right hand side.
+\label{sec:viewpatterns}
+
+Our goal is to take in input AST,
+find all occurrences of the expression |log (exp x)|,
+and replace them with just |x|.
+Here |x| can be an arbitrary subtree of the original AST.
+We will create a series of four functions to accomplish this task,
+each improving on the previous.
+Here is the first.
 \begin{lstlisting}
@@
 \end{lstlisting}
-This function works, but has some warts.
-First, our pattern contains noise from the need to alternate between the |Free| and |Sig_| constructors.
-Second, it is specialized to the |Floating| type class,
-so it won't work on any subclasses of |Floating| when it should be able to.
-
-We'll start by fixing the second problem using the |ViewPatterns| language extension.
-We will provide a very brief description of this GHC extension here;
-for a more comprehensive description, see the GHC wiki \cite{viewpatterns}.
-|ViewPatterns| lets us use functions within a pattern,
-and we pattern match on the return value of the function.
-Because functions can be polymorphic,
-we can use the |ViewPatterns| extension to make our pattern accept multiple constructors.
-Fortunately, we've already defined a function that extracts a |Sig| from its subclasses.
+This function has two syntactic warts.
+First, the pattern is noisy.
+To properly match the expression, we must alternate between the |Free| and |Sig_| constructors.
+Second, the function is specialized to the |Floating| type class.
+We would like our transformation to also works on subclasses of |Floating|.
+
+We'll start by correcting the second problem using the |ViewPatterns| language extension.
+|ViewPatterns| provides a syntactic sugar that lets us pattern match on the return value of a function.\footnote{
+    The GHC wiki \cite{viewpatterns} provides details on exactly how the desugaring works.
+}
+We can use |ViewPatterns| to write polymorphic patterns by using polymorphic functions in the pattern.
+Fortunately, we've already defined the exact function we need:
 \begin{lstlisting}
@@
 \end{lstlisting}
-By using this function in the pattern, we can write a polymorphic version of our program transformation.
+By using this function in the pattern,
+we can write a polymorphic version of our program transformation.
 \begin{lstlisting}
@@
 but the pattern matches are even uglier (so problem 1 is worse).
-Fortunately, we can use the |PatternSynonyms| extension to simplify the patterns.
-Again, see the GHC wiki for a complete description of this language extension \cite{patternsynonyms}.
-
-We define a pattern synonym for each function contained in an FAlgebra.
+
+The |PatternSynonyms| extension lets us define new patterns,
+and we will use this feature to clean up our syntax.\footnote{
+    Again, see the GHC Wiki \cite{patternsynonyms} for details on this language extension.
+}
+Specifically, we define a pattern synonym for each function contained in an FAlgebra.
 Our naming convention is to prefix |AST_| to the name of the function.
@@
 Armed with these pattern synonyms,
-we can rewrite our function in the much simpler form below.
+we can rewrite our function in a much simpler form.
 \begin{lstlisting}
@@
 \end{lstlisting}
-This simple function works in the simple motivating test case:
+This simple function works on our the motivating test case:
 \begin{lstlisting}
@@
 We haven't told our program transformations what to do when they don't pattern match!
-The correct answer is that we want them to recurse into the subexpressions,
-applying their program transformations there as well.
-In the free monad, we handle these recursions using our |Functor| instance and calls to |fmap|.
-\begin{lstlisting}
-logexpAST4 :: View Floating alg => AST alg a -> AST alg a
+The correct answer is that we want the transformations to recurse into the subexpressions,
+applying the transformations there as well.
+In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the value of |Pure|.
+This gives our final version of the program transformation.
+\begin{lstlisting}
+logexpAST4 :: View Floating alg
+    => AST alg a -> AST alg a
 logexpAST4 (AST_log (AST_exp a)) = a
@@
 \begin{lstlisting}
-ghci> logexpAST4 $ testFunc2 var1 :: AST Floating Var
+ghci> logexpAST4 $ testFunc2 var1
+    :: AST Floating Var
 ((fromInteger 1)+var1)
@@
 What if we need to apply the program transformation multiple times?
+Consider the following test function.
+\begin{lstlisting}
+testFunc3 :: Floating a => a -> a
+testFunc3 a = 1+log(log(log(exp(exp(exp a)))))
+\end{lstlisting}
+Applying our program transformation directly doesn't fix the function because the transformation needs to be applied repeatedly.
 \begin{lstlisting}
@@
 %stopzone
-We want to apply the program transformation repeatedly until it stops changing the AST.
-This is called the fixed point of the transformation.
-The |fixAST| function below does this.
+What we want is the \emph{fixed point} of the program transformation.
+That is, we want to apply the program transformation repeatedly until it stops changing the AST.
+The |fixAST| function below takes as input a transformation and returns the fixed point of that transformation.
 \begin{lstlisting}
@@
 
-\subsection{More complex rules}
-
-The rules in in the previous section are relatively simple.
-They could just as easily have been written using GHC's rewrite rules syntax \cite{jones2001playing}.
+\subsection{More complex transformations}
+
+The transformation in in the previous section is relatively simple.
+It could just as easily have been written using GHC's rewrite rules syntax \cite{jones2001playing}.
 Rewrite rules are the primary mechanism for optimising source code in GHC.
-One optimization that rewrite rules cannot support is constant folding,
-and there is code in GHC specially to perform this constant folding.
-
-Our program transformations, however, are powerful enough to fold constants.
+One limitation of rewrite rules is that they cannot perform computations on the rewritten expression.
+This limitation prevents rewrite rules from implementing the constant folding optimization.
+Our program transformations, however, are powerful enough for this transformation.
+
 The following function performs constant folding on expressions containing addition and multiplication of integers.
@@
 constFunc :: Num a => a -> a -> a
-constFunc x1 x2 = x1*2-(7-2)*x2
-\end{lstlisting}
-Then, if we call this function with two known parameters,
-our |foldConstants| function is able to take it down to a single constant.
+constFunc x1 x2 = x1*2+(7+2)*x2
+\end{lstlisting}
+If we call this function with two known parameters,
+our |foldConstants| function is able to simplify the expression into a single constant.
 \begin{lstlisting}
@@
 \end{lstlisting}
-If we instead invoke the function with only
+Alternatively, we can call the function with two variable parameters and only fold the constants in the expression.
 \begin{lstlisting}
@@
 \end{lstlisting}
-We can even invoke the function with only a single known parameter,
-and have that constant get folded into the final expression.
+We can even invoke the function with one known and one variable parameter.
 \begin{lstlisting}
@@
 
-The next step in implementing logistic regression is implementing the logistic loss function
-\begin{align}
-l &: \mathbb{R}^n \times \mathbb{R}^n \to \mathbb{R} \\
-l &(v_1,v_2) = f(<v_1,v_2>)
-\end{align}
-This function takes two vectors as inputs and returns a scalar.
-Unfortunately, GHC's numeric hierarchy has no notion of vectors.
-So you implement a simple extension to the hierarchy shown in Figure \ref{code:vector}.
+GHC's numerical class hierarchy is notoriously limited.
+For example, GHC's numeric hierarchy has no notion of vectors.
+Figure \ref{code:vector} shows a simple extension to the default hierarchy to incorporate vectors.
 
@@
 
-class (Num a, Fractional (Scalar a))
+class (Num a, Floating (Scalar a))
     => Vector a where
@@
     dotproduct :: a -> a -> Scalar a
+
+data Vec3 = Vec3 a a a
+type Scalar (Vec3 a) = a
 \end{lstlisting}
@@
     A simple extension to GHC's numeric hierarchy for vector arithmetic.
+    A vector space is a special type of number---{\ttfamily (+)} represents vector addition, and {\ttfamily (*)} represents elementwise multiplication---that has an associated {\ttfamily Scalar} type and the ability to perform scalar multiplication with {\ttfamily (.*)}.
+    A Hilbert space is a vector space with dot products.
 }
@@
 So we cannot use the construction of Section \ref{sec:homogeneous} to generate an AST for this function.
-In this section, we'll introduce the more general notion of heterogeneous functions and show how to construct ASTs for them.
-
-Informally, the key difference between a homogeneous and heterogeneous FAlgebra is that homogeneous FAlgebras contain functions only over a single type; whereas heterogeneous FAlgebras contain functions over many different types.
-Formally, we call a function \emph{heterogeneous} if:
+In this section, we introduce the more general notion of heterogeneous type classes and functions.
+These concepts are called heterogeneous because they allow the construction of ASTs over multiple types through the use of type families.
+
+\begin{defn}
+We call a type class heterogeneous if:
 \begin{enumerate}
 \item
-The function's signature contains a single type variable, which we'll call |a|.
-\item
-The function returns either |a| or a type family applied to |a|.
-\item
-%The function's parameters are either the same as the return type, they are concrete.
-Each of the function's parameters is either:
+it has a single parameter of kind |Type|; and
+\item
+the class's constraints contain only heterogeneous type classes applied either directly to the parameter \emph{or to a type family applied to a parameter}.
+\end{enumerate}
+For example, every homogeneous type class is also heterogeneous.
+The type classes in Figure 2 are all heterogeneous, but not homogeneous.
+\end{defn}
+
+\begin{defn}
+We call a function heterogeneous if:
 \begin{enumerate}
 \item
+there is exactly one type variable in the signature, which we denote by |a|;
+\item
+each of the function's parameters is either
+\begin{enumerate}
+\item
 |a|,
 \item
-a type family applied to |a|, or
-\item
-concrete.
+concrete, or
+\emph{
+\item
+a type family applied to }|a|;
 \end{enumerate}
 \item
-The constraints contain exactly one type class, and
-that type class has exactly one parameter,
-which is the function's return type.
+the function's return type is either |a| \emph{or a type family applied to} |a|; and
+\item
+the constraints contain only heterogeneous type classes applied to |a| \emph{or type family applied to} |a|.
 \end{enumerate}
+All homogeneous functions are heterogeneous.
+The |logLoss| function is heterogeneous but not homogeneous.
+\end{defn}
+
 The construction of ASTs follows for heterogeneous types follows a similar pattern as for homogeneous types, but involves considerably more type hackery.
 
-\subsection{Dealing with type families}
-Before we're ready to construct the ASTs, we need a short diversion to discuss type families.
-Type families are functions at the type level in Haskell.
-Unfortunately, they're not as convenient to work with.
-In particular, type families cannot be partially applied.
-
-To simulate partial application, we will create a new data type for each type family.
-The naming convention is to put a |T| in front of the family's name.
-For example, we declare
-\begin{lstlisting}
-data TScalar
-\end{lstlisting}
-for the type family |Scalar|.
-\begin{lstlisting}
-type family App (t::k) (a::Type) :: Type
-\end{lstlisting}
-
-Repeated applications of type families will be stored in a promoted list.
-\begin{lstlisting}
-type instance App '[]       a = a
-type instance App (x ': xs) a = App x (App xs a)
-\end{lstlisting}
-
-Since |TScalar| is an ordinary Haskell type, it has kind |Type|.
-Ideally, we would like all of these shadow types to inhabit the same kind.
-But this would require open kind families,
-an extension that has been proposed\footnote{} but not implemented.
-To make our code easier to read, we will create the following kind synonym.
-\begin{lstlisting}
-type AT = Type
-\end{lstlisting}
-
-\subsection{The problems}
-
-There are three problems with our homogeneous FAlgebra encoding that make it unsuitable for heterogeneous FAlgebras.
+\subsection{The problems of heterogeneous types}
+Before discussing how to construct the heterogeneous ASTs,
+we will look at three problems that type families introduce to the homogeneous AST construction.
 
 \begin{problem}[the superclass problem]
-The |Vector| class has a superclass constraint on a type family.
+Consider the |Vector| class in Figure \ref{code:vector}.
+This class is not homogeneous because of the superclass constraint involving a type family application.
 If we follow the formulation for making a homogeneous FAlgebra, we get
@@
 the type checker will try to match the |a| type with |Scalar a| and fail.
+To fix this problem, we will need to modify the way we create superclass constructors.
 \end{problem}
@@
 \begin{problem}[the parameter problem]
-Now consider what happens when you try to write the |Functor| instance
+Now consider what happens when we write the |Functor| instance
 \begin{lstlisting}
@@
 \end{lstlisting}
-Here type checking fails on the last line for the same reason as above,
-but also fails on the first line due to the presence of a type family application in the function's parameter.
+Here type checking fails on |fmap|'s first clause in the expression |f s|.
+The function |f| has type |a -> b|, but the |s| variable has type |Scalar a|.
+We can no longer map arbitrary functions over our |Sig| type.
+To fix this problem, we will need a modified functor class.
 \end{problem}
@@
 \begin{problem}[the return type problem]
-Consider for a moment the |Hilbert| class in Figure \ref{code:vector}.
+Consider the |Hilbert| class in Figure \ref{code:vector}.
 The |dotProduct| function is not homogeneous because of the type family in the return type.
-What happens if we try to make |Hilbert| and instance of |FAlgebra| anyways?
-We would get something like:
+If we follow the formulation for making a homogeneous FAlgebra, we get
 \begin{lstlisting}
@@
 \end{lstlisting}
-This code doesn't type check.
-|runSig| expects its return type to be |a|,
+Here, type checking fails in the last clause of |runSig|.
+The return type should be |a|,
 but |dotProduct| returns a |Scalar a|.
+To fix this problem, we need to change the way we run a signature.
 \end{problem}
 
+\subsection{Dealing with type families}
+To address these problems, we first need to consider a limitation with how type families work.
+Consider the declaration
+\begin{lstlisting}
+newtype Wrap (t::Type->Type) a = Wrap (t a)
+\end{lstlisting}
+The list constructor |[]| has kind |Type->Type|,
+so we can create an expression like
+\begin{lstlisting}
+Wrap [1] :: Wrap [] Int
+\end{lstlisting}
+The |Scalar| type family also has kind |Type->Type|,
+so we would like to use it similarly.
+\begin{lstlisting}
+Wrap 1 :: Wrap Scalar (Vec3 Int)
+\end{lstlisting}
+Unfortunately, this expression does not type check because GHC requires that type families be fully applied anywhere they appear in a type signature.
+In this section, we will create a workaround that allows us to pass type families as arguments to types.\footnote{
+    The {\ttfamily singletons} package provides a similar construction in Section 4.3 of the related paper \cite{eisenbarg2015promoting}.
+    The {\ttfamily singletons}' construction is more general in order to allow for the promotion of ordinary functions to type families, and is therefore more complicated.
+    For ease of presentation, we do not use their construction.
+}
+
+The first step is to create a new data type for each type family.
+The naming convention used by \homoiconic is to put a |T| in front of the family's name.
+For example, we declare
+\begin{lstlisting}
+data TScalar
+\end{lstlisting}
+for the type family |Scalar|.
+We call types created in this way ``tags,''
+and create the following type synonym to refer to them.
+\begin{lstlisting}
+type Tag = Type
+\end{lstlisting}
+We would get more type safety if |Tag| were introduced as a new kind distinct from |Type|;
+but this requires open kinds, which is a feature not yet available in GHC.\footnote{
+    Issue \#11080 on GHC Trac has an extensive discussion of this feature.
+    See \url{https://ghc.haskell.org/trac/ghc/ticket/11080}.
+}
+
+Next, we need a way to apply the |TScalar| type as if it were a type family.
+The |App| type class serves this role.
+\begin{lstlisting}
+type family App (t::Tag) (a::Type) :: Type
+\end{lstlisting}
+The |t| parameter is the tag representing the type family we want to apply,
+and the |a| parameter is the type to which we want to apply the type family.
+The |Scalar| instance looks like
+\begin{lstlisting}
+type instance App TScalar a = Scalar a
+\end{lstlisting}
+And now we can create a modified version of our |Wrap| type above that actually works.
+\begin{lstlisting}
+newtype Wrap2 (t::Tag) a = Wrap2 (AppTag t a)
+\end{lstlisting}
+
+Frequently, we will want to deal with not just a single type family application,
+but a whole sequence of type family applications.
+We can represent these applications with kind |[Tag]| as in the following family.
+\begin{lstlisting}
+type family AppTags (t::[Tag]) (a::Type) :: Type
+type instance AppTags '[]       a = a
+type instance AppTags (x ': xs) a
+    = AppTag x (AppTags xs a)
+\end{lstlisting}
+Now we can create an even more powerful wrapper that can handle arbitrary applications of type families.
+\begin{lstlisting}
+newtype Wrap3 (t::[Tag]) a = Wrap3 (AppTags t a)
+\end{lstlisting}
+
 \subsection{Heterogeneous FAlgebras}
 
-Our new |FAlgebra| class has a number of modifications to solve the problems mentioned above.
-It is defined as:
+At last we are ready to see our type class for heterogeneous FAlgebras.
 \begin{lstlisting}
 class HetFunctor (Sig alg) => FAlgebra alg where
-    data Sig alg (t::[AT]) a
-
-    runSig    :: alg a => proxy a
-        -> Sig alg        t  (App t a) -> App        t  a
-    runSigTag :: alg a => proxy a
-        -> Sig alg (s ':  t) (App t a) -> App (s ':  t) a
-\end{lstlisting}
-
-The most important change is that the |Sig| data family has been given a new type parameter |t|.
-This parameter indexes the return type of the operation stored in |Sig|.
-Notice that |t| has kind |[AT]|.
-If there is no type family application, then |t| will be the empty list |'[]|;
-if there is a single application by data constructor |A|, |t| will be |'[A]|;
-and if the return type is |A (B a)|, then |t| will be |'[A,B]|.
-
-When defining instances of the |Sig| family, we will need to use GADTs to enforce that the |t| parameter is set correctly.
-|Sig|'s constructors can be classified into the same categories as before.
-For superclass constructors, if the class constraint does not involve a type family,
-then we get defined as:
+    data Sig alg (t::[Tag]) a
+
+    runSig0 :: alg a => proxy a
+        -> Sig alg t (AppTags t a)
+        -> AppTags t a
+
+    runSig1 :: alg a => proxy a
+        -> Sig alg (s ': t) (AppTags t a)
+        -> AppTags (s ': t) a
+\end{lstlisting}
+This class is considerably more complex looking than for homogeneous FAlgebras,
+but it follows the same basic principles.
+The most important difference is that the |Sig| data family has been given a new type parameter |t|.
+This parameter tells us which type families have been applied to the variable |a|.
+Exactly what this means will become more clear as we see how to create instances of |FAlgebra|.
+
+First we discuss how to define instances of the |Sig| family.
+We will need to use GADTs \cite{schrijvers2009complete} to ensure the |t| parameter is set correctly.
+As before, the constructors can be classified as either function constructors or superclass constructors.
+For function constructors, the parameters remain exactly the parameters of the function they correspond to.
+The return type, however, depends on the return type of the function.
+If the function's return type does not use a type family, then then the |t| parameter is |'[]|.
+For example, the function constructor for |Vector|'s |(.*)| is
+\begin{lstlisting}
+Sig_dotmul :: Scalar a -> a -> Sig Vector '[] a
+\end{lstlisting}
+If the function's return type does use a type family, then the |t| parameter is the singleton list containing the family's tag.
+For example, the function constructor for |Hilbert|'s |dotProduct| is
+\begin{lstlisting}
+Sig_dotProduct :: a -> a -> Sig Hilbert '[TScalar] a
+\end{lstlisting}
+Unlike functor constructors, superclass constructors are parametric in their |t| parameter.
+Superclass constructors take a single parameter,
+which is the |Sig| of the superclass.
+If the corresponding constraint does not involve a type family,
+then the |t| parameter is the same for the superclass and the returned value.
+For example, the |Num a| constraint in the |Vector| instance gets the constructor
 \begin{lstlisting}
@@
 \end{lstlisting}
-if the class constraint does involve a type family,
-then we prepend the type family to |t| in the constructor's return type.
+If the class constraint does involve a type family,
+then we prepend the type family's tag to |t| in the constructor's return type.
+For example, the |Floating a| constraint in the |Vector| instance gets the constructor
 \begin{lstlisting}
@@
 \end{lstlisting}
-Function constructors are likewise divided into two cases.
-If the function's return type does not use a type family,
-then
-\begin{lstlisting}
-Sig_dotmul :: Scalar a -> a -> Sig Vector '[] a
-\end{lstlisting}
-If the function's return type does use a type family,
-then
-\begin{lstlisting}
-Sig_dotProduct :: a -> a -> Sig Hilbert '[TScalar] a
-\end{lstlisting}
-These changes to the |Sig| type are the core of the new addition.
+These changes to the |Sig| type are the core of the differences between heterogeneous and homogeneous |FAlgebra|s.
 There will be many changes to all of the other functions,
-but they all stem from adapting the same techniques used in the homogeneous case to the updated |Sig|.
-
-The |runSig| function has been split into two functions, |runSig| and |runSigTag|.
-First we'll explain the |runSig| function, then see why its deficiencies necessitate the |runSigTag| function.
-Recall that the |t| type variable represents the type families that have been applied to the content of the |Sig|.
+but they all stem from adapting the same techniques used in the homogeneous case to the modified |Sig|.
+
+The |runSig| function has been split into two functions:
+|runSig0| is for running functions that do not involve type families,
+and |runSig1| is for running functions that involve a single type family.
+|runSig0| works just like the homogeneous |runSig|,
+but has a more complicated type signature due to the addition of the |t| parameter to |Sig|.
+Recall that the |t| type variable represents the type families that have already been applied to the content of the |Sig|.
 So when we run the |Sig|, the content needs to reflect those applications.
-Thus, the parameter to |runSig| is |Sig alg t (App t a)|.
-Similarly, the return type needs to reflect those applications plus whatever applications are present in the function.
-And so the return type is |Sig alg t (App t a)|.
+Thus, the parameter to |runSig0| is |Sig alg t (App t a)|.
+Similarly, the return type needs to reflect those applications,
+and so the return type is |Sig alg t (App t a)|.
 Because the |a| type variable appears only inside type family applications,
-we need the |proxy a| argument for type resolution.
-The |runSig| function is defined exactly as before.
-\begin{lstlisting}
-runSig _ (Sig_Vector_Num s)
-    = runSig (Proxy::Proxy a) s
-runSig _ (Sig_Vector_Floating s)
-    = runSig (Proxy::Proxy (Scalar a)) s
-runSig _ (Sig_dotmul s a) = s.*a
-\end{lstlisting}
-If the function's return type involves a type family application,
-then it is impossible to use the |runSig| function on it.
-The signature is wrong.
-That is why we introduce a second function |runSigTag|.
-\begin{lstlisting}
-runSigTag p (Sig_Hilbert_Vector s) = runSigTag p s
-runSigTag p (Sig_dotproduct a1 a2) = dotproduct a1 a2
-\end{lstlisting}
+we need the |proxy a| argument to avoid an ambiguous type.
+To implement the |runSig0| function, function constructors get |runSig0| called recursively (with an appropriate |Proxy| parameter representing the type family application in the constraint);
+type 0 function constructors are evaluated exactly as in the homogeneous case,
+and type 1 function constructors are not evaluated at all since there's no way to make them type check.
+For example, here is |Vector|'s |runSig0| function
+\begin{lstlisting}
+runSig0 _ (Sig_Vector_Num s)
+    = runSig0 (Proxy::Proxy a) s
+runSig0 _ (Sig_Vector_Floating s)
+    = runSig0 (Proxy::Proxy (Scalar a)) s
+runSig0 _ (Sig_dotmul s a) = s.*a
+\end{lstlisting}
+and |Hilbert|'s |runSig0| function
+\begin{lstlisting}
+runSig0 _ (Sig_Hilbert_Vector s)
+    = runSig0 (Proxy::Proxy a) s
+\end{lstlisting}
+The return type of |runSig1| contains an extra |Tag| applied to it that the content of |Sig| does not have.
+This lets us evaluate type 1 functions.
+To implement |runSig1|, function constructors get called recursively as before,
+but this time for only type 1 functions can be evaluated,
+and type 2 functions will be ignored.
+Here are the |runSig1| functions for |Vector| and |Hilbert|.
+\begin{lstlisting}
+runSig1 _ (Sig_Vector_Num s)
+    = runSig0 (Proxy::Proxy a) s
+runSig1 _ (Sig_Vector_Floating s)
+    = runSig0 (Proxy::Proxy (Scalar a)) s
+
+runSig1 _ (Sig_Hilbert_Vector s)
+    = runSig1 (Proxy::Proxy a) s
+runSig1 _ (Sig_dotproduct a1 a2) = dotproduct a1 a2
+\end{lstlisting}
+The last component of the |FAlgebra| is the |HetFunctor| class.
+This class is used for evaluating the |AST|s and is discussed in Section \ref{sec:heteval}.
+Before we get to that,
+we'll discuss how to construct and show the ASTs.
 
@@
 
-To construct an AST out of the |Sig| functor requires a more complex version of the free monad.
-\begin{lstlisting}
-data Free (f::[AT]->Type->Type) (t::[AT]) (a::Type) where
-    FreeTag  :: f (s ': t) (Free f t a)  -> Free f (s ': t) a
-    Free     :: f       t  (Free f t a)  -> Free f       t  a
-    Pure     :: App t a -> Free f t a
+To construct an AST out of the |Sig| functor requires a heterogeneous version of the free monad.
+Like the |Sig| type, the heterogeneous free monad will carefully track the |t| parameter and so must be a GADT.
+\begin{lstlisting}
+data Free (f::[Tag]->Type->Type) (t::[Tag]) a where
+  Free1::f (s ':t)(Free f t a)->Free f (s ':t) a
+  Free0::f      t (Free f t a)->Free f      t  a
+  Pure ::App t a -> Free f t a
 \end{lstlisting}
@@
 To convert the heterogeneous AST into a |String|,
-we need to write a |Show| instance for |Free| and a show instance for each |Sig|.
-Both of these instances are considerably more complicated than before.
+we need a |Show| instance for each |Sig| and a |Show| instance for |Free|.
 
@@
 For each superclass constructor, show the contained constructor.
+|Vector|'s show instance is given below.
 \begin{lstlisting}
@@
 \end{lstlisting}
-We had to add the |Show (Scalar a)| constraint above because the |a1| variable in the last line has type |Scalar a|.
+Notice that we had to add the |Show (Scalar a)| constraint above because the |a1| variable in the last line has type |Scalar a|.
 
@@
 \end{lstlisting}
-This means that the constraint |Show (Scalar a)| will use the same |Show| instance as the one displayed above.
-In order to satisfy that constraint, we need |Show (Show (Scalar a))|.
-This nesting of |Show| constraints continues forever!
-In order to stop this explosion of |Show| constraints, we add the following overlapping instance.
+This means that when we call |show| on an |AST|,
+the constraint solver will try to satisfy the |Show (Scalar a)| with the exact same instance.
+In order to satisfy that constraint, we will need |Show (Show (Scalar a))|;
+and to satisfy that, we need |Show (Scalar (Scalar (Scalar a)))|.
+This nesting of |Scalar| type families continues forever,
+and the constraint solver loops!
+
+One solution to this problem is to add the following overlapping instance.
 \begin{lstlisting}
@@
 \end{lstlisting}
-Once there are at least three |Scalar|s applies,
-this instance becomes more specific,
+Once there are at least three |Scalar|s applied to |Show|'s parameter,
+this new instance becomes more specific,
 and is thus selected by GHC.
@@
 \subsection{Evaluating the AST}
+\label{sec:heteval}
 
@@
 eval (Pure    a) = a
-eval (Free    s) = runSig    (Proxy::Proxy a) $ mape eval s
-eval (FreeTag s) = runSigTag (Proxy::Proxy a) $ mape eval s
+eval (Free    s) = runSig0 (Proxy::Proxy a) $ mape eval s
+eval (FreeTag s) = runSig1 (Proxy::Proxy a) $ mape eval s
 \end{lstlisting}

### 2016-06-09 src/Heterogeneous/ExamplePrelude.hs

@@ -131,7 +131,7 @@ instance Floating a => Vector (Vec3 a) where
 --------------------------------------------------------------------------------
 
 logexpAST1 :: AST Floating '[] a -> AST Floating '[] a
-logexpAST1 (Free (Sig_log (Free (Sig_exp a)))) = a
+logexpAST1 (Free0 (Sig_log (Free0 (Sig_exp a)))) = a
 
 -- logexpAST2 :: forall t alg a. View Floating t alg '[] => AST alg '[] a -> AST alg '[] a
 -- logexpAST2 (Free (unsafeExtractSig -> Sig_log a)) = a
@@ -160,7 +160,7 @@ logexpAST4 (AST_log (AST_exp a)) = a
 
 logexpAST5 :: View Floating '[] alg t => AST alg t a -> AST alg t a
 logexpAST5 (AST_log (AST_exp a)) = a
-logexpAST5 (Free f) = Free $ fmapAST logexpAST4 f
+logexpAST5 (Free0 f) = Free0 $ fmapAST logexpAST4 f
 
 fmapAST :: (forall s. View Floating '[] alg s => AST alg s a -> AST alg s a) -> Sig alg t (AST alg t a) -> Sig alg t (AST alg t a)
 fmapAST f = undefined

### 2016-06-09 src/Heterogeneous/FAlgebra.hs

@@ -38,13 +38,15 @@ import Unsafe.Coerce
 
 --------------------------------------------------------------------------------
 
-type AT = Type
+type Tag = Type
 
-type family App (t::k) (a::Type) ::  Type
-type instance App '[]           a = a
-type instance App ((x::AT)':xs) a = App x (App xs a)
+type family AppTag (t::Tag) (a::Type)
 
-type family TypeConstraints (t::[AT]) (a::Type) :: Constraint
+type family AppTags (t::[Tag]) (a::Type) :: Type where
+    AppTags '[]       a = a
+    AppTags (x ': xs) a = AppTag x (AppTags xs a)
+
+type family TypeConstraints (t::[Tag]) (a::Type) :: Constraint
 
 
 type family Snoc (xs :: [k]) (y::k) where
@@ -68,28 +70,28 @@ appCoerce
     :: Proxy t
     -> Proxy s
     -> Proxy a
-    -> App t (App s a)
-    -> App (t `Snoc` s) a
+    -> AppTags t (AppTag s a)
+    -> AppTags (t `Snoc` s) a
 appCoerce _ _ _ = unsafeCoerce
 
 sigCoerce
     :: Proxy t
     -> Proxy s
     -> Proxy a
-    -> Sig alg t (App (t `Snoc` s) a)
-    -> Sig alg t (App  t (App s a))
+    -> Sig alg t (AppTags (t `Snoc` s) a)
+    -> Sig alg t (AppTags  t (AppTag s a))
 sigCoerce _ _ _ = unsafeCoerce
 
-runSigSnoc :: forall proxy alg a t u.
+runSig0Snoc :: forall proxy alg a t u.
     ( FAlgebra alg
-    , alg (App u a)
+    , alg (AppTag u a)
     ) => Proxy u
 --       -> Proxy u
       -> Proxy a
-      -> Sig alg t (App (t `Snoc` u) a)
-      -> App (t `Snoc` u) a
-runSigSnoc ps pa u
-    = appCoerce pt ps pa $ runSig (Proxy::Proxy (App u a)) $ sigCoerce pt ps pa u
+      -> Sig alg t (AppTags (t `Snoc` u) a)
+      -> AppTags (t `Snoc` u) a
+runSig0Snoc ps pa u
+    = appCoerce pt ps pa $ runSig0 (Proxy::Proxy (AppTag u a)) $ sigCoerce pt ps pa u
     where
[…]

### 2016-06-09 src/Homogeneous/FAlgebra.hs

@@ -92,7 +92,7 @@ import Test.Tasty
 import Test.Tasty.Ingredients.Basic
 import Test.Tasty.Options
 import Test.Tasty.Runners
-import qualified Test.Tasty.SmallCheck as SC
+-- import qualified Test.Tasty.SmallCheck as SC
 import qualified Test.Tasty.QuickCheck as QC
 import Test.QuickCheck hiding (Testable)
 

### 2016-06-09 src/Homogeneous/PreludeInstances.hs

@@ -201,6 +201,14 @@ constFunc x1 x2 = x1*2+(7+2)*x2
 
 --------------------------------------------------------------------------------
 
+instance
+    ( FAlgebra alg
+    , Show (Sig alg (Free (Sig alg) Var))
+    ) => Show (AST alg Var -> AST alg Var) where
+    show f = show (f var1)
+
+--------------------------------------------------------------------------------
+
 -- type family Scalar a
 -- -- mkAT ''Scalar
 --

### 2016-06-09 src/TemplateHaskellUtils.hs

@@ -16,7 +16,7 @@ import Test.Tasty
 import Test.Tasty.Ingredients.Basic
 import Test.Tasty.Options
 import Test.Tasty.Runners
-import qualified Test.Tasty.SmallCheck as SC
+-- import qualified Test.Tasty.SmallCheck as SC
 import qualified Test.Tasty.QuickCheck as QC
 import Test.QuickCheck hiding (Testable)
 

### 2016-06-10 paper/paper.tex
more paper edits

@@
 \usepackage{amsmath,amsfonts}
-\usepackage{url}
+\usepackage{hyperref}
 
@@
 \begin{filecontents}{paper.bib}
+@article{birkhoff1970heterogeneous,
+  title={Heterogeneous algebras},
+  author={Birkhoff, Garrett and Lipson, John D},
+  journal={Journal of Combinatorial Theory},
+  volume={8},
+  number={1},
+  pages={115--133},
+  year={1970},
+  publisher={Elsevier}
+}
+
 @article{goldberg1991,
@@
 
+@inproceedings{panchekha2015automatically,
+  title={Automatically improving accuracy for floating point expressions},
+  author={Panchekha, Pavel and Sanchez-Stern, Alex and Wilcox, James R and Tatlock, Zachary},
+  booktitle={Proceedings of the 36th ACM SIGPLAN Conference on Programming Language Design and Implementation},
+  pages={1--11},
+  year={2015},
+  organization={ACM}
+}
 @misc{viewpatterns,
@@
 \title{Homoiconicity Without Template Haskell}
-\subtitle{Functional Pearl}
+%\subtitle{Functional Pearl}
 
@@
 Template Haskell gives a \emph{compile time} form of homoiconicity to Haskell programs;
-but in this pearl, we demonstrate another form of homoiconicity that relies on polymorphism and GHC's type programming capabilities.
+but in this paper, we demonstrate another form of homoiconicity that relies on polymorphism and GHC's type programming capabilities.
 To motivate our technique,
@@
 
-%This pearl shows that many Haskell functions are \emph{homoiconic}.
+%This paper shows that many Haskell functions are \emph{homoiconic}.
 %That is, we can convert these functions into isomorphic abstract syntax trees (ASTs),
@@
 %seemingly correct mathematical formulae can give wildly inaccurate results.
-%This pearl shows that Haskell's type system can mitigate this danger.
+%This paper shows that Haskell's type system can mitigate this danger.
 %
@@
 
-In this pearl, we see how to write functions like |stabilize| that perform program transformations.
+In this paper, we see how to write functions like |stabilize| that perform program transformations.
 These functions follow the following pattern.
@@
 
-\section{Homogeneous functions are homoiconic}
-\label{sec:homogeneous}
-
-In this section we first define homogeneous type classes and functions.
 There are two intuitions for homogeneous type classes.
@@
 Second, we can construct an FAlgebra for these type classes.
+
+\section{Homogeneous functions are homoiconic}
+\label{sec:homogeneous}
+
+In this section we first define homogeneous type classes and functions.
+Then we show that there is an FAlgebra for each homogeneous type class,
+and that this FAlgebra makes homogeneous functions homoiconic.
 
@@
 \item
-the constraints contain only homogeneous type classes applied to |a|.
+the constraints contain exactly one homogeneous type classes applied to |a| and nothing else.
 \end{enumerate}
@@
 \end{defn}
+
+%In the remainder of this section we describe a canonical method for constructing FAlgebras from homogeneous type classes and show that this construction makes homogeneous functions homoiconic.
+%The construction is tedious to implement by hand,
+%so the \homoiconic library provides the function |mkFAlgebra| to automate the boilerplate.
+%This function is provided as a convenience and is not strictly necessary for the technique.
 
@@
 this constructor should have the same parameters as the function.
-For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three member functions |(/)|, |recip|, and |fromRational| respectively.
+For the |Fractional| example, we've defined the three constructors |Sig_div|, |Sig_recip|, and |Sig_fromRational| for the three class methods |(/)|, |recip|, and |fromRational| respectively.
 The second constructor flavor is the \emph{superclass constructor}.
@@
 \subsection{Constructing the AST}
+\label{sec:hom.cons}
 
@@
 Using our uniform representation of an AST,
-we can make the |AST| type an instance of the homogeneous type classes.
+we can make the |AST| type an instance of any homogeneous type class.
 As an example, the |Fractional| instance is shown below.
@@
 
-We can easily create ASTs using these instances.
+These instances let us easily create ASTs.
 The |Num| AST can be created with the expression
@@
 \begin{lstlisting}
-expr5 = AST Floating Double
+expr5 :: AST Floating Double
 expr5 = (1+2)+3
@@
 \end{lstlisting}
-It would be nice if the displayed function would display the actual input variables,
-rather than just substituting numbers.
+It would be nice if the displayed function could show variables,
+rather than just numbers.
 To accomplish this, we need a variable type.
@@
 Values of type |Var| cannot be plugged directly into the |logLogistic1| function because |Var| is not an instance of |Floating|.
-Fortunately, we have a way to make any type an instance of |Floating| via the |AST Floating| type.
+Fortunately, we have a way to make any type an instance of |Floating| via |Floating|'s |AST| type.
 More generically, we define the following variables.
@@
 GHC's defaulting mechanism is not powerful enough for it to properly specialize the function without a type signature.
-So unfortunately this technique doesn't give us a convenient |Show| instance,
-it does give us a slightly more verbose alternative for showing a function.
+So while this construction does give us a convenient way to inspect the contents of a function,
+it's not quite powerful enough for a proper |Show| instance.
 
@@
 
+\subsection{Cheating with template Haskell}
+
+For each homogeneous type class,
+there are many boilerplate instances that we need to write.
+The \homoiconic library provides a template Haskell function |mkFAlgebra| in the |Homogeneous.FAlgebra| module that generates the |FAlgebra|, |Functor|, |Show|, and |View| instances plus all of the pattern synonyms.
+
 %%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
@@
 
-We're now ready to write program transformations that manipulate the ASTs.
-The standard way to deconstruct an AST Haskell is via pattern matching.
-In this section, we'll first demonstrate how to pattern match our ASTs using the |ViewPatterns| and |PatternSynonyms| language extensions.
-Then we'll see three simple examples of program transformations implemented in this way.
+We're now ready to write program transformations that manipulate the ASTs created in the previous section.
+In this section, we'll first demonstrate how to pattern match on our ASTs using the |ViewPatterns| and |PatternSynonyms| language extensions.
+%Then we'll see three simple examples of program transformations implemented in this way.
+We'll also see three motivating examples of useful transformations on floating point operations.
 
@@
 \end{lstlisting}
-As in our |logLogistic1| function, |testFunc| gives incorrect answers when specialized to |Double|.
+As in our |logLogistic1| function,
+|testFunc| gives incorrect answers when |a| is specialized to |Double|.
 Inputs greater than (approximately) 710 cause the |exp| function to overflow to infinity;
@@
 But this wrong.
-Mathematically, the $\log$ function is defined to be the inverse of $\exp$,
+On the real numbers, the $\log$ function is defined to be the inverse of $\exp$,
 so the final result should be exactly the same as the input.
@@
 
-Our goal is to take in input AST,
+Our goal is to take an input AST,
 find all occurrences of the expression |log (exp x)|,
 and replace them with just |x|.
-Here |x| can be an arbitrary subtree of the original AST.
+(Here |x| can be an arbitrary AST, not just a single value.)
 We will create a series of four functions to accomplish this task,
@@
 This function has two syntactic warts.
-First, the pattern is noisy.
+First, the pattern is verbose.
 To properly match the expression, we must alternate between the |Free| and |Sig_| constructors.
+It would be better to have a single constructor per operation.
 Second, the function is specialized to the |Floating| type class.
@@
 We can use |ViewPatterns| to write polymorphic patterns by using polymorphic functions in the pattern.
-Fortunately, we've already defined the exact function we need:
+We need a polymorphic function for extracting a |Sig alg a| type from a |Sig Floating a| whenever |alg| is a subclass of |Floating|.
+Fortunately, we've already defined such a function:
 \begin{lstlisting}
@@
 \end{lstlisting}
-By using this function in the pattern,
-we can write a polymorphic version of our program transformation.
+Using it in the pattern gives us
 \begin{lstlisting}
@@
 \end{lstlisting}
-This function now has the correct type signature (so we've solved problem 2),
-but the pattern matches are even uglier (so problem 1 is worse).
-
-The |PatternSynonyms| extension lets us define new patterns,
-and we will use this feature to clean up our syntax.\footnote{
+This modified program transformation now has the correct polymorphic type signature
+(so we've solved problem 2),
+but the pattern matches are even uglier
+(so we've made problem 1 worse).
+
+The |PatternSynonyms| extension lets us define new patterns,\footnote{
     Again, see the GHC Wiki \cite{patternsynonyms} for details on this language extension.
 }
+and we will use this feature to clean up our syntax.
 Specifically, we define a pattern synonym for each function contained in an FAlgebra.
@@
 %stopzone
-but it fails to work in this slightly more complex function:
+but it fails on this slightly more complex function:
 \begin{lstlisting}
@@
 \end{lstlisting}
-We haven't told our program transformations what to do when they don't pattern match!
-The correct answer is that we want the transformations to recurse into the subexpressions,
+The problem is that we haven't told our program transformations what to do when they don't pattern match!
+When pattern matching fails, we want our transformations to recurse into the subexpressions,
 applying the transformations there as well.
-In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the value of |Pure|.
+In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the contents of |Pure|.
 This gives our final version of the program transformation.
@@
 %stopzone
-There's still one last wrinkle.
+But there's still one last wrinkle.
 What if we need to apply the program transformation multiple times?
@@
 \end{lstlisting}
-Applying our program transformation directly doesn't fix the function because the transformation needs to be applied repeatedly.
+Applying our program transformation directly doesn't correct the function because the transformation needs to be applied repeatedly.
 \begin{lstlisting}
@@
 %stopzone
+And that's all there is to writing a program transformation.
+More complex transformations just involve more complex pattern matching.
 
@@
 The transformation in in the previous section is relatively simple.
-It could just as easily have been written using GHC's rewrite rules syntax \cite{jones2001playing}.
-Rewrite rules are the primary mechanism for optimising source code in GHC.
-One limitation of rewrite rules is that they cannot perform computations on the rewritten expression.
-This limitation prevents rewrite rules from implementing the constant folding optimization.
-Our program transformations, however, are powerful enough for this transformation.
+For example, it could have been implemented using GHC's rewrite rules \cite{jones2001playing}.
+Rewrite rules are the primary mechanism for optimising source code in GHC,
+but they are limited to transformations that involve no computation.
+This limitation prevents rewrite rules from implementing a number of optimizations,
+one of them being being constant folding.
+Constant folding is the process of reducing an expression containing several constants into a single constant.
+For example, converting the expression |(1+2)+3| into just the number |6|.
+%(Due to this limitation, GHC has special ad-hoc code just for constant folding.)
+Our program transformations, however, can implement arbitrary computation and so are powerful enough to handle constant folding.
 
@@
 
-Here's an example of our constant folding in action.
-First define a simple function.
-\begin{lstlisting}
-constFunc :: Num a => a -> a -> a
-constFunc x1 x2 = x1*2+(7+2)*x2
-\end{lstlisting}
-If we call this function with two known parameters,
-our |foldConstants| function is able to simplify the expression into a single constant.
-\begin{lstlisting}
-ghci> fixAST foldConstants (constFunc 2 3)
+Let's see an example of constant folding in action.
+First define a simple function with two parameters a handful of constants.
+\begin{lstlisting}
+func :: Num a => a -> a -> a
+func x1 x2 = x1*2+(7+2)*x2
+\end{lstlisting}
+If we call |func| with two constant parameters,
+our |foldConstants| function is able to simplify the resulting expression into a single constant.
+\begin{lstlisting}
+ghci> fixAST foldConstants (func 2 3)
     :: AST Floating Double
@@
 \end{lstlisting}
-Alternatively, we can call the function with two variable parameters and only fold the constants in the expression.
-\begin{lstlisting}
-ghci> fixAST foldConstants (constFunc var1 var2)
+Alternatively, we can call the function with two variable parameters.
+Only the constants in the original function will be folded.
+\begin{lstlisting}
+ghci> fixAST foldConstants (func var1 var2)
     :: AST Floating Var
@@
 \end{lstlisting}
-We can even invoke the function with one known and one variable parameter.
-\begin{lstlisting}
-ghci> fixAST foldConstants (constFunc var1 3)
+And finally, we can call the function with one constant and one variable parameter.
+\begin{lstlisting}
+ghci> fixAST foldConstants (func var1 3)
     :: AST Floating Var
@@
 
-Our original motivation was to stabilize the |logLogistic1| function.
-Stabilizing this function is a bit harder because it actually requires changing the type signature of the function by adding an |Ord| instance.
-One of the requirements of a homogeneous function is that it have only a single class in the context.
-To accommodate this condition, we'll create a new type class that has our desired types in the constraint.
+Now let's see how to stabilize the |logLogistic1| function.
+Stabilizing this function is a bit harder because it requires changing the type signature of the function---the constraints on |logLogistic1| require only a |Floating| instance, but the constraints on |logLogistic2| require both a |Floating| and |Ord| instance.
+Recall that one of the requirements for a function to be homogeneous is that it have only a single class in the context.
+Therefore the stabilized function |logLogistic2| is not homogeneous.
+
+We can make an equivalent implementation, however, that is homogeneous by introducing the following type class
 \begin{lstlisting}
 class (Floating a, Ord a) => FloatingOrd a
-instance {-#OVERLAPPABLE#-} (Floating a, Ord a) => FloatingOrd a
-\end{lstlisting}
-Note that the |OVERLAPPABLE| pragma is required so that we can create an |Ord| instance for |AST FloatingOrd a|.
-
-Now we can define our transformation as
-\begin{lstlisting}
-stabilize :: Eq a =>
-    AST FloatingOrd a -> AST FloatingOrd a
+\end{lstlisting}
+and the following instance
+\begin{lstlisting}
+instance {-#OVERLAPPABLE#-} (Floating a, Ord a)
+    => FloatingOrd a
+\end{lstlisting}
+Note that the |OVERLAPPABLE| pragma is required so that we can create an |Ord| instance for |AST FloatingOrd a| per the construction in Section \ref{sec:hom.cons}.
+
+We can now write a function that stabilizes |logLogistic1|
+\begin{lstlisting}
+stabilize :: AST Floating a -> AST FloatingOrd a
 stabilize
@@
 \end{lstlisting}
+Notice that in this transformation, the type signature of the AST actually changes.
+
+This transformation works, but it feels like cheating.
+If we have to manually write a unique program transformation for every numeric function,
+then we're not saving any work.
+Fortunately, we don't have to do that.
+
+\subsection{Arbitrary floating point stability}
+
+Creating program transformations that stabilize floating point numbers is a hard,
+actively researched problem.
+Recently, a team from the University of Washington created a tool called Herbie that uses a number of clever techniques to stabilize these floating point expressions \cite{panchekha2015automatically}.
+We can use programs like herbie within our Haskell programs
+
+\begin{lstlisting}
+herbie :: AST Floating a -> IO (AST FloatingOrd a)
+\end{lstlisting}
 
@@
 
-GHC's numerical class hierarchy is notoriously limited.
-For example, GHC's numeric hierarchy has no notion of vectors.
-Figure \ref{code:vector} shows a simple extension to the default hierarchy to incorporate vectors.
+In the 1970's, mathematicians noticed a deficiency in the definition of FAlgebras.
+The signature of an FAlgebra can only contain a single type,
+but many structures involve multiple types.
+For example, the definition of a vector space requires both a scalar type and a vector type.
+The concept of heterogeneous FAlgebras was introduced to formalize these algebraic structures over multiple types \cite{birkhoff1970heterogeneous}.
+
+%The same pattern happens in Haskell.
+Similarly in Haskell, homogeneous type classes are not sufficient to describe all the structures we want to work with.
+For example, Figure \ref{code:vector} shows a simple extension to GHC's numerics hierarchy for working with vectors.
+The |Vector| and |Hilbert| type classes are not homogeneous because they rely on a type family to describe the relationship between multiple types.
+Using this class hierarchy, we can implement functions like the logistic loss:
+\begin{lstlisting}
+logLoss :: Vector a => a -> a -> Scalar a
+logLoss a1 a2 = logLogistic1 (dotproduct a1 a2)
+\end{lstlisting}
+And again, this function is not homogeneous due to the type family in the signature.
+Nonetheless, we would still like to be able to convert this function into an AST as we did for homogeneous functions.
+
+In this section we introduce the notion of heterogeneous type classes and functions.
+We will see that for every heterogeneous type class,
+there exists a corresponding heterogeneous FAlgebra.
+From this heterogeneous FAlgebra,
+we can construct an AST using the heterogeneous free monad.
+This construction gives us homoiconicity for heterogeneous functions.
+Unfortunately, type families as implemented in GHC Haskell have many limitations,
+and the limitations will complicate the heterogeneous constructions.
+We will require considerably more type hackery in this section.
 
@@
 
-You can now implement the logistic loss as:
-\begin{lstlisting}
-logLoss :: Vector a => a -> a -> Scalar a
-logLoss a1 a2 = logLogistic1 (dotproduct a1 a2)
-\end{lstlisting}
-Unfortunately, this function is not homogeneous due to the application of a type family in the return type.
-So we cannot use the construction of Section \ref{sec:homogeneous} to generate an AST for this function.
-In this section, we introduce the more general notion of heterogeneous type classes and functions.
-These concepts are called heterogeneous because they allow the construction of ASTs over multiple types through the use of type families.
+%Unfortunately, this function is not homogeneous due to the application of a type family in the return type.
+%So we cannot use the construction of Section \ref{sec:homogeneous} to generate an AST for this function.
+%In this section, we introduce the more general notion of heterogeneous type classes and functions.
+%These concepts are called heterogeneous because they allow the construction of ASTs over multiple types through the use of type families.
 
@@
 The type classes in Figure 2 are all heterogeneous, but not homogeneous.
+%Multiparameter type classes are not heterogeneous.
 \end{defn}
@@
 The |logLoss| function is heterogeneous but not homogeneous.
+
+Heterogeneous functions come in two types.
+Type 0 functions have no type family applications in the return type.
+For example, |(.*)| is a type 0 function.
+Type 1 functions have one type family application in the return type.
+For example, |dotProduct| is a type 1 function.
 \end{defn}
 
-The construction of ASTs follows for heterogeneous types follows a similar pattern as for homogeneous types, but involves considerably more type hackery.
-
 \subsection{The problems of heterogeneous types}
-Before discussing how to construct the heterogeneous ASTs,
+\label{sec:het.probs}
+Before describing how to construct the heterogeneous ASTs,
 we will look at three problems that type families introduce to the homogeneous AST construction.
@@
         | Sig_Vector_Num (Sig Num a)
-        | Sig_Vector_Floating (Sig Floating (Scalar a))
+        | Sig_Vector_Floating
+            (Sig Floating (Scalar a))
     runSig (Sig_dotmul s a) = s.* a
@@
 This definition does not type check.
-In the last line of |runSig|'s definition,
-the type checker will try to match the |a| type with |Scalar a| and fail.
+In the last clause of |runSig|'s definition,
+the type checker will try to match the |a| type with |Scalar a| (because |s| has type |Sig Floating (Scalar a)| instead of |Sig Floating a|) and fail.
 To fix this problem, we will need to modify the way we create superclass constructors.
@@
     fmap f (Sig_Vector_Num s) = Sig_Vector_Num (f s)
-    fmap f (Sig_Vector_Floating s) = Sig_Vector_Floating (f s)
-\end{lstlisting}
-Here type checking fails on |fmap|'s first clause in the expression |f s|.
-The function |f| has type |a -> b|, but the |s| variable has type |Scalar a|.
-We can no longer map arbitrary functions over our |Sig| type.
-To fix this problem, we will need a modified functor class.
+    fmap f (Sig_Vector_Floating s)
+        = Sig_Vector_Floating (f s)
+\end{lstlisting}
+Here type checking fails on |fmap|'s first clause in the expression |(f s)|.
+The function |f| has type |a->b|, but the |s| variable has type |Scalar a|.
+To fix this problem, we will need a modified |fmap| function.
 \end{problem}
@@
 Here, type checking fails in the last clause of |runSig|.
-The return type should be |a|,
+The return type of |runSig| should be |a|,
 but |dotProduct| returns a |Scalar a|.
@@
 \begin{lstlisting}
-newtype Wrap (t::Type->Type) a = Wrap (t a)
+newtype Wrap1 (t::Type->Type) a = Wrap1 (t a)
 \end{lstlisting}
 The list constructor |[]| has kind |Type->Type|,
-so we can create an expression like
-\begin{lstlisting}
-Wrap [1] :: Wrap [] Int
+so the following expression type checks.
+\begin{lstlisting}
+Wrap1 [1] :: Wrap1 [] Int
 \end{lstlisting}
@@
 \begin{lstlisting}
-Wrap 1 :: Wrap Scalar (Vec3 Int)
+Wrap1 1 :: Wrap1 Scalar (Vec3 Int)
 \end{lstlisting}
@@
 In this section, we will create a workaround that allows us to pass type families as arguments to types.\footnote{
-    The {\ttfamily singletons} package provides a similar construction in Section 4.3 of the related paper \cite{eisenbarg2015promoting}.
-    The {\ttfamily singletons}' construction is more general in order to allow for the promotion of ordinary functions to type families, and is therefore more complicated.
+    The {\ttfamily singletons} package provides a similar construction in Section 4.3 of the packages paper \cite{eisenberg2015promoting}.
+    %The {\ttfamily singletons}' construction is more general than the one presented here.
+    Their construction allows for the promotion of ordinary functions to type families,
+    and is therefore more complicated.
     For ease of presentation, we do not use their construction.
@@
 
-Next, we need a way to apply the |TScalar| type as if it were a type family.
-The |App| type class serves this role.
-\begin{lstlisting}
-type family App (t::Tag) (a::Type) :: Type
+Next, we need a way to apply tags to a type as if they were a type family.
+The |AppTag| type family serves this role.
+\begin{lstlisting}
+type family AppTag (t::Tag) (a::Type) :: Type
 \end{lstlisting}
@@
 \begin{lstlisting}
-type instance App TScalar a = Scalar a
-\end{lstlisting}
-And now we can create a modified version of our |Wrap| type above that actually works.
+type instance AppTag TScalar a = Scalar a
+\end{lstlisting}
+And now we can create a modified version of our |Wrap1| type above that actually works.
 \begin{lstlisting}
@@
 but a whole sequence of type family applications.
-We can represent these applications with kind |[Tag]| as in the following family.
+We can represent these applications with kind |[Tag]|.
+The following type family applies a sequence of tags to a type.
 \begin{lstlisting}
@@
 \end{lstlisting}
+For example, the type
+|AppTags '[TScalar,TScalar] a|
+is equivalent to the type
+|Scalar (Scalar a)|,
+and
+|AppTags '[] a| is equivalent to |a|.
+%\begin{lstlisting}
+%AppTags '[TScalar,TScalar] a
+%\end{lstlisting}
+%is equivalent to the type
+%\begin{lstlisting}
+%Scalar (Scalar a)
+%\end{lstlisting}
+
 Now we can create an even more powerful wrapper that can handle arbitrary applications of type families.
@@
 \begin{lstlisting}
-class HetFunctor (Sig alg) => FAlgebra alg where
+class FAlgebra alg where
   data Sig alg (t::[Tag]) a
@@
     -> AppTags (s ': t) a
-\end{lstlisting}
-This class is considerably more complex looking than for homogeneous FAlgebras,
-but it follows the same basic principles.
-The most important difference is that the |Sig| data family has been given a new type parameter |t|.
+
+  mapRun
+    :: (forall s. Free (Sig alg') s a -> AppTags s a)
+    -> Sig alg t (Free (Sig alg') t' a)
+    -> Sig alg t (AppTags t' a)
+\end{lstlisting}
+Don't panic!
+Understanding exactly how this class works in not necessary for using the ASTs it generates.
+Still, this section will carefully explain how it works.
+
+The most important difference between the homogeneous and heterogeneous |FAlgebra| class is that the |Sig| data family has been given a new type parameter |t::[Tag]|.
 This parameter tells us which type families have been applied to the variable |a|.
@@
 We will need to use GADTs \cite{schrijvers2009complete} to ensure the |t| parameter is set correctly.
-As before, the constructors can be classified as either function constructors or superclass constructors.
-For function constructors, the parameters remain exactly the parameters of the function they correspond to.
+As before, the constructors come in two flavors.
+For function constructors,
+the parameters remain exactly the parameters of the function they correspond to.
 The return type, however, depends on the return type of the function.
-If the function's return type does not use a type family, then then the |t| parameter is |'[]|.
-For example, the function constructor for |Vector|'s |(.*)| is
+For type zero functions (there are no type family applications in the return type),
+then |t| parameter is set to |'[]|.
+For example, the function constructor for |Vector|'s |(.*)| function is
 \begin{lstlisting}
@@
 \end{lstlisting}
-If the function's return type does use a type family, then the |t| parameter is the singleton list containing the family's tag.
-For example, the function constructor for |Hilbert|'s |dotProduct| is
+For type one functions (there is one type family application in the return type),
+the |t| parameter is the singleton list containing the family's tag.
+For example, the function constructor for |Hilbert|'s |dotProduct| function is
 \begin{lstlisting}
@@
 then we prepend the type family's tag to |t| in the constructor's return type.
-For example, the |Floating a| constraint in the |Vector| instance gets the constructor
+For example, the |Floating (Scalar a)| constraint in the |Vector| instance gets the constructor
 \begin{lstlisting}
@@
 These changes to the |Sig| type are the core of the differences between heterogeneous and homogeneous |FAlgebra|s.
-There will be many changes to all of the other functions,
-but they all stem from adapting the same techniques used in the homogeneous case to the modified |Sig|.
-
-The |runSig| function has been split into two functions:
-|runSig0| is for running functions that do not involve type families,
-and |runSig1| is for running functions that involve a single type family.
+There will be many changes to all of the other constructions,
+but they are all a result of these changes to the signature.
+
+The |runSig| function tells us how to evaluate the signature of an FAlgebra.
+It has been split into two functions:
+|runSig0| is for running type 0 functions (no type family in the return type),
+and |runSig1| is for running functions type 1 functions (type family in the return type).
 |runSig0| works just like the homogeneous |runSig|,
 but has a more complicated type signature due to the addition of the |t| parameter to |Sig|.
-Recall that the |t| type variable represents the type families that have already been applied to the content of the |Sig|.
-So when we run the |Sig|, the content needs to reflect those applications.
-Thus, the parameter to |runSig0| is |Sig alg t (App t a)|.
-Similarly, the return type needs to reflect those applications,
-and so the return type is |Sig alg t (App t a)|.
+Recall that the |t| type variable represents the type families that have already been applied to the signature;
+so when we run the |Sig|, the result needs to reflect those applications.
+Thus, the parameter to |runSig0| is |Sig alg t (AppTag t a)|,
+and the return type is |AppTag t a|.
 Because the |a| type variable appears only inside type family applications,
 we need the |proxy a| argument to avoid an ambiguous type.
-To implement the |runSig0| function, function constructors get |runSig0| called recursively (with an appropriate |Proxy| parameter representing the type family application in the constraint);
+To implement the |runSig0| function, function constructors recursively call |runSig0| on their argument (with an appropriate |Proxy| parameter representing the type family application in the constraint);
 type 0 function constructors are evaluated exactly as in the homogeneous case,
-and type 1 function constructors are not evaluated at all since there's no way to make them type check.
+and type 1 function constructors are not evaluated at all because there is no way to make them type check.
 For example, here is |Vector|'s |runSig0| function
@@
 This lets us evaluate type 1 functions.
-To implement |runSig1|, function constructors get called recursively as before,
+To implement |runSig1|, superclass constructors get called recursively as before,
 but this time for only type 1 functions can be evaluated,
-and type 2 functions will be ignored.
+and type 0 functions will be ignored.
 Here are the |runSig1| functions for |Vector| and |Hilbert|.
@@
 \end{lstlisting}
-The last component of the |FAlgebra| is the |HetFunctor| class.
-This class is used for evaluating the |AST|s and is discussed in Section \ref{sec:heteval}.
+
+The last component of the |FAlgebra| is the |mapRun| function.
+Since our new |Sig| type cannot be made an instance of |Functor|
+(see problem 2 in Section \ref{sec:het.probs}),
+this function performs the role that |fmap| performed in the homogeneous FAlgebras.
+That is, it will allow recursion over the AST.
+We will discuss the function in more detail in Section \ref{sec:heteval} when we discuss evaluating ASTs.
 Before we get to that,
@@
 To construct an AST out of the |Sig| functor requires a heterogeneous version of the free monad.
-Like the |Sig| type, the heterogeneous free monad will carefully track the |t| parameter and so must be a GADT.
+It is defined as
 \begin{lstlisting}
@@
   Free0::f      t (Free f t a)->Free f      t  a
-  Pure ::App t a -> Free f t a
-\end{lstlisting}
-There are now three constructors.
-The |Free| and |Pure| constructors work exactly as they did before.
-The only difference is that |Pure| does not store the |a| type inside of it,
-but instead it applies the |t| type families to |a|.
-The |FreeTag| constructor represents a branch like the |Free| constructor,
-but it also represents a change in return type.
-When embedding a |Sig| into an AST,
-we will use the |FreeTag| constructor whenever the function's return type uses a type family.
-
-The AST is then defined as
+  Pure ::AppTag t a -> Free f t a
+\end{lstlisting}
+and the corresponding AST is
 \begin{lstlisting}
@@
 \end{lstlisting}
+There are two major changes in the heterogeneous free monad.
+First, the kind of the |Free| type constructor has changed.
+We've added an additional parameter |t::[Tag]| that represents the type families that need to get applied to |a| to evaluate the syntax tree.
+The type signature of the |runAST| function (discussed in detail in Section \ref{sec:heteval}) makes this clear.
+\begin{lstlisting}
+runAST :: (FAlgebra alg, alg a)
+    => Free (Sig alg) t a -> AppTag t a
+\end{lstlisting}
+Notice that the return type applies |t| to |a|.
+Changing the type of |t| changes the type we get when evaluating the AST.
+
+The second change is that there are now three constructors instead of two.
+The |Pure| constructor again represents leaves in the AST.
+These leaves now have type |AppTag t a| instead of just |a| to accommodate the modified return type of |runAST|.
+The |Free0| and |Free1| constructors represent branches in the AST.
+These branches correspond to type 0 and type 1 functions respectively.
+Notice that in type 1 branches, all of the subtrees have a different type than the tree itself.
+
+An example should illustrate the relationship between the three constructors.
+Consider the expression
+\begin{lstlisting}
+exp (dotProduct (Vec3 1 2 3) (Vec3 2 3 4))
+\end{lstlisting}
+|Vec3| is an instance of |Hilbert|, so we can create a corresponding AST using |Hilbert|'s |FAlgebra| instance.
+\begin{lstlisting}
+expr1 :: AST Hilbert '[TScalar] (Vec3 Double)
+expr1 = Free0
+  ( Sig_Hilbert_Vector
+    ( Sig_Vector_Floating
+      ( Sig_exp
+        ( Free1
+          ( Sig_dotProduct
+            ( Pure (Vec3 1 2 3)
+              :: AST Hilbert '[] (Vec3 Double)
+            )
+            ( Pure (Vec3 2 3 4)
+              :: AST Hilbert '[] (Vec3 Double)
+            )
+          )
+          :: Sig Hilbert '[TScalar]
+            (AST Hilbert '[] (Vec3 Double))
+        )
+        :: AST Hilbert '[TScalar] (Vec3 Double)
+      )
+      :: Sig Hilbert '[TScalar]
+        (AST Hilbert '[TScalar] (Vec3 Double))
+    )
+    :: Sig Hilbert '[TScalar]
+      (AST Hilbert '[TScalar] (Vec3 Double))
+  )
+\end{lstlisting}
+The type signatures on each expression are labeled to emphasize how the type of the tree changes with the application of the |Free0| and |Free1| constructors.
+Specifically, because the dot product is a type 1 function (its return type is |Scalar a|),
+the |Sig_dotProduct| function gets embedded into the AST using the |Free1| constructor.
+This changes the type of the tree by adding the |TScalar| tag.
+When we evaluate the resulting tree, we will get a scalar,
+so it is now safe to perform scalar operations.
+The |exp| function is an example of a scalar operation.
+The design of the |Sig_Vector_Floating| constructor lets us embed the |Sig_exp| constructor into the tree only because of the |TScalar| tag generated by |Sig_dotProduct|.
+Since |Sig_exp| is a type 0 function,
+it is embedded into the tree using the |Free0| constructor.
 
@@
 \end{lstlisting}
-
+Instances of |View| should satisfy the property that
+
+\begin{lstlisting}
+type Scalar (AST alg t a) = AST alg (TScalar ': t) a
+\end{lstlisting}
+
+\begin{lstlisting}
+instance ()
+    => Vector (AST alg t a) where
+    (.*) s a = Free0 $ embedSig $ Sig_dotmul s a
+\end{lstlisting}
+
+\begin{lstlisting}
+instance ()
+    => Hilbert (AST alg t a) where
+    dotProduct a1 a2
+        = Free1 $ embedSig $ Sig_dotProduct a1 a2
+\end{lstlisting}
+
+%\begin{lstlisting}
+%ghci> exp (dotProduct var1 var2) :: AST Hilbert '[TScalar] Var
+%(exp (dotProduct var1 var2))
+%\end{lstlisting}
 %\begin{lstlisting}
@@
 
-The |Show| instance for the |Sig| type follows the same pattern as before.
+The |Show| instance for the |Sig| type follows the same pattern as the homogeneous case.
 For each function constructor,
 simply show the constructor's arguments connected in an appropriate way by the function's name.
-For each superclass constructor, show the contained constructor.
-|Vector|'s show instance is given below.
+For each superclass constructor, recursively call show the superclass.
+As an example, |Vector|'s show instance is given below.
 \begin{lstlisting}
@@
 Notice that we had to add the |Show (Scalar a)| constraint above because the |a1| variable in the last line has type |Scalar a|.
-
-This instance can be defined, but it can't yet be used.
+In general, whenever a function constructor contains a parameter with a type class,
+we will need to add a |Show| constraint for the corresponding class.
+
+This |Show| instance can be defined, but it can't yet be used.
 Recall that we defined
@@
 This means that when we call |show| on an |AST|,
-the constraint solver will try to satisfy the |Show (Scalar a)| with the exact same instance.
+the constraint solver will try to satisfy the |Show (Scalar a)| constraint with the exact same instance.
 In order to satisfy that constraint, we will need |Show (Show (Scalar a))|;
@@
 
-We must perform a related type hack to get the |Show| instance for |Free|.
-The |Show| instance for |Free| is shown below.
+We need another type hack to get the |Show| instance for |Free|.
+%The difficulty here arises due to trying to show the contents of the |Free1| constructor.
+%Below is a |Show| instance ignoring this constructor.
+If we ignore the |Free1| constructor, then the |Show| instance is the exactly the same as for the homogeneous free monad.
 \begin{lstlisting}
 instance
-    ( Show      (App t a)
+    ( Show (AppTag t a)
+    , Show (f t (Free f t a))
+    ) => Show (Free f t a) where
+    show (Pure  a ) = show a
+    show (Free0 f0) = "("++show f0++")"
+    show (Free1 f1) = undefined
+\end{lstlisting}
+If we try to show |f1| the same way we show |f0|,
+we run into type problems.
+
+A solution is to create a closed type family to apply a |Show| constraint only when it makes sense.
+\begin{lstlisting}
+type family ShowUntag (f::Type) :: Constraint where
+    ShowUntag (f (s ': t) (Free f (s ': t) a))
+        = Show (f (s ': t) (Free f t a))
+    ShowUntag a = ()
+\end{lstlisting}
+Now we can write the complete |Show| instance as
+\begin{lstlisting}
+instance
+    ( Show      (AppTag t a)
     , Show      (f t (Free f t a))
     , ShowUntag (f t (Free f t a))
-    ) => Show (Free f t a)
-        where
-    show (FreeTag     f) = "("++show f++")"
-    show (Free        f) = "("++show f++")"
+    ) => Show (Free f t a) where
     show (Pure  a) = show a
-
-type family ShowUntag (f::Type) :: Constraint where
-    ShowUntag (f (s ':  t) (Free f (s ':  t) a))  = Show (f (s ':  t) (Free f          t  a))
-    ShowUntag a = ()
-\end{lstlisting}
-The |FreeTag| constructor must be undressed in order to satisfy the |Show| requirement.
+    show (Free0 f) = "("++show f++")"
+    show (Free1 f) = "("++show f++")"
+\end{lstlisting}
+
+Now that these |Show| instances are written,
+actually showing expressions is straightforward.
+You just have to add the appropriate type signature to the expression.
+\begin{lstlisting}
+ghci> logLogistic1 3 :: AST Floating '[] Double
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate (fromInteger 3))))))
+\end{lstlisting}
+
+In order to display functions,
+we create the |Var| type exactly as before.
+\begin{lstlisting}
+newtype Var a = Var String
+\end{lstlisting}
+Now when we create values of type |Var|,
+we want them to be usable in expressions of any tagging.
+This means they need to be parametric in both the |alg| and |t| parameters.
+\begin{lstlisting}
+var1 :: AppTags t Var~Var => AST alg t Var
+var1 = Pure $ Var "var1"
+
+var2 :: AppTags t Var~Var => AST alg t Var
+var2 = Pure $ Var "var2"
+
+var3 :: AppTags t Var~Var => AST alg t Var
+var3 = Pure $ Var "var3"
+\end{lstlisting}
+%stopzone
+To do this, we need to enforce that any type family applied to |Var| is equivalent to |Var|.
+
+Now we can show our ASTs like before.
+The only difference is that we must now include the |t| parameter as well.
+\begin{lstlisting}
+ghci> logLogistic1 var1 :: AST Floating '[] Var
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate var1)))))
+\end{lstlisting}
+Recall that |Floating| appeared as a constraint for the |Vector| class.
+So you might think we could specialize the expression above to the |Vector| AST.
+But if you try, you get a type error.
+\begin{lstlisting}
+ghci> logLogistic1 var1 :: AST Vector '[] Var
+No instance for (View Floating ('[] Tag) Vector ('[] Tag))
+      arising from a use of ‘logLogistic1’
+\end{lstlisting}
+Just because something implements |Vector| does not guarantee that it implements |Floating| as well.
+It only guarantees that its |Scalar| implements |Floating|.
+
+\begin{lstlisting}
+ghci> logLogistic1 1 :: AST Vector '[TScalar] Var
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate (fromInteger 1))))))
+\end{lstlisting}
 
@@
 
-Running the heterogeneous AST is almost the same as the homogeneous case.
-\begin{lstlisting}
-eval :: forall alg t a.
-    ( FAlgebra alg
-    , alg a
-    ) => Free (Sig alg) t a -> App t a
-eval (Pure    a) = a
-eval (Free    s) = runSig0 (Proxy::Proxy a) $ mape eval s
-eval (FreeTag s) = runSig1 (Proxy::Proxy a) $ mape eval s
-\end{lstlisting}
-The only difference is we need to add a line for the |FreeTag| parameter,
-and we need to manually specify the type |a| with the |Proxy|.
-
-\subsection{Heterogeneous homoiconicity at last!}
+Running the heterogeneous AST is similar to running the homogeneous AST with three exceptions.
+First, the type signature is different as we've already discussed.
+Second, since the free monad has an additional constructor,
+the |runAST| function will need an additional clause pattern matching against this constructor.
+Finally, we must use the |mapAST| function instead of |fmap| for recursion.
+The full implementation is shown below.
+\begin{lstlisting}
+runAST :: forall alg t a. (FAlgebra alg, alg a)
+    => Free (Sig alg) t a -> AppTag t a
+runAST (Pure  a) = a
+runAST (Free0 s)
+    = runSig0 (Proxy::Proxy a) $ mapAST runAST s
+runAST (Free1 s)
+    = runSig1 (Proxy::Proxy a) $ mapAST runAST s
+\end{lstlisting}
 

### 2016-06-10 src/Heterogeneous/ExamplePrelude.hs

@@ -62,6 +62,7 @@ mkFAlgebra ''FloatingOrd
 type family Scalar a
 mkAT ''Scalar
 
+type instance Scalar Double = Double
 type instance Scalar Var = Var
 
 scalar1 :: AST alg '[TScalar] Var
@@ -89,6 +90,9 @@ class (Num a, Floating (Scalar a)) => Vector a where
 --     show (Sig_Vector_Num s) = show s
 --     show (Sig_Vector_Floating_Scalar s) = show s
 --     show (Sig_dotmul s a) = show s++".*"++show a
+
+-- instance {-#OVERLAPS#-} Show (Sig Vector (t0 ': t1 ': t) a) where
+--     show _ = "<<overlaps>>"
 
 mkFAlgebra ''Vector
 
@@ -127,6 +131,9 @@ instance Num a => Num (Vec3 a) where
 
 instance Floating a => Vector (Vec3 a) where
     (.*) s (Vec3 a1 a2 a3) = Vec3 (s*a1) (s*a2) (s*a3)
+
+instance Floating a => Hilbert (Vec3 a) where
+    dotProduct (Vec3 a1 a2 a3) (Vec3 b1 b2 b3) = a1*b1+a2*b2+a3*b3
 
 --------------------------------------------------------------------------------
 
@@ -189,3 +196,6 @@ logLogistic2 :: FloatingOrd x => x -> x
 logLogistic2 x = m+log(1/(exp(m)+exp(-x+m)))
     where
         m = min 0 x
+
+logLoss :: Hilbert x => x -> x -> Scalar x
+logLoss x1 x2 = logLogistic1 (dotProduct x1 x2)

### 2016-06-10 src/Heterogeneous/FAlgebra.hs

@@ -146,6 +146,16 @@ data Free (f::[Tag]->Type->Type) (t::[Tag]) (a::Type) where
     Free0 :: TypeConstraints t a => f       t  (Free f t a)  -> Free f       t  a
     Pure  :: AppTags t a -> Free f t a
 
+-- instance
+--     ( Show (AppTag s (AppTags t a))
+--     , Show (f (s ': t) (Free f (s ': t) a))
+--     , Show (f (s ': t) (Free f       t  a))
+--     ) => Show (Free f (s ': t) a)
+--         where
+--     show (Free1 f) = "("++show f++")"
+--     show (Free0 f) = "("++show f++")"
+--     show (Pure  a) = show a
+
 instance
     ( Show      (AppTags t a)
     , Show      (f t (Free f t a))
@@ -160,13 +170,13 @@ type family ShowUntag (f::Type) :: Constraint where
     ShowUntag (f (s ':  t) (Free f (s ':  t) a))  = Show (f (s ':  t) (Free f          t  a))
     ShowUntag a = ()
 
-eval :: forall alg t a.
+runAST :: forall alg t a.
     ( FAlgebra alg
     , alg a
     ) => Free (Sig alg) t a -> AppTags t a
-eval (Pure    a) = a
-eval (Free0   s) = runSig0    (Proxy::Proxy a) $ mape eval s
-eval (Free1 s) = runSig1 (Proxy::Proxy a) $ mape eval s
+runAST (Pure  a) = a
+runAST (Free0 s) = runSig0 (Proxy::Proxy a) $ mapRun runAST s
+runAST (Free1 s) = runSig1 (Proxy::Proxy a) $ mapRun runAST s
 
 class NoCxt (a::k)
 instance NoCxt a
@@ -271,7 +281,7 @@ class Typeable alg => FAlgebra (alg::Type->Constraint) where
     runSig1 :: alg a => proxy a -> Sig alg (s ':  t) (AppTags t a) -> AppTags (s ':  t) a
     runSig0    :: alg a => proxy a -> Sig alg        t  (AppTags t a) -> AppTags        t  a
 
-    mape :: (TypeConstraints t' a)
+    mapRun :: (TypeConstraints t' a)
          => (forall s. Free (Sig alg') s a -> AppTags s a)
          -> Sig alg t (Free (Sig alg') t' a)
          -> Sig alg t (AppTags t' a)
[…]

### 2016-06-12 paper/paper.tex
final edits

@@
 \newtheorem{defn}{Definition}
+\newtheorem{note}{Note}
 
@@
              {<=}{{$\leq$}}2 {=>}{{$\Rightarrow$}}2
+             {~}{\tiny{$\sim$}}1
     %keywordstyle=\lst@ifdisplaystyle\color{blue}\fi,
@@
 
+@inproceedings{sheard2002template,
+  title={Template meta-programming for Haskell},
+  author={Sheard, Tim and Jones, Simon Peyton},
+  booktitle={Proceedings of the 2002 ACM SIGPLAN workshop on Haskell},
+  pages={1--16},
+  year={2002},
+  organization={ACM}
+}
+
 @article{swierstra2008,
@@
 %\title{The (floating) point of homoiconicity}
-\title{Homoiconicity Without Template Haskell}
+%\title{Homoiconicity Without Template Haskell}
+\title{Homoicoicity with FAlgebras}
 %\subtitle{Functional Pearl}
@@
 This AST can then be manipulated arbitrarily and converted back into a function.
-Template Haskell gives a \emph{compile time} form of homoiconicity to Haskell programs;
-but in this paper, we demonstrate another form of homoiconicity that relies on polymorphism and GHC's type programming capabilities.
-To motivate our technique,
-we construct a small library that automatically transforms numerically unstable code into stable code.
+Template Haskell gives a limited form of homoiconicity to Haskell programs,
+but this paper introduces another form.
+The technique relies on an isomorphism between certain Haskell classes and FAlgebras.
+The \homoiconic library provides tools for using the technique.
+%To make the technique more widely applicable,
+%we introduce the Haskell community to the heterogeneous FAlgebra
+%polymorphism and GHC's type programming capabilities.
+%We also present the \homoiconic library that facilitates working with this technique.
+%To motivate our technique,
+%we construct a small library that automatically transforms numerically unstable code into stable code.
 
@@
 
-%Imagine you've just downloaded and installed a machine learning library for Haskell.
-%You open up ghci and import the corresponding modules.
-%In the documentation, you see a function called |logLogistic|,
-%so you ask ghci its type.
-%\begin{lstlisting}
-%ghci> :t logLogistic
-%logLogistic :: Floating a => a -> a
-%\end{lstlisting}
-%Now imagine you want to know how the function is implemented.
-%\begin{lstlisting}
-%ghci> logLogistic var1 :: AST Floating Var
-%\end{lstlisting}
-%Oh no!
-%The |logLogistic| function has a numerical instability in it.
-
-Imagine you're writing a machine learning library in Haskell.
-At some point, you'll need to implement the log-logistic function
-\begin{equation}
-f(x) = \log\left(\frac{1}{1+\exp{(-x)}}\right)
-\label{eq:log-logistic}
-\end{equation}
-This function is used in logistic regression and many families of neural networks.
-As a first attempt, you might write:
-\begin{lstlisting}
-logLogistic1 :: Floating x => x -> x
-logLogistic1 x = log(1/(1+exp(-x)))
-\end{lstlisting}
-Unfortunately, this straightforward implementation is numerically unstable.
+Template Haskell \cite{sheard2002template} is the standard tool for metaprogramming in Haskell.
+%Using quotations, we can convert Haskell code into an abstract syntax tree (AST),
+%and using splices, we can insert arbitrary ASTs into our code.
+%Template Haskell is enormously useful, but it also has limitations.
+In this paper, we introduce a different style of metaprogramming and an accompanying library called \homoiconic.
+
+For example, imagine you're using a library that exports a function
+\begin{lstlisting}
+logLogistic :: Floating a => a -> a
+\end{lstlisting}
+There is no way to use Template Haskell to get the abstract syntax tree (AST) of this function.
+With \homoiconic, however, we \emph{can} inspect the |logLogistic| function and get its AST.
+A ghci session to do this might look like:
+\begin{lstlisting}
+ghci> logLogistic var1 :: AST Floating Var
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate var1)))))
+\end{lstlisting}
+In this paper, we'll call any function homoiconic if we can convert it into an AST.
+So the |logLogistic| function above is homoiconic.
+
+The \homoiconic library also lets us perform arbitrary program transformations.
+A careful look at the output above will reveal that |logLogistic| is numerically unstable.
 When specialized to the |Double| type,
@@
 returning negative infinity when the true answer is actually very close to $-710$.
-Such an inaccurate answer is unacceptable,
-so the function must be rewritten.
-
-But like all good programmers, you're lazy.
-Your machine learning library will contain hundreds of numeric functions,
-all of which will need to be carefully written to avoid numeric instabilities.
-What a maintenance nightmare!
-Soon you start daydreaming about a function
-\begin{lstlisting}
-stabilize :: Floating a => (a -> a) -> (a -> a)
-\end{lstlisting}
-that takes a naively written floating point function as input and outputs a numerically stable version.
-Armed with such a function, you could write a stable implementation of the log-logistic function as
-\begin{lstlisting}
-logLogistic2 :: Floating x => x -> x
-logLogistic2 x = stabilize $ log(1/(1+exp(-x)))
-\end{lstlisting}
-%stopzone
-
-In this paper, we see how to write functions like |stabilize| that perform program transformations.
-These functions follow the following pattern.
-First, convert the input function into an abstract syntax tree (AST).
-Next, manipulate the AST to remove the numerical instabilities.
-Finally, convert the AST back into a callable function.
-%The key observation is that certain Haskell functions are \emph{homoiconic}.
-%That is, we can convert them into isomorphic abstract syntax trees (ASTs).
-%The key observation is that certain Haskell functions (which we call \emph{homogeneous} and \emph{heterogeneous}) can be converted into equivalent abstract syntax trees (ASTs).
-%The |stabilize| function converts its input function into an AST,
-%then manipulate that AST to remove numerical stabilities,
-%and finally convert the AST back into a callable function.
-
-The main idea of the |homoiconic| library is that we can associate with every type class an FAlgebra that captures the algebraic structure of that class.
-
-In Section \ref{sec:homogeneous},
-%In Section \ref{sec:homogeneous}, we introduce the idea of a \emph{homogeneous} function and show that these functions are homoiconic.
-%In Section \ref{sec:stabilize}, we use the construction above to implement several versions of the |stabilize| function.
-%In Section \ref{sec:heterogeneous}, we introduce the idea of \emph{heterogeneous} functions and show that they are also homoiconic.
-%Finally, we conclude by introducing a Haskell package called {\ttfamily homoiconic} that makes it easy for you to .
-
-There are two intuitions for homogeneous type classes.
-First, they have ASTs that contain only a single type variable.
-Second, we can construct an FAlgebra for these type classes.
+Using \homoiconic, we can write a function
+\begin{lstlisting}
+stabilize :: AST Floating a -> AST Floating a
+\end{lstlisting}
+that converts the AST of an unstable function into a stable function.
+We could write a similar transformation in Template Haskell,
+but there would be no way for us to get the AST to apply it on!
+
+The key idea of the \homoiconic library is simple:
+Certain type classes have FAlgebras that completely capture their structure.
+From FAlgebras, we can create ASTs.
+And we can use these ASTs to represent expressions generated by the corresponding type classes.
+%It's important to note that \homoiconic is much more limited than Template Haskell in other ways.
+%In particular, it will only work on certain (but not all!) expressions.
+%\footnote{
+%FAlgebras are a standard Haskell design pattern for domain specific languages.
+%There are many existing references, but Data Types \`a la Carte \cite{swierstra2008} is probably the most famous.
+%}
+%From these FAlgebras, we can in turn generate ASTs for functions.
+
+The remainder of this paper describes these constructions.
+Section \ref{sec:homogeneous} introduces \emph{homogeneous} type classes and functions.
+Homogeneous type classes have a particularly simple FAlgebra construction that is a natural extension of the Data Types \`a la Carte approach \cite{swierstra2008}.
+Section \ref{sec:stabilize} illustrates how to write program transformations like |stabilize| with a handful of examples operating on numerical functions.
+Finally, Section 4 introduces \emph{heterogeneous} type classes.
+The naming stems from the intuition that homogeneous classes generate ASTs that contain a single type,
+but heterogeneous classes generate ASTs that contain multiple types.
+The resulting constructions for the heterogeneous case is significantly more complicated.
 
@@
 and that this FAlgebra makes homogeneous functions homoiconic.
+All of the constructions discussed here can be found in \homoiconic's |Homogeneous.FAlgebra| module.
 
@@
 \end{enumerate}
-The |logLogistic1| function is homogeneous,
-and Figure \ref{code:ghc} shows many more examples taken from the Prelude's numeric hierarchy.
+The |logLogistic| function is homogeneous,
+and Figure \ref{code:ghc} shows many more examples and counterexamples taken from the |Prelude|'s numeric hierarchy.
 \end{defn}
@@
 \begin{lstlisting}
-instance View Fractional alg
-    => Fractional (AST alg a) where
+instance
+    ( View Fractional alg
+    , View Num alg
+    ) => Fractional (AST alg a) where
     (/) e1 e2 = Free $ embedSig $ Sig_div e1 e2
@@
 we simply embed the function's constructor into the AST by prepending the expression |Free $ embedSig $|.
+The constrainst |View num alg| is not used directly in the instance.
+It is inherrited from the |Num (AST alg a)| instance,
+where it is used directly.
+In general, a type class instance for |AST| will require a |View| constraint for every superclass,
+not just those parent classes listed in the context.
 
@@
 It would be easy to create a pretty printer that does not display |fromInteger| or the excess parentheses,
-but this is left as an exercise for the reader.
+%but this is left as an exercise for the reader :)
+but we that would take us too far afield here.
 
 \subsection{Showing functions}
-Reconsider our |logLogistic1| function.
+Reconsider our |logLogistic| function.
 It has type
 \begin{lstlisting}
-logLogistic1 :: Floating a => a -> a
-\end{lstlisting}
-What happens if we specialize the |a| parameter to |AST Floating Double|?
-We get a peek at the function's internals.
-\begin{lstlisting}
-ghci> logLogistic1 1 :: AST Floating Double
+logLogistic :: Floating a => a -> a
+\end{lstlisting}
+What happens if we specialize the |a| parameter to be an |AST|?
+We'll get a peek at the function's internals.
+\begin{lstlisting}
+ghci> logLogistic 1 :: AST Floating Double
 (log ((fromInteger 1)/((fromInteger 1)+(exp
@@
 \begin{lstlisting}
-ghci> logLogistic1 (1*3+4) :: AST Floating Double
+ghci> logLogistic (1*3+4) :: AST Floating Double
 (log ((fromInteger 1)/((fromInteger 1)+(exp
@@
 \begin{lstlisting}
-ghci> logLogistic1 (logLogistic1 1)
+ghci> logLogistic (logLogistic 1)
     :: AST Floating Double
@@
 \end{lstlisting}
-Values of type |Var| cannot be plugged directly into the |logLogistic1| function because |Var| is not an instance of |Floating|.
+Values of type |Var| cannot be plugged directly into the |logLogistic| function because |Var| is not an instance of |Floating|.
 Fortunately, we have a way to make any type an instance of |Floating| via |Floating|'s |AST| type.
@@
 \begin{lstlisting}
-ghci> logLogistic1 var1 :: AST Floating Var
+ghci> logLogistic var1 :: AST Floating Var
 (log ((fromInteger 1)/((fromInteger 1)+(exp
@@
 \begin{lstlisting}
-ghci> logLogistic1
+ghci> logLogistic
 \end{lstlisting}
@@
 \begin{lstlisting}
-ghci> logLogistic1 10 :: Double
+ghci> logLogistic 10 :: Double
 -4.539889921682063e-5
 
-ghci> runAST (logLogistic1 10 :: AST Floating Double)
+ghci> runAST (logLogistic 10 :: AST Floating Double)
 -4.539889921682063e-5
@@
 
-\subsection{Cheating with template Haskell}
+\subsection{Boilerplate}
 
@@
 there are many boilerplate instances that we need to write.
-The \homoiconic library provides a template Haskell function |mkFAlgebra| in the |Homogeneous.FAlgebra| module that generates the |FAlgebra|, |Functor|, |Show|, and |View| instances plus all of the pattern synonyms.
+The \homoiconic library provides a template Haskell function |mkFAlgebra| in the |Homogeneous.FAlgebra| module that generates the |FAlgebra|, |Functor|, |Show|, and |View| instances plus all of the pattern synonyms (discussed in the next section) for a given class.
+All of constructions above can be generated by the single invocation
+\begin{lstlisting}
+mkFAlgebra ''Fractional
+\end{lstlisting}
 
@@
 \subsection{Pattern matching the AST}
+\label{sec:patternmatch}
 
@@
 \end{lstlisting}
-As in our |logLogistic1| function,
+As in our |logLogistic| function,
 |testFunc| gives incorrect answers when |a| is specialized to |Double|.
@@
 
-We'll start by correcting the second problem using the |ViewPatterns| language extension.
-|ViewPatterns| provides a syntactic sugar that lets us pattern match on the return value of a function.\footnote{
+We'll start by correcting the second problem.
+The |ViewPatterns| language extension provides syntactic sugar that lets us pattern match on the return value of a function.\footnote{
     The GHC wiki \cite{viewpatterns} provides details on exactly how the desugaring works.
@@
 \end{lstlisting}
-Using it in the pattern gives us
+Using it in the pattern gives us our second program transformation.
 \begin{lstlisting}
@@
 \end{lstlisting}
-This modified program transformation now has the correct polymorphic type signature
+This modified transformation has the correct polymorphic type signature
 (so we've solved problem 2),
@@
 In the free monad, recursion is handled by a call to |fmap| and the base case is handled by extracting the contents of |Pure|.
-This gives our final version of the program transformation.
+
+Here is our final version of the program transformation.
 \begin{lstlisting}
@@
 \end{lstlisting}
-Now we correctly recurse into subexpressions.
+It correctly recurses into subexpressions.
 \begin{lstlisting}
@@
 %(Due to this limitation, GHC has special ad-hoc code just for constant folding.)
-Our program transformations, however, can implement arbitrary computation and so are powerful enough to handle constant folding.
+Our program transformations can implement arbitrary computation and so are powerful enough to handle constant folding.
 
@@
 Let's see an example of constant folding in action.
-First define a simple function with two parameters a handful of constants.
+First define a simple function with two parameters and a handful of constants.
 \begin{lstlisting}
@@
 
-Now let's see how to stabilize the |logLogistic1| function.
-Stabilizing this function is a bit harder because it requires changing the type signature of the function---the constraints on |logLogistic1| require only a |Floating| instance, but the constraints on |logLogistic2| require both a |Floating| and |Ord| instance.
+Now let's see how to stabilize the |logLogistic| function.
+Here's an example of what a stable implementation might look like.
+\begin{lstlisting}
+logLogistic2 :: (Floating x, Ord x) => x -> x
+logLogistic2 x = m+log(1/(exp(m)+exp(-x+m)))
+    where
+        m = min 0 x
+\end{lstlisting}
+Notice that it has a different type signature---we've had to add the |Ord| constraint due to the use of the |min| function.
 Recall that one of the requirements for a function to be homogeneous is that it have only a single class in the context.
@@
 
-We can now write a function that stabilizes |logLogistic1|
+We can now write a function that stabilizes |logLogistic|
 \begin{lstlisting}
@@
 If we have to manually write a unique program transformation for every numeric function,
-then we're not saving any work.
-Fortunately, we don't have to do that.
-
-\subsection{Arbitrary floating point stability}
-
-Creating program transformations that stabilize floating point numbers is a hard,
-actively researched problem.
-Recently, a team from the University of Washington created a tool called Herbie that uses a number of clever techniques to stabilize these floating point expressions \cite{panchekha2015automatically}.
-We can use programs like herbie within our Haskell programs
-
-\begin{lstlisting}
-herbie :: AST Floating a -> IO (AST FloatingOrd a)
-\end{lstlisting}
+then we're not reducing our workload.
+Fortunately we don't have to.
+Designing generic program transformations that stabilize floating point numbers is an actively researched problem.
+For example, a team from the University of Washington recently created the Herbie tool to stabilize floating point expressions \cite{panchekha2015automatically}.
+The output of such a program can easily be used to implement a significantly more advanced version of |stabilize|.
+%Since we can perform arbitrary computations inside our program transformations,
+%there's nothing stopping us from just plugging
+%Actually doing this is left as an exercise for the reader.
+%Any of these techniques can be used with these program transformations described in this section.
+%We can use programs like herbie within our Haskell programs.
+%In order to call out to a foreign program, we need to be in the |IO| monad.
+%\begin{lstlisting}
+%herbie :: AST Floating a -> IO (AST FloatingOrd a)
+%\end{lstlisting}
+%The is essentially how the Herbie GHC plugin works.
 
@@
 logLoss :: Vector a => a -> a -> Scalar a
-logLoss a1 a2 = logLogistic1 (dotproduct a1 a2)
+logLoss a1 a2 = logLogistic (dotproduct a1 a2)
 \end{lstlisting}
@@
 We will require considerably more type hackery in this section.
+Fortunately, this complexity is contained within the construction of the heterogeneous AST.
+From a user's perspective, the heterogeneous AST is just as easy to work with as the homogeneous one.
 
@@
 Type 1 functions have one type family application in the return type.
-For example, |dotProduct| is a type 1 function.
+For example, |dotProduct| is a type 1 function.\footnote{
+In general, a type $n$ function has $n$ type family applications in the return type.
+The techniques in this section can be used to support type $n$ functions for arbitrary $n$,
+but for simplicity, we restrict ourselves to just type 0 and 1 functions.
+}
 \end{defn}
+
+\begin{note}
+The heterogeneous constructions in this section will use many of the same names as the homogeneous constructions in Section \ref{sec:homogeneous}.
+Unless its clear from context, we will always be referring to the heterogeneous version of the construction for the remainder of the paper.
+In the \homoiconic library, the constructions in this section are contained in the |Heterogeneous.FAlgebra| module.
+\end{note}
 
@@
 \subsection{Dealing with type families}
-To address these problems, we first need to consider a limitation with how type families work.
+To address these problems, we first need to consider a limitation of GHC's type families.
 Consider the declaration
@@
 The |t| parameter is the tag representing the type family we want to apply,
-and the |a| parameter is the type to which we want to apply the type family.
+and the |a| parameter is the type to which we want to apply the family.
 The |Scalar| instance looks like
@@
 \end{lstlisting}
-And now we can create a modified version of our |Wrap1| type above that actually works.
+And now we can create a modified version of our |Wrap1| type above that works for type families.
 \begin{lstlisting}
@@
 Frequently, we will want to deal with not just a single type family application,
-but a whole sequence of type family applications.
-We can represent these applications with kind |[Tag]|.
+but rather a sequence of zero or more type family applications.
+We can represent these applications with kind |[Tag]| using GHC's type level lists.
 The following type family applies a sequence of tags to a type.
@@
 At last we are ready to see our type class for heterogeneous FAlgebras.
+\newpage
 \begin{lstlisting}
@@
 Don't panic!
-Understanding exactly how this class works in not necessary for using the ASTs it generates.
-Still, this section will carefully explain how it works.
-
-The most important difference between the homogeneous and heterogeneous |FAlgebra| class is that the |Sig| data family has been given a new type parameter |t::[Tag]|.
-This parameter tells us which type families have been applied to the variable |a|.
-Exactly what this means will become more clear as we see how to create instances of |FAlgebra|.
+Understanding exactly how this class works is not necessary for using the ASTs it generates.
+(To just get a feel for using the heterogeneous ASTs,
+skim the rest of this paper for code snippets containing \textbf{\footnotesize\ttfamily ghci}.)
+
+%We now begin the step-by-step explanation of the heterogeneous |FAlgebra|.
+%The most important difference between the homogeneous and heterogeneous |FAlgebra| class is that the |Sig| data family has been given a new type parameter |t::[Tag]|.
+%This parameter tells us which type families have been applied to the variable |a|.
+%Exactly what this means will become more clear as we see how to create instances of |FAlgebra|.
 
@@
 The return type, however, depends on the return type of the function.
-For type zero functions (there are no type family applications in the return type),
-then |t| parameter is set to |'[]|.
+For type 0 functions (no type family applications in the return type),
+the |t| parameter is set to |'[]|.
 For example, the function constructor for |Vector|'s |(.*)| function is
@@
 \end{lstlisting}
-For type one functions (there is one type family application in the return type),
+For type 1 functions (one type family application in the return type),
 the |t| parameter is the singleton list containing the family's tag.
-For example, the function constructor for |Hilbert|'s |dotProduct| function is
+For example, the return type for |Hilbert|'s |dotProduct| function is |Scalar a|,
+so the corresponding function constructor is
 \begin{lstlisting}
@@
 \end{lstlisting}
-Unlike functor constructors, superclass constructors are parametric in their |t| parameter.
+Unlike function constructors, superclass constructors are parametric in their |t| parameter.
 Superclass constructors take a single parameter,
@@
 \begin{lstlisting}
-Sig_Vector_Floating_Scalar
+Sig_Vector_Floating
   :: Sig Floating t a -> Sig Vector (TScalar ': t) a
@@
 |runSig0| is for running type 0 functions (no type family in the return type),
-and |runSig1| is for running functions type 1 functions (type family in the return type).
+and |runSig1| is for running type 1 functions (one type family in the return type).
+Calling the wrong |run| function will result in a runtime error,
+but our construction of the heterogeneous free monad (in Section \ref{sec:het.eval}) will make this impossible.
 |runSig0| works just like the homogeneous |runSig|,
@@
 so when we run the |Sig|, the result needs to reflect those applications.
-Thus, the parameter to |runSig0| is |Sig alg t (AppTag t a)|,
-and the return type is |AppTag t a|.
+Thus, the parameter to |runSig0| is |Sig alg t (AppTags t a)|,
+and the return type is |AppTags t a|.
 Because the |a| type variable appears only inside type family applications,
@@
 To implement |runSig1|, superclass constructors get called recursively as before,
-but this time for only type 1 functions can be evaluated,
-and type 0 functions will be ignored.
+but this time only type 1 functions can be evaluated,
+and type 0 functions will not type check.
 Here are the |runSig1| functions for |Vector| and |Hilbert|.
@@
 That is, it will allow recursion over the AST.
-We will discuss the function in more detail in Section \ref{sec:heteval} when we discuss evaluating ASTs.
-Before we get to that,
-we'll discuss how to construct and show the ASTs.
+We will discuss the function in more detail in Section \ref{sec:het.eval} when we discuss evaluating ASTs.
+%Before we get to that,
+%we'll discuss how to construct and show the ASTs.
 
@@
   Free0::f      t (Free f t a)->Free f      t  a
-  Pure ::AppTag t a -> Free f t a
+  Pure ::AppTags t a -> Free f t a
 \end{lstlisting}
@@
 We've added an additional parameter |t::[Tag]| that represents the type families that need to get applied to |a| to evaluate the syntax tree.
-The type signature of the |runAST| function (discussed in detail in Section \ref{sec:heteval}) makes this clear.
+The type signature of the |runAST| function (discussed in detail in Section \ref{sec:het.eval}) makes this clear.
 \begin{lstlisting}
 runAST :: (FAlgebra alg, alg a)
-    => Free (Sig alg) t a -> AppTag t a
+    => Free (Sig alg) t a -> AppTags t a
 \end{lstlisting}
@@
 The |Pure| constructor again represents leaves in the AST.
-These leaves now have type |AppTag t a| instead of just |a| to accommodate the modified return type of |runAST|.
+These leaves now have type |AppTags t a| instead of just |a| in order to accommodate the modified return type of |runAST|.
 The |Free0| and |Free1| constructors represent branches in the AST.
@@
 the |Sig_dotProduct| function gets embedded into the AST using the |Free1| constructor.
-This changes the type of the tree by adding the |TScalar| tag.
+This changes the type of the tree by adding |TScalar| to the list of tags.
 When we evaluate the resulting tree, we will get a scalar,
@@
 
-We can make the AST type instances of the type classes similarly to how we did for homogeneous FAlgebras.
-The main difference is that we will need a new |View| class for converting between |Sig|s.
-This |View| class not only converts the |alg| parameter, but also the |t| parameter.
+We can make |AST| an instance of any heterogeneous type class.
+This is a two step process.
+First, we generate appropriate type instances.
+For each type family used in a heterogeneous type class,
+define a type instance for |AST| that adds the family's tag to the list of tags.
+For example, the |Scalar| of an |AST| is given by
+\begin{lstlisting}
+type Scalar (AST alg t a) = AST alg (TScalar ': t) a
+\end{lstlisting}
+
+%The main difference is that we will need a new |View| class for converting between |Sig|s.
+The next step is to define an appropriate |View| class.
+The heterogeneous |View| class not only converts the |alg| parameter, but also the |t| parameter.
 \begin{lstlisting}
@@
 \end{lstlisting}
-Instances of |View| should satisfy the property that
-
-\begin{lstlisting}
-type Scalar (AST alg t a) = AST alg (TScalar ': t) a
-\end{lstlisting}
-
-\begin{lstlisting}
-instance ()
-    => Vector (AST alg t a) where
+%Instances of |View| should satisfy the property that
+
+Armed with these |View| instances, creating heterogeneous type class instances is easy.
+It follows the same pattern as the homogeneous case,
+except that type 0 functions are embedded in the tree using the |Free0| constructor,
+and type 1 functions are embedded using the |Free1| constructor.
+The |Vector| instance is shown below.
+\begin{lstlisting}
+instance
+    ( View Vector '[] Num '[]
+    , View Vector '[] Floating '[TScalar]
+    , View Vector '[] Fractional '[TScalar]
+    , View Vector '[] Num '[TScalar]
+    ) => Vector (AST alg t a) where
     (.*) s a = Free0 $ embedSig $ Sig_dotmul s a
 \end{lstlisting}
-
-\begin{lstlisting}
-instance ()
+And here is the |Hilbert| instance.
+\begin{lstlisting}
+instance
+    ( View Hilbert '[] Vector '[]
+    , View Hilbert '[] Num '[]
+    , View Hilbert '[] Floating '[TScalar]
+    , View Hilbert '[] Fractional '[TScalar]
+    , View Hilbert '[] Num '[TScalar]
     => Hilbert (AST alg t a) where
@@
 
-%\begin{lstlisting}
-%ghci> exp (dotProduct var1 var2) :: AST Hilbert '[TScalar] Var
-%(exp (dotProduct var1 var2))
-%\end{lstlisting}
-%\begin{lstlisting}
-%embedSigTag :: View alg1 (t ': t1) alg2 (t ': t2) => Sig alg1 (t ': t1) a -> Sig alg2 (t ': t2) a
-%embedSigTag = embedSig
-%\end{lstlisting}
+These instances let us easily create ASTs.
+The only difference from the homogeneous case is that we must now explicitly tag the type signature to indicate how the return type of the expression relates to the AST's root type.
+For example, here's an expression over |Double|s.
+\begin{lstlisting}
+expr1 :: AST Floating '[] Double
+expr1 = (1+2)+3 :: AST Floating '[]
+\end{lstlisting}
+And here is the same expression expressed as the |Scalar| of |Vec3 Double|.
+\begin{lstlisting}
+expr2 :: AST Hilbert '[TScalar] (Vec3 Double)
+expr2 = (1+2)+3
+\end{lstlisting}
+We will see more complex examples in the next section's discussion of showing ASTs.
 
@@
 To convert the heterogeneous AST into a |String|,
-we need a |Show| instance for each |Sig| and a |Show| instance for |Free|.
+we need a |Show| instance for for |Free| and a |Show| instance for each |Sig|.
+As is typical for heterogeneous ASTs,
+writing these instances is painful,
+but using them is easy.
+
+The |Show| instance for |Free| is given below.
+\begin{lstlisting}
+instance
+    ( Show      (AppTags t a)
+    , Show      (f t (Free f t a))
+    , ShowUntag (f t (Free f t a))
+    ) => Show (Free f t a) where
+    show (Pure  a ) = show a
+    show (Free0 f0) = "("++show f0++")"
+    show (Free1 f1) = "("++show f1++")"
+\end{lstlisting}
+The instance is similar to the homogeneous instance except for the |ShowUntag| type family in the constraints.
+This family provides the |Show| instance needed to apply |show| to the |f1| variable.
+This variable is declared in the |Free1| constructor,
+but because of this constructor's definition,
+the type of |f1| is not mentioned anywhere in the instance signature!
+The |ShowUntag| function extracts the type of |f1| from instance parameter |t| when it is available and provides a |Show| instance.
+When the type cannot be extracted, GHC is able to prove that the pattern match on |Free1| always fails, and so no constraint is needed.
+The |ShowUntag| family is defined as:
+
+\newpage
+\begin{lstlisting}
+type family ShowUntag (f::Type) :: Constraint where
+    ShowUntag (f (s ': t) (Free f (s ': t) a))
+        = Show (f (s ': t) (Free f t a))
+    ShowUntag a = ()
+\end{lstlisting}
 
@@
 simply show the constructor's arguments connected in an appropriate way by the function's name.
-For each superclass constructor, recursively call show the superclass.
+For each superclass constructor, recursively call show on the superclass.
 As an example, |Vector|'s show instance is given below.
 \begin{lstlisting}
-instance
-    ( Show a
-    , Show (Scalar a)
-    ) => Show (Sig Vector t a)
-        where
+instance (Show a, Show (Scalar a))
+    => Show (Sig Vector t a) where
     show (Sig_Vector_Num s) = show s
     show (Sig_Vector_Floating s) = show s
-    show (Sig_dotmul a1 a2) = show a1++".*"++show a2
+    show (Sig_dotmul a1 a2)
+        = show a1++".*"++show a2
 \end{lstlisting}
@@
 
-This |Show| instance can be defined, but it can't yet be used.
+Unfortunately, we can't yet use these two instances to show an |AST|.
+When we call |show| on an expression of type |AST Vector t a|,
+the constraint solver looks for a |Show| instance of |Scalar (AST Vector t a)|.
+This sends the constraint solver into a loop.
 Recall that we defined
 \begin{lstlisting}
-type Scalar (AST alg t a) = AST alg (TScalar ': t) a
-\end{lstlisting}
-This means that when we call |show| on an |AST|,
-the constraint solver will try to satisfy the |Show (Scalar a)| constraint with the exact same instance.
-In order to satisfy that constraint, we will need |Show (Show (Scalar a))|;
-and to satisfy that, we need |Show (Scalar (Scalar (Scalar a)))|.
-This nesting of |Scalar| type families continues forever,
-and the constraint solver loops!
-
-One solution to this problem is to add the following overlapping instance.
+type Scalar (AST alg t a)
+    = AST alg (TScalar ': t) a
+\end{lstlisting}
+So in order to satisfy the |Show| instance for |AST Vector t a|,
+the constraint solver needs a |Show| instance for |AST Vector '[TScalar] a|,
+which in turn needs a |Show| instance for |AST Vector '[TScalar,TScalar] a|,
+and so on.
+The nesting of these |TScalar| tags continues indefinitely.
+The solution is to create the following overlapping instance.
 \begin{lstlisting}
@@
 \end{lstlisting}
-Once there are at least three |Scalar|s applied to |Show|'s parameter,
+Once there are at least three |TScalar|s in the list of tags,
 this new instance becomes more specific,
@@
 that ends the recursion.
-
-We need another type hack to get the |Show| instance for |Free|.
-%The difficulty here arises due to trying to show the contents of the |Free1| constructor.
-%Below is a |Show| instance ignoring this constructor.
-If we ignore the |Free1| constructor, then the |Show| instance is the exactly the same as for the homogeneous free monad.
-\begin{lstlisting}
-instance
-    ( Show (AppTag t a)
-    , Show (f t (Free f t a))
-    ) => Show (Free f t a) where
-    show (Pure  a ) = show a
-    show (Free0 f0) = "("++show f0++")"
-    show (Free1 f1) = undefined
-\end{lstlisting}
-If we try to show |f1| the same way we show |f0|,
-we run into type problems.
-
-A solution is to create a closed type family to apply a |Show| constraint only when it makes sense.
-\begin{lstlisting}
-type family ShowUntag (f::Type) :: Constraint where
-    ShowUntag (f (s ': t) (Free f (s ': t) a))
-        = Show (f (s ': t) (Free f t a))
-    ShowUntag a = ()
-\end{lstlisting}
-Now we can write the complete |Show| instance as
-\begin{lstlisting}
-instance
-    ( Show      (AppTag t a)
-    , Show      (f t (Free f t a))
-    , ShowUntag (f t (Free f t a))
-    ) => Show (Free f t a) where
-    show (Pure  a) = show a
-    show (Free0 f) = "("++show f++")"
-    show (Free1 f) = "("++show f++")"
-\end{lstlisting}
+Of course there is nothing special about the choice of three elements in the list above;
+the only requirement is that the list be longer than any sequence of type family applications your program actually uses.
 
@@
 actually showing expressions is straightforward.
-You just have to add the appropriate type signature to the expression.
-\begin{lstlisting}
-ghci> logLogistic1 3 :: AST Floating '[] Double
+As in the homogeneous case,
+you just have to add the appropriate type signature to the expression.
+\begin{lstlisting}
+ghci> logLogistic 3 :: AST Floating '[] Double
 (log ((fromInteger 1)/((fromInteger 1)+(exp
@@
 \end{lstlisting}
-
+This same expression can be shown via the |Hilbert| AST as well.
+\begin{lstlisting}
+ghci> logLogistic 3 :: AST Hilbert '[TScalar] (Vec3 Double)
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate (fromInteger 3))))))
+\end{lstlisting}
 In order to display functions,
-we create the |Var| type exactly as before.
+we create the |Var| type as before.
 \begin{lstlisting}
@@
 \end{lstlisting}
-Now when we create values of type |Var|,
-we want them to be usable in expressions of any tagging.
+When we define values of type |Var|,
+we want them to be usable in all expressions regardless of the tag.
 This means they need to be parametric in both the |alg| and |t| parameters.
+Recall that the argument to the |Pure| construct takes type |AppTags t a|,
+so in order to embed a |Var| into an AST using |Pure|,
+we need to guarantee that |AppTags t Var~Var|.\footnote{The {\ttfamily mkFAlgebra} function generates type instances for {\ttfamily Var} for all type families used in a heterogeneous class.
+All of these instances satisfy the property that {\ttfamily AppTags t Var~Var},
+but GHC is unable to prove that all the instances satisfy this property.
+That is why the constraint is explicitly added to the type signature of each {\ttfamily varN} variable.
+}
 \begin{lstlisting}
@@
 %stopzone
-To do this, we need to enforce that any type family applied to |Var| is equivalent to |Var|.
-
-Now we can show our ASTs like before.
-The only difference is that we must now include the |t| parameter as well.
-\begin{lstlisting}
-ghci> logLogistic1 var1 :: AST Floating '[] Var
+
+\begin{lstlisting}
+ghci> logLogistic var1 :: AST Floating '[] Var
 (log ((fromInteger 1)/((fromInteger 1)+(exp
@@
 \end{lstlisting}
-Recall that |Floating| appeared as a constraint for the |Vector| class.
-So you might think we could specialize the expression above to the |Vector| AST.
-But if you try, you get a type error.
-\begin{lstlisting}
-ghci> logLogistic1 var1 :: AST Vector '[] Var
-No instance for (View Floating ('[] Tag) Vector ('[] Tag))
-      arising from a use of ‘logLogistic1’
-\end{lstlisting}
-Just because something implements |Vector| does not guarantee that it implements |Floating| as well.
+Recall that |Floating| appeared as a constraint for the |Hilbert| class.
+So you might think we could specialize the expression above to the |Hilbert| AST.
+\begin{lstlisting}
+ghci> logLogistic var1 :: AST Hilbert '[] Var
+No instance for
+    (View Floating ('[] Tag) Hilbert ('[] Tag))
+\end{lstlisting}
+%\begin{lstlisting}
+%ghci> logLogistic var1 :: AST Hilbert '[] Var
+%\end{lstlisting}
+%This code give us an error message about the following missing instance.
+%\begin{lstlisting}
+%View Floating ('[] Tag) Hilbert ('[] Tag)
+%\end{lstlisting}
+Just because something implements |Hilbert| does not guarantee that it implements |Floating| as well.
 It only guarantees that its |Scalar| implements |Floating|.
-
-\begin{lstlisting}
-ghci> logLogistic1 1 :: AST Vector '[TScalar] Var
+The following modified expression works just fine.
+\begin{lstlisting}
+ghci> logLogistic var1 :: AST Hilbert '[TScalar] Var
 (log ((fromInteger 1)/((fromInteger 1)+(exp
-    (negate (fromInteger 1))))))
-\end{lstlisting}
-
+    (negate var1)))))
+\end{lstlisting}
+Now we demonstrate this technique works with the |logLoss| function,
+which is heterogeneous but not homogeneous.
+\begin{lstlisting}
+ghci> logLoss var1 var2 :: AST Hilbert '[TScalar] Var
+(log ((fromInteger 1)/((fromInteger 1)+(exp
+    (negate (dotProduct var1 var2))))))
+\end{lstlisting}
+And finally, we can have arbitrarily complex expressions in both the scalar and vector section of an expression.
+\begin{lstlisting}
+ghci> (var1+exp var2).*(var3+var1.*var3)
+    :: AST Hilbert '[] Var
+((var1+(exp var2)).*(var3+(var1.*var3)))
+\end{lstlisting}
+Note that in the above expression, |var1| and |var2| are used as scalars whereas |var3| is used as a vector.
+ %(var1+exp var2).*var3 :: AST Hilbert '[] Var
+%((var1+(exp var2)).*var3)
 \subsection{Evaluating the AST}
-\label{sec:heteval}
-
-Running the heterogeneous AST is similar to running the homogeneous AST with three exceptions.
-First, the type signature is different as we've already discussed.
-Second, since the free monad has an additional constructor,
-the |runAST| function will need an additional clause pattern matching against this constructor.
-Finally, we must use the |mapAST| function instead of |fmap| for recursion.
-The full implementation is shown below.
+\label{sec:het.eval}
+
+Our last step in demonstrating the homoiconicity of heterogeneous functions is to evaluate their ASTs.
+We will first discuss the |runAST| function.
+In so doing we will see the motivation for the |mapRun| function from the definition of the heterogeneous |FAlgebra| type.
+Finally, we'll see some examples.
+
+The |runAST| function is shown below.
 \begin{lstlisting}
 runAST :: forall alg t a. (FAlgebra alg, alg a)
-    => Free (Sig alg) t a -> AppTag t a
+    => Free (Sig alg) t a -> AppTags t a
 runAST (Pure  a) = a
@@
 \end{lstlisting}
+It has three differences from the homogeneous case.
+First, (as we've already discussed)
+the type signature is different.
+The |t| parameter to |AST| tracks the type represented by the tree.
+So the type from running the tree must apply |t| to |a|.
+Second, since the free monad has an additional constructor,
+the |runAST| function needs an additional clause pattern matching against this constructor.
+The clauses for the |Free0| and |Free1| constructors are essentially the same though.
+Finally, we use the |mapRun| function instead of |fmap| for recursion.
+Recall the signature of |mapRun|.
+\begin{lstlisting}
+mapRun
+  :: (forall s. Free (Sig alg') s a -> AppTags s a)
+  -> Sig alg t (Free (Sig alg') t' a)
+  -> Sig alg t (AppTags t' a)
+\end{lstlisting}
+This function uses |Rank2Types| to ensure that the function we are mapping can be applied to any syntax tree no matter what type it represents.
+
+As in the homogeneous case,
+when we evalaute an AST for an expression,
+we get the same result as if we had just evaluated the expression directly.
+\begin{lstlisting}
+ghci> logLoss (Vec3 1 2 3) (Vec3 2 3 4)
+-2.0611536942919273e-9
+\end{lstlisting}
+
+\begin{lstlisting}
+ghci> runAST (logLoss (Vec3 1 2 3) (Vec3 2 3 4)
+    :: AST Hilbert '[TScalar] (Vec3 Double))
+-2.0611536942919273e-9
+\end{lstlisting}
+We've finally demonstrated that heterogeneous functions are also homoiconic.
+
+\subsection{Boilerplate}
+
+Constructing heterogeneous FAlgebras is a pain.
+The \homoiconic library provides a |mkFAlgebra| function in the |Heterogeneous.FAlgebra| module to automate this process.
 
@@
 
-There are two more points worth discussion.
-
-The first is efficiency.
-The |stabilize| function is absurdly slow.
-If you view the core, you can see two very different outputs.
-But the output of |stabilize| is highly structured and ready to be optimized.
-It's left as an exercise for the reader to write a GHC plugin to fix the problem.
-
-The second is improved class hierarchies.
-Most of the interesting alternative numeric hierarchies involve many non-homogeneous functions.
-In particular, they use type classes.
+The \homoiconic library lets you construct ASTs from type classes.
+The construction is easiest for homogeneous type classes and more difficult for heterogeneous type classes.
+But in either case, \homoiconic provides Template Haskell functions to shield you from the details and boilerplate.
+Your existing Haskell code bases likely include many type classes on which these constructions work.
+
+%\subsection{Transforming heterogeneous functions}
+%
+%In order to perform program transformations on heterogeneous functions,
+%we need two things.
+%First, we need convenient way to pattern match.
+%The construction is identical to the construction presented in Section \ref{sec:patternmatch}.
+%For each function heterogeneous function in a heterogeneous type class,
+%we get a pattern that is the name of the function prefixed by |AST_|.
+%
+%For example, recall the |testFunc1| function we defined before.
+%\begin{lstlisting}
+%testFunc1 :: Floating a => a -> a
+%testFunc1 = log (exp a)
+%\end{lstlisting}
+%We wrote a program transformation that removed the |log| and |exp| function calls.
+%The same program transformation works, but with a modified type signature to account for the heterogeneous ASTs.
+%\begin{lstlisting}
+%logexpAST3 :: View Floating '[] alg t
+    %=> AST alg t a -> AST alg t a
+%logexpAST3 (AST_log (AST_exp a)) = a
+%\end{lstlisting}
+%Finally, we need a way to perform recursion.
+%
+%\begin{lstlisting}
+%mapTrans :: (forall s. View Floating '[] alg s => AST alg s a -> AST alg s a)
+    %-> Sig alg t (AST alg t a)
+    %-> Sig alg t (AST alg t a)
+%mapTrans f = undefined
+%\end{lstlisting}
+%
+%\begin{lstlisting}
+%\end{lstlisting}
+
+%\section{Conclusion}
+%
+%Heterogeneous FAlgebras probably seem scary at this point.
+
+%The \homoiconic package gives Haskell programmers a new metaprogramming tool based on FAlgebras.
+%Homogeneous FAlgebras are closely related
+%
+%Heterogeneous FAlgebras are scary to construct,
+%but \homoiconic provides a |mkFAlgebra| function that generates the obnoxious details for you.
+%Most of the pain from the heterogeneous FAlgebra type class is due to short comings in the way GHC handles type families.
+
+%It's possible that some future development in the type system will make heterogeneous FAlgebras more pleasant to work with.
+
+%In this paper, we've provided a slight twist on the homogeneous FAlgebra construction that Haskellers know and love and introduced the heterogeneous FAlgebra to the community.
+%Both of these structures give us a new form of metaprogr
+%Heterogeneous FAlgebras are a pain to define in Haskell.
+%But they're not that bad to work with.
+%Essentially the only difference is that we must tag our ASTs with the appropriate type families.
 

### 2016-06-12 src/Heterogeneous/ExamplePrelude.hs

@@ -167,10 +167,15 @@ logexpAST4 (AST_log (AST_exp a)) = a
 
 logexpAST5 :: View Floating '[] alg t => AST alg t a -> AST alg t a
 logexpAST5 (AST_log (AST_exp a)) = a
-logexpAST5 (Free0 f) = Free0 $ fmapAST logexpAST4 f
+-- logexpAST5 (Free1 f) = Free1 $ fmapAST1 logexpAST4 f
+logexpAST5 (Free0 f) = Free0 $ fmapAST0 logexpAST4 f
+logexpAST5 (Pure  a) = Pure a
 
-fmapAST :: (forall s. View Floating '[] alg s => AST alg s a -> AST alg s a) -> Sig alg t (AST alg t a) -> Sig alg t (AST alg t a)
-fmapAST f = undefined
+fmapAST1 :: (forall s. View Floating '[] alg s => AST alg s a -> AST alg s a) -> Sig alg t (AST alg t a) -> Sig alg t (AST alg t a)
+fmapAST1 f = undefined
+
+fmapAST0 :: (forall s. View Floating '[] alg s => AST alg s a -> AST alg s a) -> Sig alg t (AST alg t a) -> Sig alg t (AST alg t a)
+fmapAST0 f = undefined
 
 --     fmapTag' :: Proxy s -> Proxy a -> Proxy b -> App s a -> App s b
 
